### 2017-07-18 cabios-template/main.tex
put gdoc in latex

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\usepackage{graphicx}
\graphicspath{{../figures/}}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Genome-wide genetic studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. This increase in data size has led to the emergence of a range of specialized software for every part of the analysis pipeline. Yet, it is often difficult to choose which piece of software to use and how to combine all these software. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of the RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the Celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1M markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

Genome-wide datasets produced for association studies have dramatically increased in size over the past years. A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation. This diverse and extremely rich software environment has been of tremendous benefit for the genetic community. However, it has two limitations: analysis pipelines become very complex and researchers have limited access to diverse analysis tools due to growing data sizes.

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools available either independently or incorporated within large framework (VCFtools, BCFtools, PLINK...). Similarly genome-wide analysis quality control software have been developed (e.g. Bioconductor GWASTools and PLINK). Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (4 REFs). Then, GWAS analysis itself depends on the genotype format (ProbABEL or SNPTEST for dosage data).  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred, PRSice and other. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis, already implemented by the scientific community. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read/write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory (\cite{Kane2013}). Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

\section{Approach}

We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping. Package bigstatsr implements many statistical tools for several types of ``big.matrix'' objects (raw, char, short, integer, float and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of ``big.matrix'' object to store the genotypes. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (pruning, testing, plotting).

We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.

\begin{methods}
\section{Methods}

\subsection{Memory-mapped files}

The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory. It uses what we could call an hybrid solution: memory-mapping. Memory-mapping is used to access data, possibly stored on disk, as if it were in memory. This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory (\cite{Kane2013}).

We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (REF). With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (REF). The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (REFs).

Usually, memory-mapping provides faster and seamless access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in the memory and every second access would be fast.

\subsection{Data management, preprocessing and imputation} \label{sec:preprocess}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=160pt]{conversion+QC.pdf}}
\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
\end{figure}

Fast read/write operations from/to bed/bim/fam PLINK files are available. One should use PLINK to convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure 1). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage (use of one byte per element, 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''). With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure D). The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (REF). XGBoost build decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{imputation.pdf}}
\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
\end{figure}

\subsection{Population structure and SNP thinning based on Linkage Disequilibrium}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array.}\label{fig:svd}
\end{figure}

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for the computation of a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other on randomized projections (Figure 2).

The function based on randomized projections runs in linear time in all dimensions (REF). FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra (REF). The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix (REF). There are different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance (e.g. association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the MAF as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all (URL).

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (REF). Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs (URL). However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and remove them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (REF). Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have a strong weight (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

\subsection{Association tests and Polygenic Risk Scores}

For association purposes, statistical tests based on linear and logistic regressions are available. Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (REF).

$\forall j \in \{1, \dots, m\}$, 
\itemize{
\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as Age or Gender).
\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K$, where $\hat{p} = \mathbb{P}(Y = 1)$.
}\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

The R packages also implement functions to compute Polygenic Risk Scores.

First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in equations X and Y (REF evans2009). Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients.

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso (REF) and elastic-net (REF) regularizations, which reduce the number of  predictors (SNPs) included in the predictive models. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (REF). The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (REF). These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

\subsection{Data analyzed}

\end{methods}

\section{Results}

\subsection{Overview}

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of memory and 12 cores, and a laptop with only 8GB of memory and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

\subsection{Application}

We performed an association study and computed a polygenic risk score for the Celiac cohort. The data have been pre-processed following steps from figure 1, resulting in 15,250 individuals and 281,122 SNPs. We note that if we used a standard R matrix to store the genotypes, it would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

We used bigstatsr and bigsnpr R functions to get the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure 3). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure 4). As illustrated in the supplementary data, the whole pipeline is user-friendly and takes only 20 lines of code.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. The y-axis has been cut in order to not see only the very strong effects on chromosome 6.}\label{fig:gwas}
\end{figure}

The Celiac dataset is relatively small as compared to modern genetic cohorts. To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, on the whole dataset, we made a linear association test for each SNP, using the 10 first PCs as covariables.

\subsection{Method Comparison}

We first compared the GWAS and PRS computations with the package against PLINK and EIGENSOFT.

(TABLEAU GWAS (feuille 1) et PRS (feuille 2))

For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g. pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table 1). Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same of order of magnitude as when using PLINK and EIGENSOFT (Table x). However, the whole analysis pipeline make use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it. The code used for these two comparisons is available as Supplementary Material, as long as the code for the following comparisons.

On our desktop computer, we compared the computation times between FlashPCA2 and the similar function implemented in bigstatsr, big\_randomSVD. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We replicated the individuals of the celiac disease cohort in order to have three datasets of growing sizes: 15,250, 76,250 and 152,500 individuals. Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure 5). We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition (e.g. SmartPCA). FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 99.9\% when using FlashPCA2 or bigstatsr (Figure 25).

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{benchmark-pca}}
\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the computation time in minutes as a function of the number of samples. The computation corresponds to 10 Principal Components and 93,083 SNPs which remain after thinning.}\label{fig:bench-pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{precision-pca}}
\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA) and package bigstatsr. It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the celiac disease dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
\end{figure}

\subsection{Automatic detection of long-range LD regions}

For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the celiac and POPRES datasets (REFs). For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table SA). We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions (URL) and found a mean correlation of 89.6\% between PCs mainly due to a rotation of PC7 and PC8 (Table SA2). For the celiac dataset, we found 5 long-range LD regions and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping followed by the removing of predetermined long-range LD regions (Table SB and SB2).

For the celiac dataset, we compare results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures E and SE). When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure E). Moreover, loadings are more equally distributed among SNPs (Figure SE) which is confirmed by Gini coefficients (measure of dispersion) of each squared loadings reported in Figure SF, which are all around the theoretical value of $2 / pi$ (for gaussian loadings) and significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{scores}}
\caption{PC4 and PC5 of the celiac disease dataset. Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$). Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4. Right panel, PC scores obtained with the automatic detection and removal of long-range LD regions. Individuals are coloured according to their population of origin.}\label{fig:scores}
\end{figure}

\subsection{Imputation of missing values for genotyped SNPs}

For the fast imputation method based on XGBoost, we compared on the POPRES dataset (REF) composed of 1382 individuals and 344,614 SNPs in both terms of imputation accuracy and computation times. The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure S1). We introduced missing values using a Beta-binomial distribution with a mean of 3\% of missing values. Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17). Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6h to complete while our method only took 42 minutes (20 times less). For the Celiac disease dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48h. We also show that our method's estimation of the number of imputation errors is accurate (Figure X).

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{error-impute}}
\caption{Number of imputation errors vs the estimated number of imputation errors by SNP. For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect. The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost (REF).}\label{fig:error-impute}
\end{figure}

\section{Discussion}

We have developed two R packages, bigstatsr and bigsnpr, which enable multiple analyses of large-scale genotype datasets in a single comprehensive framework. Linkage Disequilibrium pruning, Principal Component Analysis, association tests and computations of polygenic risk scores are made available in this software. Implemented algorithms are both fast and memory-efficient, allowing the use of laptops or desktop computers to make genome-wide analyses. Technically, bigstatsr and bigsnpr could handle any size of datasets. However, if accesses demand the OS to often swap between the file and the memory, this would slow your analysis down. For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times. If the number of samples times the number of SNP remaining after pruning is larger than the available memory, this slowdown would happen. For instance, a 32GB computer would be slow when computing PCs on more than 100K samples and 300K SNPs remaining after LD pruning.

The two R packages don't use some specific file format nor load the entire matrix in memory but rather use a special type of matrix. Using a matrix-like format makes it easy to develop new functions in order to experiment and develop new ideas. Integration in R makes it possible to take advantage of all what R has to offer, for example using the excellent machine learning algorithm XGBoost to easily make a fast yet accurate imputation algorithm for genotyped SNPs. Other functions, not presented here, are also available and all the functions available within the package bigstatsr are not specific to SNP arrays, so that they could be used for other omic data or in completely different fields of research.

We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data. For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use (REF P+T). As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files (REF).

\section*{Acknowledgements}

Text Text Text Text Text Text  Text Text.

\section*{Funding}

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

\bibliographystyle{natbib}
\bibliography{document}

\newpage
\section*{Supplementary results}

\subsection{Long-range LD regions}

\begin{table}[ht]
\centering
\begin{tabular}{lccc}
  \hline
 & Chromosome & Start (Mb) & Stop (Mb) \\ 
  \hline
1 &  2 & 134.7 (134.5) & 137.3 (138) \\ 
  2 &  6 & 27.5 (25.5) & 33.1 (33.5) \\ 
  3 &  8 & 6.6 (8) & 13.2 (12) \\ 
   \hline
\end{tabular}
\caption{Regions found by snp\_autoSVD for the POPRES dataset. In parentheses are regions referenced in (REF)Price et al. (2008).} 
\label{tab:lrldr-popres}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jul 18 15:59:51 2017
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
  \hline
 & Chromosome & Start (Mb) & Stop (Mb) \\ 
  \hline
1 &  2 & 134.4 (134.5) & 138.1 (138) \\ 
  2 &  6 & 23.8 (25.5) & 35.8 (33.5) \\ 
  3 &  8 & 6.3 (8) & 13.5 (12) \\ 
  4 &  3 & 163.1 (n/a) & 164.9 (n/a) \\ 
  5 & 14 & 46.6 (n/a) & 47.5 (n/a) \\ 
   \hline
\end{tabular}
\caption{Regions found by snp\_autoSVD for the celiac dataset. In parentheses are regions referenced in (REF)Price et al. (2008).} 
\label{tab:lrldr-celiac}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:29:13 2017
\begin{table}[ht]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.0 & 0.1 & -0.1 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\
  PC2 & 0.1 & 100.0 & -0.0 & 0.1 & -0.0 & -0.0 & -0.0 & 0.2 & -0.1 & -0.0 \\
  PC3 & 0.0 & -0.0 & 99.9 & 0.9 & 0.1 & -0.1 & -0.3 & 0.2 & 0.4 & 0.1 \\
  PC4 & -0.1 & -0.1 & -0.9 & 99.7 & -1.0 & 0.7 & 0.6 & 0.2 & 0.3 & 0.9 \\
  PC5 & 0.1 & 0.0 & -0.1 & 1.1 & 99.3 & 1.3 & -0.8 & 1.3 & -4.2 & -2.4 \\
  PC6 & -0.0 & 0.0 & 0.1 & -0.7 & -1.0 & 97.7 & -3.5 & 6.1 & 7.9 & -6.2 \\
  PC7 & -0.0 & -0.1 & 0.2 & -0.3 & -1.7 & 0.3 & 58.3 & 73.2 & -25.9 & 9.1 \\
  PC8 & 0.1 & -0.1 & -0.3 & 0.4 & -0.5 & -5.3 & -73.5 & 59.5 & 15.8 & 13.2 \\
  PC9 & 0.0 & 0.1 & -0.4 & -0.8 & 5.0 & -7.6 & 27.8 & 11.0 & 91.9 & 9.0 \\
  PC10 & 0.1 & 0.0 & 0.0 & -0.9 & 1.6 & 10.2 & 3.9 & -19.6 & -6.3 & 89.2 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the POPRES dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-popres}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:25:56 2017
\begin{table}[ht]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
  PC2 & 0.1 & 100.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 \\
  PC3 & 0.1 & -0.0 & 99.9 & 0.2 & -0.0 & 0.1 & 0.1 & 0.1 & 0.0 & -0.1 \\
  PC4 & -0.0 & -0.0 & -0.3 & 99.9 & -0.1 & 0.1 & -0.1 & 0.0 & 0.1 & 0.1 \\
  PC5 & 0.0 & 0.0 & 0.0 & 0.1 & 99.7 & 0.9 & -0.3 & 0.1 & -0.8 & -0.6 \\
  PC6 & -0.0 & 0.0 & -0.1 & -0.2 & -0.8 & 99.6 & 0.5 & -0.5 & -0.2 & -0.4 \\
  PC7 & -0.0 & 0.0 & -0.1 & 0.0 & 0.5 & -0.4 & 98.9 & 3.1 & 0.7 & 1.6 \\
  PC8 & 0.0 & 0.0 & -0.2 & -0.0 & -0.2 & 0.5 & -3.2 & 98.4 & -4.5 & -1.5 \\
  PC9 & -0.0 & -0.0 & -0.0 & 0.0 & 0.6 & 0.1 & -0.7 & 4.6 & 96.9 & -10.7 \\
  PC10 & -0.0 & -0.0 & 0.1 & -0.1 & 0.3 & 0.1 & -1.2 & 1.5 & 8.6 & 92.7 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-popres}
\end{table}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=220pt]{loadings}}
\caption{Loadings of first 6 PCs of the celiac disease dataset plotted as hexbins (2-D histogram with hexagonal cells). On the left, without removing any long range LD region (only clumping at $R^2 > 0.2$). On the right, with the automatic detection and removal of long-range LD regions.}\label{fig:loadings}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=220pt]{gini}}
\caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $R^2 > 0.2$) and with the automatic detection and removal of long-range LD regions. The dashed line corresponds to the theoretical value for gaussian loadings.}\label{fig:gini}
\end{figure}

\subsection{Imputation}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=180pt]{hist-maf}}
\caption{Histogram of the minor allele frequencies of the POPRES dataset used for comparing imputation methods.}\label{fig:maf}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=180pt]{hist-NA}}
\caption{Histogram of the number of missing values by SNP. These numbers were generated using a Beta-binomial distribution.}\label{fig:NA}
\end{figure}

\end{document}

### 2017-07-19 cabios-template/main.tex
in preprocessing of results

@@
 In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs.
-We illustrate applications of the two R packages by analysing a case-control genomic dataset for the Celiac disease, performing an association study and computing Polygenic Risk Scores.
-Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1M markers on a single desktop computer.
+We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
+Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
 \\
@@
 Consider first the basic tools necessary to perform a standard genome-wide analysis.
-Conversions between standard file formats has become a field by itself with several tools available either independently or incorporated within large framework (VCFtools, BCFtools, PLINK...).
-Similarly genome-wide analysis quality control software have been developed (e.g.
-Bioconductor GWASTools and PLINK).
-Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (4 REFs).
-Then, GWAS analysis itself depends on the genotype format (ProbABEL or SNPTEST for dosage data).
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred, PRSice and other.
+Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework (\cite{Danecek2011,Li2011,Purcell2007}).
+Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools (\cite{Gogarten2012}).
+Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (\cite{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}).
+Then, GWAS analysis itself depends on the genotype format, e.g.
+ProbABEL or SNPTEST for dosage data (\cite{Aulchenko2010,Marchini2010}.
+Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice (\cite{Vilhjalmsson2015,Euesden2015}).
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
@@
 The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory.
-It uses what we could call an hybrid solution: memory-mapping.
+They use what we could call an hybrid solution: memory-mapping.
 Memory-mapping is used to access data, possibly stored on disk, as if it were in memory.
@@
 
-We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (REF).
-With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (REF).
-The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R.
-For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (REFs).
+We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (\cite{Nielsen2008}).
+With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (\cite{Eddelbuettel2011}).
+The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R (\cite{R2017}.
+For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (\cite{Zeng2017,RSpectra2016}).
 
-Usually, memory-mapping provides faster and seamless access than standard read/write operations.
+Usually, memory-mapping provides seamless and faster access than standard read/write operations.
 When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory.
@@
 It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on.
-Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in the memory and every second access would be fast.
+Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.
 
@@
 One should use PLINK to convert data to this format.
-In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure 1).
+In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}).
 PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs.
@@
 
-Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure D).
-The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R.
+Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
+The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R (\cite{Browning2009}).
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
@@
 It only relies on some local XGBoost models.
-XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (REF).
-XGBoost build decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices.
+XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (\cite{Chen2016}).
+XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices.
 For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets.
@@
 
-The function based on randomized projections runs in linear time in all dimensions (REF).
-FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra (REF).
+The function based on randomized projections runs in linear time in all dimensions (\cite{Lehoucq1996}).
+FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
 The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.
@@
 Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed.
-For instance, thinning is mandatory when computing PCs of a genotype matrix (REF).
+For instance, thinning is mandatory when computing PCs of a genotype matrix (\cite{Abdellaoui2013}).
 There are different options to thin SNPs based on Linkage Disequilibrium.
@@
 association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest.
-Furthermore, we advise to always use clumping instead of pruning (by using the MAF as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all (URL).
+Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.
 
-The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (REF).
-Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs (URL).
-However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and remove them.
-This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (REF).
+The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (\cite{Price2008}).
+Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs\footnote{https://goo.gl/8TngVE}.
+However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them.
+This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (\cite{Luu2017}).
 Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm.
-Indeed, a long-range LD region would cause SNPs in this region to have a strong weight (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive.
+Indeed, a long-range LD region would cause SNPs in this region to have strong weights (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive.
 This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.
@@
 Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
-Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (REF).
+Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (\cite{Zheng2012}).
 
@@
 \itemize{
-\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as Age or Gender).
-\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K$, where $\hat{p} = \mathbb{P}(Y = 1)$.
+\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
+\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $\hat{p} = \mathbb{P}(Y = 1)$.
 }\\
@@
 
-First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in equations X and Y (REF evans2009).
+First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
 Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
@@
 Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
-Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients.
+Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients (\cite{Dudbridge2013,Chatterjee2013,Golan2014}).
 
@@
 The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
-These models include lasso (REF) and elastic-net (REF) regularizations, which reduce the number of  predictors (SNPs) included in the predictive models.
-Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (REF).
-The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (REF).
+These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models (\cite{Tibshirani1996,Zou2005,Friedman2010}).
+Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (\cite{Tibshirani2012}).
+The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (\cite{Zeng2017}).
 These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).
@@
 \subsection{Data analyzed}
+
+In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets (\cite{Dubois2010,Nelson2008}).
+The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
+The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
+
+For computation timing purposes only, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structure of Linkage Disequilibrium and Population Structure as the original dataset.
+To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.
 
@@
 We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort.
-We compare the performance on two computers, a desktop computer with 64GB of memory and 12 cores, and a laptop with only 8GB of memory and 4 cores.
+We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores.
 For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
@@
 We performed an association study and computed a polygenic risk score for the Celiac cohort.
-The data have been pre-processed following steps from figure 1, resulting in 15,250 individuals and 281,122 SNPs.
+The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 (\cite{Purcell2007}).
+For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
+For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that we imputed with our new method (see section~\ref{sec:impute}).
 We note that if we used a standard R matrix to store the genotypes, it would take 32GB of memory.
@@
 
-\subsection{Imputation of missing values for genotyped SNPs}
+\subsection{Imputation of missing values for genotyped SNPs}\label{sec:impute}
 
@@
 \caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
-\label{tab:pc-popres}
+\label{tab:pc-celiac}
 \end{table}

### 2017-07-19 preprocessing.Rmd
in preprocessing of results

---
title: "Preprocessing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quality Control

```{r}
library(bigsnpr)

NCORES <- parallel::detectCores() / 2
plink <- "../plink_linux_x86_64/plink"

popresQC.bed <- snp_plinkQC(
  plink.path = plink,
  prefix.in = "../POPRES_data/POPRES_allchr",
  file.type = "--file", # ped/map
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

popresQC2.bed <- snp_plinkIBDQC(plink, popresQC.bed, ncores = NCORES)
```

```{r}
celiacQC.bed <- snp_plinkQC(
  plink.path = plink,
  prefix.in = "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300",
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

celiacQC2.bed <- snp_plinkIBDQC(plink, celiacQC.bed, ncores = NCORES)
```

## Transform PLINK files in big.matrix format

```{r}
popresQC.rds <- snp_readBed(popresQC2.bed, "popresQC")

system.time(
  celiacQC.rds <- snp_readBed(celiacQC2.bed, "celiacQC")
)
```

## Imputation of Celiac dataset

```{r}
system.time(
  celiac <- snp_attach(celiacQC.rds)
)
str(celiac, max.level = 2)
```

```{r}
G <- celiac$genotypes
big_counts(G, ind.col = 1:10)
```

```{r}

```

### 2017-07-20 imputation.Rmd
add imputation

---
title: "Imputation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generate missing values

```{r, results='hide'}
library(bigsnpr)
library(dplyr)
library(ggplot2)

NCORES <- parallel::detectCores() / 2
```

```{r}
popres <- snp_attach("backingfiles/popresQC.rds")

G <- popres$genotypes
n <- nrow(G)
m <- ncol(G)
```

```{r}
# Compute minor allele frequencies
maf <- snp_MAF(G)

bigstatsr:::MY_THEME(ggplot(), coeff = 1.8) +
  geom_histogram(aes(maf)) + 
  labs(x = "Minor Allele Frequency")

ggsave("figures/hist-maf.png", width = 1300, height = 900, scale = 1/75)
```

```{r}
# Generate number of missing values by SNP with a Beta-Binomial distribution
nbNA <- VGAM::rbetabinom.ab(m, size = n, shape1 = 0.6, shape2 = 20)

sum(nbNA) / (n * m)

bigstatsr:::MY_THEME(ggplot(), coeff = 1.8) +
  geom_histogram(aes(nbNA), bins = 100) + 
  labs(x = "Number of missing values")

ggsave("hist-NA.png", width = 1300, height = 900, scale = 1/75)
```

```{r}
# Generate indices of missing values
indNA <- cbind(
  lapply(nbNA, function(nb) {
    `if`(nb > 0, sample(n, size = nb), NULL)
  }) %>% unlist(), 
  rep(cols_along(G), nbNA)
)

# Fill a copy of the matrix with NAs (coded as 03)
X <- attach.BM(G)
XNA <- as.BM.code(deepcopy(X), code = bigsnpr:::CODE_IMPUTE_PRED)
XNA[indNA] <- as.raw(3)

GNA <- describe(XNA)
```

## Imputation with our new method

```{r}
system.time(
  test <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
) # Latest timings: 42 / 48 minutes
```

```{r}
# Errors
mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 %
table(XNA[indNA], X[indNA])

# Plot the estimation of the errors of imputation
data_frame(col = indNA[, 2], error = XNA[indNA] != X[indNA]) %>%
  group_by(col) %>%
  summarise(nb_err = sum(error)) %>%
  mutate(nb_err_est = (test$pError * test$pNA * n)[col]) %>%
  ggplot() %>%
  bigstatsr:::MY_THEME(coeff = 1.8) + 
  geom_abline(slope = 1, intercept = 0, col = "red") + 
  geom_point(aes(nb_err, nb_err_est), alpha = 0.4) +
  labs(x = "Number of errors", y = "Estimated number of errors")

ggsave("error-impute.png", width = 1000, height = 1000, scale = 1/75)
```

## Imputation with BEAGLE

```{r}
# Cleaning from previous runs 
file.remove(list.files("backingfiles", pattern = "tmp*", full.names = TRUE))
```

```{r}
# Write a new bedfile
popresNA <- popres
popresNA$genotypes <- GNA
bedfile <- snp_writeBed(popresNA, bedfile = "backingfiles/tmp.bed")
```

```{r}
system.time(
  bedfile2 <- snp_beagleImpute(beagle.path = "../../beagle.21Jan17.6cc.jar",
                               plink.path = "../plink_linux_x86_64/plink",
                               bedfile.in = "backingfiles/tmp.bed",
                               memory.max = 30,
                               ncores = NCORES)
) # 14.6 / 14.2 hours
```

```{r}
# Read bedfile into R and assess the errors of imputation
popres.beagle <- snp_readBed(bedfile2, backingfile = "tmp_impute") %>%
  snp_attach()
X_beagle <- attach.BM(popres.beagle$genotypes)

mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 %
table(X_beagle[indNA], X[indNA])
```

### 2017-07-20 cabios-template/main.tex
review mike 1

@@
 
+\usepackage{adjustbox}
+\usepackage{multirow}
 \usepackage{graphicx}
@@
 Consider first the basic tools necessary to perform a standard genome-wide analysis.
-Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework (\cite{Danecek2011,Li2011,Purcell2007}).
-Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools (\cite{Gogarten2012}).
-Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (\cite{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}).
+Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
+Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}.
+Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}.
 Then, GWAS analysis itself depends on the genotype format, e.g.
-ProbABEL or SNPTEST for dosage data (\cite{Aulchenko2010,Marchini2010}.
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice (\cite{Vilhjalmsson2015,Euesden2015}).
+ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.
+Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice \cite[]{Vilhjalmsson2015,Euesden2015}.
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
@@
 For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read/write operations.
-Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory (\cite{Kane2013}).
+Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
 Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
@@
 
-\begin{methods}
 \section{Methods}
@@
 Memory-mapping is used to access data, possibly stored on disk, as if it were in memory.
-This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory (\cite{Kane2013}).
+This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory \cite[]{Kane2013}.
 
-We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (\cite{Nielsen2008}).
-With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (\cite{Eddelbuettel2011}).
-The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R (\cite{R2017}.
-For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (\cite{Zeng2017,RSpectra2016}).
+We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
+With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
+The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
+For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{Zeng2017,RSpectra2016}.
 
@@
 Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
-The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R (\cite{Browning2009}).
+The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
@@
 It only relies on some local XGBoost models.
-XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (\cite{Chen2016}).
+XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
 XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices.
@@
 
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for the computation of a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other on randomized projections (Figure 2).
+For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for the computation of a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other on randomized projections (Figure~\ref{fig:svd}).
 
-The function based on randomized projections runs in linear time in all dimensions (\cite{Lehoucq1996}).
+The function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
 FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
@@
 Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed.
-For instance, thinning is mandatory when computing PCs of a genotype matrix (\cite{Abdellaoui2013}).
+For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}.
 There are different options to thin SNPs based on Linkage Disequilibrium.
@@
 
-The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (\cite{Price2008}).
+The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
 Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs\footnote{https://goo.gl/8TngVE}.
 However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them.
-This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (\cite{Luu2017}).
+This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
 Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm.
@@
 Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
-Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (\cite{Zheng2012}).
+Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
 
@@
 Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
-Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients (\cite{Dudbridge2013,Chatterjee2013,Golan2014}).
+Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Dudbridge2013,Chatterjee2013,Golan2014}.
 
@@
 The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
-These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models (\cite{Tibshirani1996,Zou2005,Friedman2010}).
-Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (\cite{Tibshirani2012}).
-The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (\cite{Zeng2017}).
+These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Tibshirani1996,Zou2005,Friedman2010}.
+Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}.
+The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}.
 These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).
@@
 
-In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets (\cite{Dubois2010,Nelson2008}).
+In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
 The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
@@
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.
+\label{sec:rep}
 
-\end{methods}
+\subsection{Reproducibility}
+
+All the code used in this paper along with results, such as execution times and figures, are available as HTML notebooks in the Supplementary Data.
 
@@
 
-\subsection{Overview}
+\subsection{Overview}\label{sec:overview}
 
@@
 We performed an association study and computed a polygenic risk score for the Celiac cohort.
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 (\cite{Purcell2007}).
+
+The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
 For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
-For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that we imputed with our new method (see section~\ref{sec:impute}).
-We note that if we used a standard R matrix to store the genotypes, it would take 32GB of memory.
+For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with our new method (see section~\ref{sec:impute}).
+We note that if we used a standard R matrix to store the genotypes, this data would take 32GB of memory.
 On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.
 
-We used bigstatsr and bigsnpr R functions to get the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure 3).
-We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure 4).
+We used bigstatsr and bigsnpr R functions to get the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}).
+We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
 As illustrated in the supplementary data, the whole pipeline is user-friendly and takes only 20 lines of code.
@@
 
-We first compared the GWAS and PRS computations with the package against PLINK and EIGENSOFT.
+We first compared the GWAS and PRS computations with the R packages against PLINK 1.9 and EIGENSOFT 6.1.4.
 
-(TABLEAU GWAS (feuille 1) et PRS (feuille 2))
+\begin{table}[!tpb]
+\centering
+\begin{adjustbox}{max width=235pt}
+\begin{tabular}{|c|cc|cc|}
+\hline
+Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\ 
+\hline
+Reading PLINK files & n/a & n/a & 5 / 20 sec & snp\_readBed \\
+Pruning & plink -{}-exclude -{}-range -{}-indep-pairwise & 4 / 6 sec & 13 / 46 sec & snp\_indLRLDR \& snp\_pruning \\
+\multirow{2}{*}{Computing 10 PCs} & plink -{}-extract -{}-make-bed & 2 sec & \multirow{2}{*}{45 / 136 sec} & \multirow{2}{*}{big\_randomSVD} \\
+& flashpcaR::flashpca (FlashPCA2) & 6 / 7 min &  &  \\
+GWAS (binary phenotype) & plink -{}-logistic -{}-covar & 5 min & 5 / 14.5 min & big\_univLogReg \\
+\hline
+Total & PLINK and flashpcaR & 11 / 12 min & 6 / 18 min & bigsnpr and bigstatsr \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
+The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\label{tab:bench-gwas}
+\end{table}
+
+\begin{table}[!tpb]
+\centering
+\begin{adjustbox}{max width=235pt}
+\begin{tabular}{|c|cc|cc|}
+\hline
+Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\
+\hline
+GWAS (binary phenotype) & plink -{}-keep -{}-logistic -{}-covar & 4 / min & 3 / min & big\_univLogReg \\
+Clumping & plink -{}-keep -{}-clump -{}-clump-p1
+-{}-clump-p2 -{}-clump-r2 & 49 / sec & 9 / sec & snp\_clumping \\
+PRS & plink -{}-keep -{}-extract -{}-score -{}-q-score-range & 9 / sec & 4 / sec & snp\_PRS \\
+\hline
+Total & PLINK & 5 / min & 3 / min & bigsnpr and bigstatsr \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
+The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\label{tab:bench-prs}
+\end{table}
 
 For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.
-pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table 1).
-Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same of order of magnitude as when using PLINK and EIGENSOFT (Table x).
-However, the whole analysis pipeline make use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.
-The code used for these two comparisons is available as Supplementary Material, as long as the code for the following comparisons.
+pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
+Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}).
+However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.
 
-On our desktop computer, we compared the computation times between FlashPCA2 and the similar function implemented in bigstatsr, big\_randomSVD.
+On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 and the similar function implemented in bigstatsr, big\_randomSVD.
 For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs.
-We replicated the individuals of the celiac disease cohort in order to have three datasets of growing sizes: 15,250, 76,250 and 152,500 individuals.
-Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure 5).
+We used the datasets of growing size simulated from the Celiac dataset (Section~\ref{sec:rep}).
+Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure~\ref{fig:bench-pca}).
 We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition (e.g.
 SmartPCA).
-FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 99.9\% when using FlashPCA2 or bigstatsr (Figure 25).
+FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
 
@@
 
-For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the celiac and POPRES datasets (REFs).
+For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets.
 For the POPRES dataset, the algorithm converged in two iterations.
-The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table SA).
-We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions (URL) and found a mean correlation of 89.6\% between PCs mainly due to a rotation of PC7 and PC8 (Table SA2).
-For the celiac dataset, we found 5 long-range LD regions and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping followed by the removing of predetermined long-range LD regions (Table SB and SB2).
+The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}).
+We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
+For the Celiac dataset, we found 5 long-range LD regions and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
 
-For the celiac dataset, we compare results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$).
-When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures E and SE).
-When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure E).
-Moreover, loadings are more equally distributed among SNPs (Figure SE) which is confirmed by Gini coefficients (measure of dispersion) of each squared loadings reported in Figure SF, which are all around the theoretical value of $2 / pi$ (for gaussian loadings) and significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed.
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{scores}}
-\caption{PC4 and PC5 of the celiac disease dataset.
-Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$).
-Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4.
-Right panel, PC scores obtained with the automatic detection and removal of long-range LD regions.
-Individuals are coloured according to their population of origin.}\label{fig:scores}
-\end{figure}
+For the Celiac dataset, we compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$).
+When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
+When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure~\ref{fig:scores}).
+Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}) which is confirmed by Gini coefficients (measure of dispersion) of each squared loadings reported in Figure~\ref{fig:gini}, which are all around the theoretical value of $2 / pi$ (for gaussian loadings) and significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed.
 
@@
 
-For the fast imputation method based on XGBoost, we compared on the POPRES dataset (REF) composed of 1382 individuals and 344,614 SNPs in both terms of imputation accuracy and computation times.
-The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure S1).
-We introduced missing values using a Beta-binomial distribution with a mean of 3\% of missing values.
+For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset.
+The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}).
+We introduced missing values using a Beta-binomial distribution resulting in approximately 3\% of missing values.
 Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17).
-Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6h to complete while our method only took 42 minutes (20 times less).
-For the Celiac disease dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48h.
-We also show that our method's estimation of the number of imputation errors is accurate (Figure X).
+Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
+For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
+We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
 
@@
 For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
-The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost (REF).}\label{fig:error-impute}
+The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
 \end{figure}
@@
 For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times.
-If the number of samples times the number of SNP remaining after pruning is larger than the available memory, this slowdown would happen.
-For instance, a 32GB computer would be slow when computing PCs on more than 100K samples and 300K SNPs remaining after LD pruning.
+If the number of samples times the number of SNPs remaining after pruning is larger than the available memory, this slowdown would happen.
+For instance, a 32GB computer would be slow when computing PCs on more than 100K samples and 300K SNPs remaining after LD thinning.
 
@@
 We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use (REF P+T).
-As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files (REF).
+For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use.
+As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files \cite[]{Palmer2016}.
 
@@
 \newpage
-\section*{Supplementary results}
+\section*{Supplementary Data}
 
@@
 
-\begin{table}[ht]
+\begin{table}[!tpb]
 \centering
@@
 \end{tabular}
-\caption{Regions found by snp\_autoSVD for the POPRES dataset. In parentheses are regions referenced in (REF)Price et al. (2008).} 
+\caption{Regions found by snp\_autoSVD for the POPRES dataset. In parentheses are regions referenced in \cite[]{Price2008}.}
 \label{tab:lrldr-popres}
@@
 % Tue Jul 18 15:59:51 2017
-\begin{table}[ht]
+\begin{table}[!tpb]
 \centering
@@
 \end{tabular}
-\caption{Regions found by snp\_autoSVD for the celiac dataset. In parentheses are regions referenced in (REF)Price et al. (2008).} 
+\caption{Regions found by snp\_autoSVD for the celiac dataset. In parentheses are regions referenced in \cite[]{Price2008}.}
 \label{tab:lrldr-celiac}
@@
 % Tue Jun 13 11:29:13 2017
-\begin{table}[ht]
+\begin{table}[!tpb]
 \centering
@@
 % Tue Jun 13 11:25:56 2017
-\begin{table}[ht]
+\begin{table}[!tpb]
 \centering
@@
 \end{table}
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=235pt]{scores}}
+\caption{PC4 and PC5 of the celiac disease dataset.
+Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$).
+Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4.
+Right panel, PC scores obtained with the automatic detection and removal of long-range LD regions.
+Individuals are coloured according to their population of origin.}\label{fig:scores}
+\end{figure}
 

### 2017-07-20 imputation.Rmd
review mike 1

@@
 NCORES <- parallel::detectCores() / 2
+PLINK <- "../plink_linux_x86_64/plink"
+BEAGLE <- "../../beagle.21Jan17.6cc.jar"
 ```
@@
 system.time(
-  test <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
+  infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
 ) # Latest timings: 42 / 48 minutes
@@
   summarise(nb_err = sum(error)) %>%
-  mutate(nb_err_est = (test$pError * test$pNA * n)[col]) %>%
+  mutate(nb_err_est = (infos$pError * infos$pNA * n)[col]) %>%
   ggplot() %>%
-  bigstatsr:::MY_THEME(coeff = 1.8) + 
+  bigstatsr:::MY_THEME(coeff = 2) + 
   geom_abline(slope = 1, intercept = 0, col = "red") + 
@@
 system.time(
-  bedfile2 <- snp_beagleImpute(beagle.path = "../../beagle.21Jan17.6cc.jar",
-                               plink.path = "../plink_linux_x86_64/plink",
+  bedfile2 <- snp_beagleImpute(beagle.path = BEAGLE,
+                               plink.path = PLINK,
                                bedfile.in = "backingfiles/tmp.bed",

### 2017-07-20 preprocessing.Rmd
review mike 1

@@
 ```{r}
+# Fast imputation
+system.time(
+  infos <- snp_fastImpute(G, celiac$map$chromosome, ncores = NCORES)
+)
+```
 
+```{r}
+plot(subset(infos, pNA > 0.001), pch = 19, cex = 0.5)
+pvals <- c(0.01, 0.005, 0.002, 0.001); colvals <- 2:5
+idc <- lapply(seq_along(pvals), function(i) {
+  curve(pvals[i] / x, from = 0, lwd = 2, 
+        col = colvals[i], add = TRUE)
+})
+legend("topright", legend = pvals, title = "p(NA & Error)",
+       col = colvals, lty = 1, lwd = 2)
 ```
+
+```{r}
+big_counts(G, ind.col = 1:10)
+```
+
+```{r}
+{{G@code <- bigsnpr:::CODE_IMPUTE_PRED}}
+big_counts(G, ind.col = 1:10)
+```

### 2017-07-22 imputation.Rmd
3rd run of imputation

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 15)
 ```
@@
 
-```{r, results='hide'}
+```{r, message=FALSE}
 library(bigsnpr)
@@
 
-bigstatsr:::MY_THEME(ggplot(), coeff = 1.8) +
+bigstatsr:::MY_THEME(ggplot(), coeff = 2) +
   geom_histogram(aes(maf)) + 
@@
 
-bigstatsr:::MY_THEME(ggplot(), coeff = 1.8) +
+bigstatsr:::MY_THEME(ggplot(), coeff = 2) +
   geom_histogram(aes(nbNA), bins = 100) + 
@@
 
-## Imputation with our new method
+## Imputation with the new method based on XGBoost
 
@@
   infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
-) # Latest timings: 42 / 48 minutes
+) # Previous runs: 42 / 48 / 48 minutes
 ```
@@
 # Errors
-mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 %
+mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 / 4.8 %
 table(XNA[indNA], X[indNA])
@@
 
-## Imputation with BEAGLE
+## Imputation with BEAGLE (called via bigsnpr)
 
@@
                                ncores = NCORES)
-) # 14.6 / 14.2 hours
+) # 14.6 / 14.2 / 14.0 hours
 ```
@@
 
-mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 %
+mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 / 4.8 %
 table(X_beagle[indNA], X[indNA])

### 2017-07-22 cabios-template/draft.tex
draft 2

--- 2017-07-20 cabios-template/main.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
+%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
+%% Do not edit unless you really know what you are doing.
+\documentclass[english, 12pt]{article}
+\usepackage{times}
+%\usepackage{algorithm2e}
+\usepackage{url}
+\usepackage{bbm}
+\usepackage[T1]{fontenc}
+\usepackage[latin1]{inputenc}
+\usepackage{geometry}
+\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
+\usepackage{rotating}
+\usepackage{color}
+\usepackage{graphicx}
+\usepackage{amsmath, amsthm, amssymb}
+\usepackage{setspace}
+\usepackage{lineno}
+\usepackage{hyperref}
+\usepackage{bbm}
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+\linenumbers
+\doublespacing
+%\usepackage[authoryear]{natbib}
+\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
+
+%Pour les rajouts
+\usepackage{color}
+\definecolor{trustcolor}{rgb}{0,0,1}
 
@@
 
+\makeatletter
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
+%% Bold symbol macro for standard LaTeX users
+%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
+
+%% Because html converters don't know tabularnewline
+\providecommand{\tabularnewline}{\\}
+
+\usepackage{babel}
+\makeatother
+
 \begin{document}
-\firstpage{1}
-
-\subtitle{Subject Section}
 
 \title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
-\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\author[Privé \textit{et~al}.]{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
 $^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
-$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}
+$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
 
-\corresp{$^\ast$To whom correspondence should be addressed.}
+$^\ast$To whom correspondence should be addressed.
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+\date{~ }
 
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{\textbf{Motivation:} Genome-wide genetic studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
+\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
 This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses.
-This increase in data size has led to the emergence of a range of specialized software for every part of the analysis pipeline.
-Yet, it is often difficult to choose which piece of software to use and how to combine all these software.
+Specialized software for every part of the analysis pipeline have been developed to handle large genomic data.
+However, combining all these software into a single data analysis pipeline might be technically difficult.
 \\
 \textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework.
-To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of the RAM.
+To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM.
 To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods.
@@
 We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
-Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
+Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
 \\
 \textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
-\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 Consider first the basic tools necessary to perform a standard genome-wide analysis.
-Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
-Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}.
-Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}.
-Then, GWAS analysis itself depends on the genotype format, e.g.
+Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
+Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}.
+Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}.
+Then, implementation of GWAS analyses also depends on the genotype format, e.g.
 ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice \cite[]{Vilhjalmsson2015,Euesden2015}.
+Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple variants \cite[]{Vilhjalmsson2015,Euesden2015}.
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
-Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.
+Overall, this might be a brake on data exploration.
+%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.
 
@@
 Hopefully, strategies have been developed to avoid loading large datasets in RAM.
-For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read/write operations.
+For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations.
 Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
@@
 
-We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
+We are aware the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
 With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
 The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
-For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{Zeng2017,RSpectra2016}.
+For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.
 
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=160pt]{conversion+QC.pdf}}
-\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
-\end{figure}
-
-Fast read/write operations from/to bed/bim/fam PLINK files are available.
-One should use PLINK to convert data to this format.
+Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this binary PLINK format.
 In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}).
@@
 
-We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage (use of one byte per element, 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'').
+We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
+Indeed, each element is stored on one byte which is 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
 With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{imputation.pdf}}
-\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
-\end{figure}
-
 \subsection{Population structure and SNP thinning based on Linkage Disequilibrium}
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{svd.pdf}}
-\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array.}\label{fig:svd}
-\end{figure}
-
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for the computation of a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other on randomized projections (Figure~\ref{fig:svd}).
+For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD, (Figure~\ref{fig:svd}).
 
@@
 FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
-The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.
+The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the ``big.matrix'' format, which enables parallel computations.
 
@@
 For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}.
-There are different options to thin SNPs based on Linkage Disequilibrium.
+There are at least 3 different options to thin SNPs based on Linkage Disequilibrium.
 The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.
@@
 The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
-Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs\footnote{https://goo.gl/8TngVE}.
+Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}.
 However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them.
 This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
-Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm.
-Indeed, a long-range LD region would cause SNPs in this region to have strong weights (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive.
+Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions by the algorithm.
+Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA.
 This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.
@@
 
-For association purposes, statistical tests based on linear and logistic regressions are available.
 Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
 Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
-
-$\forall j \in \{1, \dots, m\}$, 
-\itemize{
-\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
-\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $\hat{p} = \mathbb{P}(Y = 1)$.
-}\\
+Statistical tests based on linear and logistic regressions are also available.
+For the linear regression, for each SNP $j$, a T-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
+Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$ and $\hat{p} = \mathbb{P}(Y = 1)$.
+\\
 The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.
@@
 The R packages also implement functions to compute Polygenic Risk Scores. 
-
 First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
@@
 
-In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
+In this paper, two real datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
 The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
@@
 
-For computation timing purposes only, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structure of Linkage Disequilibrium and Population Structure as the original dataset.
+For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structures of Population and Linkage Disequilibrium as the original dataset.
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.
@@
 
-All the code used in this paper along with results, such as execution times and figures, are available as HTML notebooks in the Supplementary Data.
+All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data.
 
@@
 Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes.
-We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort.
+We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort.
 We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores.
@@
 We performed an association study and computed a polygenic risk score for the Celiac cohort. 
-
 The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
@@
 As illustrated in the supplementary data, the whole pipeline is user-friendly and takes only 20 lines of code.
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=200pt]{celiac-pca}}
-\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
-\end{figure}
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
-\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. The y-axis has been cut in order to not see only the very strong effects on chromosome 6.}\label{fig:gwas}
-\end{figure}
 
@@
 We first compared the GWAS and PRS computations with the R packages against PLINK 1.9 and EIGENSOFT 6.1.4.
-
-\begin{table}[!tpb]
-\centering
-\begin{adjustbox}{max width=235pt}
-\begin{tabular}{|c|cc|cc|}
-\hline
-Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\ 
-\hline
-Reading PLINK files & n/a & n/a & 5 / 20 sec & snp\_readBed \\
-Pruning & plink -{}-exclude -{}-range -{}-indep-pairwise & 4 / 6 sec & 13 / 46 sec & snp\_indLRLDR \& snp\_pruning \\
-\multirow{2}{*}{Computing 10 PCs} & plink -{}-extract -{}-make-bed & 2 sec & \multirow{2}{*}{45 / 136 sec} & \multirow{2}{*}{big\_randomSVD} \\
-& flashpcaR::flashpca (FlashPCA2) & 6 / 7 min &  &  \\
-GWAS (binary phenotype) & plink -{}-logistic -{}-covar & 5 min & 5 / 14.5 min & big\_univLogReg \\
-\hline
-Total & PLINK and flashpcaR & 11 / 12 min & 6 / 18 min & bigsnpr and bigstatsr \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
-The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
-\label{tab:bench-gwas}
-\end{table}
-
-\begin{table}[!tpb]
-\centering
-\begin{adjustbox}{max width=235pt}
-\begin{tabular}{|c|cc|cc|}
-\hline
-Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\
-\hline
-GWAS (binary phenotype) & plink -{}-keep -{}-logistic -{}-covar & 4 / min & 3 / min & big\_univLogReg \\
-Clumping & plink -{}-keep -{}-clump -{}-clump-p1
--{}-clump-p2 -{}-clump-r2 & 49 / sec & 9 / sec & snp\_clumping \\
-PRS & plink -{}-keep -{}-extract -{}-score -{}-q-score-range & 9 / sec & 4 / sec & snp\_PRS \\
-\hline
-Total & PLINK & 5 / min & 3 / min & bigsnpr and bigstatsr \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
-The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
-\label{tab:bench-prs}
-\end{table}
 
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=200pt]{benchmark-pca}}
-\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the computation time in minutes as a function of the number of samples.
-The computation corresponds to 10 Principal Components and 93,083 SNPs which remain after thinning.}\label{fig:bench-pca}
-\end{figure}
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=200pt]{precision-pca}}
-\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA) and package bigstatsr.
-It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the celiac disease dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
-\end{figure}
-
 \subsection{Automatic detection of long-range LD regions}
@@
 We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
-For the Celiac dataset, we found 5 long-range LD regions and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
+For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
 
@@
 The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}).
-We introduced missing values using a Beta-binomial distribution resulting in approximately 3\% of missing values.
+We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values.
 Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17).
@@
 We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=200pt]{error-impute}}
-\caption{Number of imputation errors vs the estimated number of imputation errors by SNP.
-For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
-The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
-\end{figure}
 
@@
 Technically, bigstatsr and bigsnpr could handle any size of datasets.
-However, if accesses demand the OS to often swap between the file and the memory, this would slow your analysis down.
+However, if accesses demand the OS to often swap between the file and the memory, this would slow down your analysis.
 For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times.
@@
 We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use.
+For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing PRS.
 As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files \cite[]{Palmer2016}.
 
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{conversion+QC.pdf}}
+\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
+\end{figure}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{imputation.pdf}}
+\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
+\end{figure}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{svd.pdf}}
+\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
+\end{figure}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{celiac-pca}}
+\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
+\end{figure}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
+\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. The y-axis has been cut in order to not see only the very strong effects on chromosome 6.}\label{fig:gwas}
+\end{figure}
+
+\newpage
+\begin{table}[!h]
+\centering
+\begin{adjustbox}{max width=\textwidth}
+\begin{tabular}{|c|c|c|}
+\hline
+\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times} \\
+ \cline{2-3}
+ & without R packages & with R packages \\
+\hline
+Reading PLINK files & n/a & 5 / 20 sec \\
+Pruning & 4 / 6 sec & 13 / 46 sec \\
+Computing 10 PCs & 6 / 7 min & 45 / 136 sec \\
+GWAS (binary phenotype) & 5 min & 5 / 14.5 min \\
+\hline
+Total & 11 / 12 min & 6 / 18 min \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
+The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\label{tab:bench-gwas}
+\end{table}
+
+\newpage
+\begin{table}[!h]
+\centering
+\begin{adjustbox}{max width=\textwidth}
+\begin{tabular}{|c|c|c|}
+\hline
+\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times} \\
+ \cline{2-3}
+ & without R packages & with R packages \\
+\hline
+GWAS (binary phenotype) & 4 / min & 3 / min \\
+Clumping & 49 / sec & 9 / sec \\
+PRS & 9 / sec & 4 / sec \\
+\hline
+Total & 5 / min & 3 / min \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
+The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\label{tab:bench-prs}
+\end{table}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{benchmark-pca}}
+\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
+It shows the computation time in minutes as a function of the number of samples.
+The computation corresponds to 10 Principal Components and 93,083 SNPs which remain after thinning.}\label{fig:bench-pca}
+\end{figure}
+
+\newpage
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{precision-pca}}
+\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA) and package bigstatsr.
+It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the celiac disease dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
+\end{figure}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{error-impute}}
+\caption{Number of imputation errors vs the estimated number of imputation errors by SNP.
+For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
+The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
+\end{figure}
+
+\newpage
 \section*{Acknowledgements}
@@
 
-\begin{table}[!tpb]
+\begin{table}[!h]
 \centering
@@
 % Tue Jul 18 15:59:51 2017
-\begin{table}[!tpb]
+\begin{table}[!h]
 \centering
@@
 % Tue Jun 13 11:29:13 2017
-\begin{table}[!tpb]
+\begin{table}[!h]
 \centering
@@
 % Tue Jun 13 11:25:56 2017
-\begin{table}[!tpb]
+\begin{table}[!h]
 \centering
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{scores}}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{scores}}
 \caption{PC4 and PC5 of the celiac disease dataset.
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=220pt]{loadings}}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{loadings}}
 \caption{Loadings of first 6 PCs of the celiac disease dataset plotted as hexbins (2-D histogram with hexagonal cells).
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=220pt]{gini}}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{gini}}
 \caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $R^2 > 0.2$) and with the automatic detection and removal of long-range LD regions.
@@
 
+\newpage
 \subsection{Imputation}
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=180pt]{hist-maf}}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{hist-maf}}
 \caption{Histogram of the minor allele frequencies of the POPRES dataset used for comparing imputation methods.}\label{fig:maf}
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=180pt]{hist-NA}}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{hist-NA}}
 \caption{Histogram of the number of missing values by SNP. These numbers were generated using a Beta-binomial distribution.}\label{fig:NA}

### 2017-07-22 imputation.Rmd
error? need rerun

@@
 
-ggsave("hist-NA.png", width = 1300, height = 900, scale = 1/75)
+ggsave("figures/hist-NA.png", width = 1300, height = 900, scale = 1/75)
 ```
@@
 GNA <- describe(XNA)
+```
+
+## Imputation with BEAGLE (called via bigsnpr)
+
+```{r}
+# Cleaning from previous runs 
+file.remove(list.files("backingfiles", pattern = "tmp*", full.names = TRUE))
+```
+
+```{r}
+# Write a new bedfile
+popresNA <- popres
+popresNA$genotypes <- GNA
+bedfile <- snp_writeBed(popresNA, bedfile = "backingfiles/tmp.bed")
+```
+
+```{r}
+system.time(
+  bedfile2 <- snp_beagleImpute(beagle.path = BEAGLE,
+                               plink.path = PLINK,
+                               bedfile.in = bedfile,
+                               memory.max = 30,
+                               ncores = NCORES)
+) # 14.6 / 14.2 / 14.0 hours
+```
+
+```{r}
+# Read bedfile into R and assess the errors of imputation
+popres.beagle <- snp_readBed(bedfile2, backingfile = "tmp_impute") %>%
+  snp_attach()
+X_beagle <- attach.BM(popres.beagle$genotypes)
+
+mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 / 4.8 %
+table(X_beagle[indNA], X[indNA])
 ```
@@
 
-ggsave("error-impute.png", width = 1000, height = 1000, scale = 1/75)
+ggsave("figures/error-impute.png", width = 1000, height = 1000, scale = 1/75)
 ```
-
-## Imputation with BEAGLE (called via bigsnpr)
-
-```{r}
-# Cleaning from previous runs 
-file.remove(list.files("backingfiles", pattern = "tmp*", full.names = TRUE))
-```
-
-```{r}
-# Write a new bedfile
-popresNA <- popres
-popresNA$genotypes <- GNA
-bedfile <- snp_writeBed(popresNA, bedfile = "backingfiles/tmp.bed")
-```
-
-```{r}
-system.time(
-  bedfile2 <- snp_beagleImpute(beagle.path = BEAGLE,
-                               plink.path = PLINK,
-                               bedfile.in = "backingfiles/tmp.bed",
-                               memory.max = 30,
-                               ncores = NCORES)
-) # 14.6 / 14.2 / 14.0 hours
-```
-
-```{r}
-# Read bedfile into R and assess the errors of imputation
-popres.beagle <- snp_readBed(bedfile2, backingfile = "tmp_impute") %>%
-  snp_attach()
-X_beagle <- attach.BM(popres.beagle$genotypes)
-
-mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 / 4.8 %
-table(X_beagle[indNA], X[indNA])
-```

### 2017-07-23 preprocessing.Rmd
run 2 or 4

@@
 ```{r}
-{{G@code <- bigsnpr:::CODE_IMPUTE_PRED}}
+# You need to change the code of G
+G@code <- bigsnpr:::CODE_IMPUTE_PRED
 big_counts(G, ind.col = 1:10)
 ```
+
+```{r}
+# To make this permanent, you need to save (modify) the file on disk
+celiac$genotypes@code <- bigsnpr:::CODE_IMPUTE_PRED
+saveRDS(celiac, "backingfiles/celiacQC.rds")
+```

### 2017-07-24 imputation.Rmd
third run

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 15)
+knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 15, fig.asp = 0.7)
 ```
@@
 
-```{r, message=FALSE}
+```{r, message=FALSE, warning=FALSE}
 library(bigsnpr)
@@
                                ncores = NCORES)
-) # 14.6 / 14.2 / 14.0 hours
+) # 14.6 / 14.8 /  hours
 ```
@@
 
-mean(X_beagle[indNA] != X[indNA]) # 3.1 / 4.8 / 4.8 %
+mean(X_beagle[indNA] != X[indNA]) # 3.1 / 3.1 / %
 table(X_beagle[indNA], X[indNA])
@@
   infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
-) # Previous runs: 42 / 48 / 48 minutes
+) # Previous runs: 42 / 47 /  minutes
 ```
 
-```{r}
+```{r, fig.asp=1}
 # Errors
-mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 / 4.8 %
+mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 /  %
 table(XNA[indNA], X[indNA])

### 2017-07-24 cabios-template/draft.tex
draft2

@@
 
-\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author[Privé \textit{et~al}.]{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
-$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
-$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
-
-$^\ast$To whom correspondence should be addressed.
+\title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
+\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
 
 \date{~ }
+\maketitle
 
+\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
+\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
+
+\noindent$^\ast$To whom correspondence should be addressed.
+
+\newpage
 \abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses.
+This increase in data size is a major challenge severely slowing down genomic analyses.
 Specialized software for every part of the analysis pipeline have been developed to handle large genomic data.
@@
 \\
-\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
+\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} and \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
 \textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 
-\maketitle
-
+\newpage
 \section{Introduction}
@@
 ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple variants \cite[]{Vilhjalmsson2015,Euesden2015}.
+Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple genetic variants \cite[]{Vilhjalmsson2015,Euesden2015}.
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
@@
 This is particularly a burden for commonly used analysis languages such as R, Python and Perl.
-Solving the memory issues for these languages would give access to a broad range of tools for data analysis, already implemented by the scientific community.
+Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented.
 Hopefully, strategies have been developed to avoid loading large datasets in RAM.
@@
 
-We are aware the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
+We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
 With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
@@
 All this is managed by the Operating System so that it is seamless and efficient.
-It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on.
+It means that if you use the same chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on.
 Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.
@@
 
-Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this binary PLINK format.
+Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this format.
 In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}).
@@
 We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
-Indeed, each element is stored on one byte which is 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
+Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
 With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
@@
 Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
-The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}.
+The first function is a wrapper to PLINK and Beagle which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
-This algorithm doesn't use phasing and is very fast.
+This algorithm doesn't use phasing and is very fast, as previous Machine Learning Approaches for genetic imputation \cite[]{Wang2012}.
 It only relies on some local XGBoost models.
@@
 
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD, (Figure~\ref{fig:svd}).
+For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
 
@@
 A variant of pruning is clumping.
-Clumping is useful if a statistic is available to sort the SNPs by importance (e.g.
-association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest.
+Clumping is useful if a statistic is available to sort the SNPs by importance, e.g.
+association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest.
 Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.
@@
 This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
-Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions by the algorithm.
+Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions.
 Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA.
@@
 Statistical tests based on linear and logistic regressions are also available.
-For the linear regression, for each SNP $j$, a T-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
-Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$ and $\hat{p} = \mathbb{P}(Y = 1)$.
-\\
-The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.
+For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
+Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
 
@@
 
-Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors.
+Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
 The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
@@
 
-In this paper, two real datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
+In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
 The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
@@
 
-For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structures of Population and Linkage Disequilibrium as the original dataset.
+For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset.
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.
@@
 First, we illustrate the application of R packages bigstatsr and bigsnpr.
-Secondly, we compared the performance of the packages against standard software.
+Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT).
 Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes.
@@
 We performed an association study and computed a polygenic risk score for the Celiac cohort. 
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
+The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
 For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
-For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with our new method (see section~\ref{sec:impute}).
-We note that if we used a standard R matrix to store the genotypes, this data would take 32GB of memory.
-On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.
+For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with the XGBoost method.
+If we used a standard R matrix to store the genotypes, this data would take 32GB of memory.
+On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB only.
 
-We used bigstatsr and bigsnpr R functions to get the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}).
+We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}).
 We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
-As illustrated in the supplementary data, the whole pipeline is user-friendly and takes only 20 lines of code.
+As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.
 
-The Celiac dataset is relatively small as compared to modern genetic cohorts.
+%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
 To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs.
@@
 
-We first compared the GWAS and PRS computations with the R packages against PLINK 1.9 and EIGENSOFT 6.1.4.
-
+We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
 For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.
@@
 
-On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 and the similar function implemented in bigstatsr, big\_randomSVD.
+On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr.
 For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs.
-We used the datasets of growing size simulated from the Celiac dataset (Section~\ref{sec:rep}).
+We used the datasets of growing size simulated from the Celiac dataset.
 Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure~\ref{fig:bench-pca}).
-We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition (e.g.
-SmartPCA).
+We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA.
 FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
@@
 When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure~\ref{fig:scores}).
-Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}) which is confirmed by Gini coefficients (measure of dispersion) of each squared loadings reported in Figure~\ref{fig:gini}, which are all around the theoretical value of $2 / pi$ (for gaussian loadings) and significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed.
+Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}).
+This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
 
@@
 We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values.
-Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17).
+Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017).
 Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
 For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
-We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
+We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
 
@@
 Technically, bigstatsr and bigsnpr could handle any size of datasets.
-However, if accesses demand the OS to often swap between the file and the memory, this would slow down your analysis.
+However, if accesses demand the OS to often swap between the file and the memory, this would slow down analysis.
 For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times.
@@
 Using a matrix-like format makes it easy to develop new functions in order to experiment and develop new ideas.
-Integration in R makes it possible to take advantage of all what R has to offer, for example using the excellent machine learning algorithm XGBoost to easily make a fast yet accurate imputation algorithm for genotyped SNPs.
-Other functions, not presented here, are also available and all the functions available within the package bigstatsr are not specific to SNP arrays, so that they could be used for other omic data or in completely different fields of research.
+Integration in R makes it possible to take advantage of all what R has to offer, for example using the widely-used machine learning algorithm XGBoost to easily make a fast and accurate imputation algorithm for genotyped SNPs.
+Other functions, not presented here, are also available and all the functions available within the package bigstatsr are not specific to SNP arrays, so that they could be used for other omic data or in other fields of research.
 
@@
 \centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
-\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. The y-axis has been cut in order to not see only the very strong effects on chromosome 6.}\label{fig:gwas}
+\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr.
+Some SNPs in chromosome 6 have p-values smaller than the $10^{-10}$ threshold used for vizualisation purposes.}\label{fig:gwas}
 \end{figure}
@@
  \cline{2-3}
- & without R packages & with R packages \\
+ & PLINK and FastPCA & bigstatsr and bigsnpr \\
 \hline
@@
 \end{adjustbox}
-\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
-The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
+The first time is with a desktop computer and the second time is with a laptop computer.}
 \label{tab:bench-gwas}
@@
  \cline{2-3}
- & without R packages & with R packages \\
+ & PLINK & bigstatsr and bigsnpr \\
 \hline
@@
 \end{adjustbox}
-\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset.
-The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for the Celiac dataset.
+The first time is with a desktop computer and the second time is with a laptop computer.}
 \label{tab:bench-prs}
@@
 
-\newpage
 \begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{precision-pca}}
-\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA) and package bigstatsr.
-It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the celiac disease dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
-\end{figure}
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{error-impute}}
-\caption{Number of imputation errors vs the estimated number of imputation errors by SNP.
-For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
-The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
+\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
+It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
 \end{figure}
@@
 
-Text Text Text Text Text Text  Text Text.
+Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
 
-\section*{Funding}
-
-This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
+\vspace*{-12pt}
 
@@
 \section*{Supplementary Data}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
 
@@
 
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\textwidth]{error-impute}}
+\caption{Number of imputation errors vs the estimated number of imputation errors by SNP.
+For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
+The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
+\end{figure}
+
 \end{document}

### 2017-07-24 imputation.Rmd
draft2

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 15, fig.asp = 0.7)
+knitr::opts_chunk$set(fig.align = "center", out.width = "70%",
+                      fig.width = 15, fig.asp = 0.7)
 ```
@@
                                ncores = NCORES)
-) # 14.6 / 14.8 /  hours
+) # 14.6 / 14.8 / 14.8 hours
 ```
@@
 
-mean(X_beagle[indNA] != X[indNA]) # 3.1 / 3.1 / %
+mean(X_beagle[indNA] != X[indNA]) # 3.1 / 3.1 / 3.1 %
 table(X_beagle[indNA], X[indNA])
@@
   infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
-) # Previous runs: 42 / 47 /  minutes
+) # Previous runs: 42 / 47 / 47 minutes
 ```
@@
 # Errors
-mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 /  %
+mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 / 4.8 %
 table(XNA[indNA], X[indNA])

### 2017-07-26 GWAS-biobank-scale.Rmd
biobank scale

---
title: "Imputation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 15, fig.asp = 0.7)
```

## Form a larger dataset

```{r}
library(bigsnpr)

celiac <- snp_attach("backingfiles/celiacQC.rds") 
G <- celiac$genotypes

system.time(
  bedfile <- snp_writeBed(celiac, bedfile = "backingfiles/tmp_biobank.bed", 
                          ind.row = sort(rep_len(rows_along(G), 500e3)),
                          ind.col = sort(rep_len(cols_along(G), 1e6)))
)

# Size if the ".bed" file in GB
file.size(bedfile) / 1024^3
```

## Make a GWAS on this dataset

### Read the dataset in bigSNP format

```{r}
system.time(
  rdsfile <- snp_readBed(bedfile, backingfile = "tmp_biobank")
) # 1.6h
```

```{r}
bkfile <- sub("\\.rds$", ".bk", rdsfile)
# Size if the ".bk" file in GB
file.size(bkfile) / 1024^3
```

```{r}
system.time(
  biobank <- snp_attach(rdsfile)
)
```

### Clumping with MAF before PCA

```{r}
NCORES <- parallel::detectCores() / 2
# Get SNPs in long-range LD regions
system.time(
  ind.excl <- snp_indLRLDR(biobank$map$chromosome,
                           biobank$map$physical.pos)
)
# No parallelization for this huge dataset
system.time(
  ind.keep <- snp_clumping(biobank$genotypes, 
                           infos.chr = biobank$map$chromosome,
                           exclude = ind.excl,
                           ncores = 1)
) # 6.3h
```

### Computation of first 10 PCs

```{r}
# Parallelization is OK for PCA because 
# only a subset of this huge dataset is used 
system.time(
  obj.svd <- big_randomSVD(biobank$genotypes, 
                           fun.scaling = snp_scaleBinom(),
                           ind.col = ind.keep,
                           ncores = NCORES)
) # 2.4h
```

### GWAS on a continuous variable

```{r}
# Simulate some continous phenotype
y <- rnorm(nrow(biobank$genotypes))

# No parallelization for this huge dataset
system.time(
  gwas <- big_univLinReg(biobank$genotypes,
                         y.train = y,
                         covar.train = obj.svd$u, 
                         ncores = 1)
) # 3h
```

```{r}
snp_manhattan(gwas, biobank$map$chromosome, biobank$map$physical.pos,
              npoints = 50e3, coeff = 2)
```

### 2017-07-27 GWAS-biobank-scale.Rmd
biobank-scale

@@
 ---
-title: "Imputation"
+title: "GWAS on a biobank-scale dataset"
 output: html_document
@@
   rdsfile <- snp_readBed(bedfile, backingfile = "tmp_biobank")
-) # 1.6h
+) # 1.6h / 1h
 ```
@@
                            ncores = 1)
-) # 6.3h
+) # 6.3h / 5.4h
 ```
@@
                            ncores = NCORES)
-) # 2.4h
+) # 2.4h / 1.9h
 ```

### 2017-07-30 benchmark-pca.Rmd
why bug? Error in mutate_impl(.data, dots): Evaluation error: l'argument est de longueur nulle.

---
title: "Benchmarking Principal Component Analysis software"
output: html_document
---

In this notebook, we will benchmark three different methods for computing first Principal Components (PCs) of a large genotype matrix.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7, error = TRUE)
options(width = 80)
```

## Get the data

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # devtools::install_github("gabraham/flashpca", subdir = "flashpcaR")
  library(flashpcaR)
  # install.packages("tidyverse")
  library(tidyverse)
  # install.packages("glue")
  library(glue)   
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"

NCORES <- parallel::detectCores() / 2

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes

# Thinning of SNPs
ind.excl <- snp_indLRLDR(celiac$map$chromosome,
                         celiac$map$physical.pos)
length(
  ind.keep <- snp_clumping(celiac$genotypes, celiac$map$chromosome,
                         exclude = ind.excl, 
                         ncores = NCORES)
)
```

```{r, include=FALSE}
# Some cleaning
file.remove(list.files(path = "backingfiles/", 
                       pattern = "celiacQC_keep*", 
                       full.names = TRUE))
```

```{r}
# Write new bed files keeping only a subset of SNPs and replicating individuals
snp_writeBed(celiac, "backingfiles/celiacQC_keep1.bed",
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep1.bed", "celiacQC_keep1")

snp_writeBed(celiac, "backingfiles/celiacQC_keep5.bed",
             ind.row = rep(rows_along(G), each = 5),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep5.bed", "celiacQC_keep5")

snp_writeBed(celiac, "backingfiles/celiacQC_keep10.bed",
             ind.row = rep(rows_along(G), each = 10),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep10.bed", "celiacQC_keep10")
```

## Functions for computing PCs

```{r}
# FlashPCA2: devtools::install_github("gabraham/flashpca", subdir = "flashpcaR")
flashpca <- function(prefix) {
  time <- system.time(
    svd <- flashpcaR::flashpca(prefix)
  )
  list(time[3], svd$vectors)
}

fastpca <- function(prefix, fast = TRUE) {
  parfile <- glue("{prefix}.pca.par")

  ## Renaming so that Eigensoft can recognize file type
  system(glue("cp {prefix}.bim {prefix}.pedsnp"))  
  ## Renaming so that Eigensoft can recognize file type
  system(glue("cp {prefix}.fam {prefix}.pedind"))  
  writeLines(glue(
    "genotypename: {prefix}.bed
    snpname: {prefix}.pedsnp
    indivname: {prefix}.pedind
    evecoutname: {prefix}.pca.evec
    evaloutname: {prefix}.eval
    altnormstyle: NO
    numoutevec: 10
    numoutlieriter: 0
    numoutlierevec: 10
    outliersigmathresh: 6
    qtmode: 0
    fastmode: {as.integer(fast)}"
  ), con = parfile)
  ## PCA
  time <- system.time(
    system(glue("{smartpca} -p {parfile}"), ignore.stdout = TRUE)
  ) 
  ## removing temp files
  system(glue("rm {prefix}.pedsnp")) 
  system(glue("rm {prefix}.pedind"))
  system(glue("rm {parfile}"))

  evec <- data.table::fread(glue("{prefix}.pca.evec"), 
                            data.table = FALSE, skip = 1)

  list(time[3], as.matrix(evec[, 2:11]))
}

# bigstatsr::big_randomSVD
mypca <- function(prefix, ncores = 1) {
  rdsfile <- paste0(prefix, ".rds")
  bigsnp <- snp_attach(rdsfile)

  time <- system.time(
    svd <- big_randomSVD(bigsnp$genotypes, 
                         fun.scaling = snp_scaleBinom(),
                         ncores = ncores)
  )

  list(time[3], svd$u)
}
```

## Evaluating all models

```{r}
# Preparing a dataset for the results
models <- tribble(
  ~method,               ~fun,     ~params,
  "FastPCA/EIGENSOFT",   fastpca,  list(),
  "FlashPCA2",           flashpca, list(),
  "bigstatsr - 1 core",  mypca,    list(),
  "bigstatsr - 6 cores", mypca,    list(ncores = NCORES)
) %>%
  mutate(size = list(c(1, 5, 10))) %>%
  unnest(size, .drop = FALSE) %>%
  mutate(time = NA, pcs = list(NA))

# Filling this data frame with time and PC scores for each method and dataset
for (i in rows_along(models)) { 
  size <- models$size[[i]]
  prefix <- glue("backingfiles/celiacQC_keep{size}")
  fun <- models$fun[[i]]
  params <- c(prefix, models$params[[i]])
  res <- do.call(fun, args = params)
  models[["time"]][[i]] <- res[[1]]
  models[["pcs"]][[i]] <- res[[2]]
}
```

## Plotting benchmark

```{r}
(models <- mutate(models, size = size * nrow(G)))
```

```{r}
models %>%
  ggplot(aes(size / 1000, time / 60, color = method)) %>%
  bigstatsr:::MY_THEME(coeff = 2.5) + 
  geom_point(aes(shape = method), cex = 8) +
  geom_line(aes(linetype = method), lwd = 3) + 
  theme(legend.position = c(0.25, 0.75),
        legend.key.width = unit(8, "line")) + 
  labs(x = "Number of individuals (x1000)", y = "Time (in minutes)", 
       color = "Methods:", linetype = "Methods:", shape = "Methods:")

ggsave("figures/benchmark-pca.png", width = 1271, height = 945, scale = 1/75)
```

## Comparing precision of results

```{r}
true <- fastpca("backingfiles/celiacQC_keep1", fast = FALSE)
```

```{r}
# Tidy and plot results
(preci <- models %>% 
  mutate(
    size = as.factor(size),
    cor = lapply(pcs, function(x) {
      N <- nrow(x) / nrow(true)
      if (N == 1) {
        diag(cor(x, true))
      } else {
        diag(cor(x[c(TRUE, rep(FALSE, N - 1)), ], true))
      }
    }),
    PC = list(as.factor(1:10))
  ) %>%
  select(-c(fun, params, time, pcs)))

preci %>%
  unnest(cor, PC) %>%
  ggplot(aes(PC, cor^2, color = size, shape = size)) %>%
  bigstatsr:::MY_THEME(coeff = 2) +
  geom_point(cex = 5) +
  geom_line(aes(group = size, linetype = size), lwd = 2) +
  facet_wrap(~method, nrow = 2) +
  labs(x = "Principal Component", y = "Squared correlation between estimated and true PCs",
       color = "Number of individuals", shape = "Number of individuals", linetype = "Number of individuals") + 
  theme(legend.position = c(0.8, 0.15),
        strip.text.x = element_text(size = rel(3)),
        legend.key.width = unit(4.5, "line")) + 
  coord_cartesian(ylim = c(0, 1))

ggsave("figures/precision-pca.png", width = 1200, height = 900, scale = 1/75)
```

### 2017-07-30 benchmark-pca.Rmd
benchmark pca OK

@@
                       fig.width = 16, fig.asp = 0.7, error = TRUE)
-options(width = 80)
+options(width = 85)
 ```
@@
 ```{r}
-true <- fastpca("backingfiles/celiacQC_keep1", fast = FALSE)
+# SmartPCA
+smartpca <- fastpca("backingfiles/celiacQC_keep1", fast = FALSE)
+true <- smartpca[[2]]
+cat(sprintf("Standard PCA takes forever: %.2gh.", smartpca[[1]] / 3600))
 ```
@@
 ```
+
+```{r, include=FALSE}
+library(gmailr)
+send_message(
+  mime(To = "florian.prive.21@gmail.com",
+       From = "florian.prive.21@gmail.com",
+       Subject = sprintf("Knitting of %s has finished",
+                         knitr::current_input()))
+)
+```

### 2017-07-30 autoSVD.Rmd
autoSVD OK

---
title: "Auto SVD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7, error = TRUE)
options(width = 85)
```

## Get the data

```{r}
suppressMessages({
  library(bigsnpr)
  library(tidyverse)
  library(foreach)
})

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- parallel::detectCores() / 2
```

## Without removing long-range LD regions

```{r}
system.time(
  ind.keep <- snp_clumping(G, CHR, ncores = NCORES)
)
system.time(
  obj.svd <- big_randomSVD(G, snp_scaleBinom(), 
                           ind.col = ind.keep,
                           ncores = NCORES)
)
```

```{r, fig.asp=1}
(plot_loadingsA <- plot(obj.svd, type = "loadings", loadings = 1:6, coeff = 1))
```

```{r}
oneSNP <- which.max(abs(obj.svd$v[, 4]))
(plot_scoresA <- plot(obj.svd, type = "scores", scores = 4:5, coeff = 2) +
    aes(color = as.factor(attach.BM(G)[, ind.keep[oneSNP]])) + 
    labs(title = NULL, color = "Genotype") +
    theme(legend.position = c(0.15, 0.15)) +
    guides(color = guide_legend(override.aes = list(size = 4))))
```

## Automatically removing long-range LD regions

```{r}
system.time(
  test <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
)
htmlTable::htmlTable(attr(test, "lrldr"), rnames = FALSE,
                     css.table = "margin-top: 1em; 
                                  margin-bottom: 1em;   
                                  margin-left: auto;
                                  margin-right: auto",
                     css.cell = "padding: 5px 5px 5px 5px")
```

```{r, comment=""}
attr(test, "lrldr") %>%
  transmute(
    Chromosome = Chr,
    `Start (Mb)` = Start / 1e6,
    `Stop (Mb)` = Stop / 1e6
  ) %>%
  xtable::xtable(digits = 1, caption = "", label = "", auto = TRUE)
```

```{r, fig.asp=1}
(plot_loadingsB <- plot(test, type = "loadings", loadings = 1:6, coeff = 1))
```

```{r}
# Get population from external files
pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/",
                        pattern = "cluster_*",
                        full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")

(plot_scoresB <- plot(test, type = "scores", scores = 4:5, coeff = 2) +
    aes(color = pop.names[pop]) + 
    labs(title = NULL, color = "Population") +
    theme(legend.position = c(0.2, 0.75)) +
    guides(color = guide_legend(override.aes = list(size = 4))))
```

```{r}
# save plots for paper
p <- cowplot::plot_grid(plot_loadingsA, plot_loadingsB, align = "hv", 
                        ncol = 2, labels = c("A", "B"), label_size = 30)
ggsave("figures/loadings.png", width = 1620, height = 920, scale = 1/75)

p <- cowplot::plot_grid(plot_scoresA, plot_scoresB, align = "hv", 
                        ncol = 2, labels = c("A", "B"), label_size = 30)
ggsave("figures/scores.png", width = 1640, height = 750, scale = 1/75)
```

```{r, fig.asp=0.7, fig.width=18}
doParallel::registerDoParallel(NCORES)

nBoot <- 500

tribble(
  ~name,                                               ~svd,
  "When automatically removing long-range LD regions", test, 
  "When keeping long-range LD regions",                obj.svd
) %>%
  apply(1, function(method) {
    V <- method[["svd"]][["v"]]^2
    m <- nrow(V)
    foreach(ic = seq_len(nBoot), .combine = 'rbind') %dopar% {
      ind <- sort(sample(m, replace = TRUE))
      apply(V[ind, ], 2, ineq::Gini)
    } %>%
      reshape2::melt() %>%
      cbind(method = method[["name"]])
  }) %>%
  Reduce(f = 'rbind') %>%
  ggplot(aes(as.factor(Var2), value)) %>%
  bigstatsr:::MY_THEME(coeff = 2) +
  geom_hline(yintercept = 2 / pi, lty = 2) +
  geom_boxplot(aes(color = method), lwd = 1.5) +
  labs(y = "Gini coefficient of squared loadings", 
       x = "Principal Component", color = "") +
  theme(legend.position = "top") +
  guides(color = guide_legend(ncol = 1))

doParallel::stopImplicitCluster()

ggsave("figures/gini.png", width = 1341, height = 865, scale = 1/75)
```

## Removing long-range LD regions based on predefined table

```{r}
ind.excl0 <- snp_indLRLDR(CHR, POS)
ind.keep0 <- snp_clumping(G, CHR, exclude = ind.excl0, ncores = NCORES)
obj.svd0 <- big_randomSVD(G, snp_scaleBinom(), 
                          ind.col = ind.keep0,
                          ncores = NCORES)
```

```{r}
corr <- cor(test$u, obj.svd0$u)
rownames(corr) <- colnames(corr) <- paste0("PC", 1:10)
htmlTable::htmlTable(round(100 * corr, 2), 
                     align = paste(c("|", rep("c", 10)), collapse = ""),
                     css.table = "margin-top: 1em; 
                                  margin-bottom: 1em;   
                                  margin-left: auto;
                                  margin-right: auto",
                     css.cell = "padding: 5px 5px 5px 5px")
```

```{r}
mean(diag(abs(corr)))
```

```{r, include=FALSE}
library(gmailr)
send_message(
  mime(To = "florian.prive.21@gmail.com",
       From = "florian.prive.21@gmail.com",
       Subject = sprintf("Knitting of %s has finished",
                         knitr::current_input()))
)
```

### 2017-07-30 compare-GWAS-PRS-desktop.Rmd
remains

---
title: "Comparing bigstatsr and bigsnpr with PLINK and EIGENSOFT"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7)
options(width = 85)
```

## Prerequisites

Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file containing 15,250 individuals and 281,122 SNPs.

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # install.packages("glue")
  library(glue)      
  # install.packages("ggplot2")
  library(ggplot2)
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"

write.table2 <- bigsnpr:::write.table2
fread2 <- function(...) data.table::fread(data.table = FALSE, ...)

NCORES <- parallel::detectCores() / 2
plink <- "../plink_linux_x86_64/plink"
```

## Example of GWAS

In this section, we will show how to perform a GWAS with or without our packages. For simplicity, everything will be done inside R.

### With bigstastr and bigsnpr

```{r}
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos

# Pruning
system.time(
  ind.keep <- snp_pruning(G, infos.chr = CHR, 
                          exclude = snp_indLRLDR(CHR, POS),
                          ncores = NCORES)
)
```

```{r}
# PCA
system.time(
  obj.svd <- big_randomSVD(G, snp_scaleBinom(), 
                           ind.col = ind.keep,
                           ncores = NCORES)
)

# Get population from external files
pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/",
                        pattern = "cluster_*",
                        full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
# Plot PC1 & PC2
plot(obj.svd, type = "scores", coeff = 2) +
  aes(color = pop.names[pop]) +
  labs(color = "Population", title = NULL) +
  theme(legend.position = c(0.2, 0.75)) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))

ggsave("figures/celiac-pca.png", width = 1124, height = 945, scale = 1/75)
```

```{r}
# GWAS
system.time(
  gwas <- big_univLogReg(G, y01.train = celiac$fam$affection - 1,
                         covar.train = obj.svd$u, ncores = NCORES)
)

labels <- 1:22; labels[c(15, 17, 19, 21)] <- ""
snp_manhattan(snp_gc(gwas), CHR, POS, 
              coeff = 2, labels = labels) + 
  geom_hline(yintercept = -log10(5e-8), col = "red") +
  labs(title = NULL) +
  coord_cartesian(ylim = c(0, 30))

ggsave("figures/celiac-gwas-cut.png", width = 1475, height = 788, scale = 1/75)
```

### With PLINK and EIGENSOFT (and R)

```{r}
# Pruning
snp_writeBed(celiac, bedfile = "plink0.bed")
prefix_bed <- "plink0"
write.table2(LD.wiki34, "LRLDR.txt")
system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --exclude LRLDR.txt --range",
              " --indep-pairwise 50 1 0.2",
              " --threads {NCORES}"))
)

ind.keep2 <- match(scan("plink.prune.in", what = ""), celiac$map$marker.ID)
mean(ind.keep %in% ind.keep2)

# Make a pruned dataset
write.table2(as.data.frame(celiac$map[ind.keep, "marker.ID"]), "keep.in")
system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --extract keep.in --make-bed",
              " --threads {NCORES}"))
)

# PCA with EIGENSOFT
fastpca <- function(prefix, fast = TRUE) {
  parfile <- glue("{prefix}.pca.par")

  ## Renaming so that Eigensoft can recognize file type
  system(glue("cp {prefix}.bim {prefix}.pedsnp"))  
  ## Renaming so that Eigensoft can recognize file type
  system(glue("cp {prefix}.fam {prefix}.pedind"))  
  writeLines(glue(
    "genotypename: {prefix}.bed
    snpname: {prefix}.pedsnp
    indivname: {prefix}.pedind
    evecoutname: {prefix}.pca.evec
    evaloutname: {prefix}.eval
    altnormstyle: NO
    numoutevec: 10
    numoutlieriter: 0
    numoutlierevec: 10
    outliersigmathresh: 6
    qtmode: 0
    fastmode: {as.integer(fast)}"
  ), con = parfile)
  ## PCA
  system(glue("{smartpca} -p {parfile}"), ignore.stdout = TRUE)
  ## removing temp files
  system(glue("rm {prefix}.pedsnp")) 
  system(glue("rm {prefix}.pedind"))
  system(glue("rm {parfile}"))

  evec <- data.table::fread(glue("{prefix}.pca.evec"), 
                            data.table = FALSE, skip = 1)

  as.matrix(evec[, 2:11])
}

system.time(
  obj.svd2 <- fastpca("plink")
)

# FastPCA is not accurate enough (see benchmark-pca.html)
round(100 * cor(obj.svd2, obj.svd$u), 1)
```

```{r}
# GWAS
write.table2(cbind(celiac$fam[, 1:2], obj.svd$u), "covar.txt")
system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --logistic hide-covar",
              " --covar covar.txt",
              " --threads {NCORES}"))
)

gwas2 <- fread2("plink.assoc.logistic")
all.equal(gwas2$P, 10^predict(gwas)) 
```

## Example of Polygenic Risk Scores (PRS)

In this section, we will show how to create a PRS with or without our packages. For simplicity, everything will be done inside R.

```{r}
# Split in training/test set (80%/20%)
ind.train <- sort(sample(nrow(G), 0.8 * nrow(G)))
ind.test <- setdiff(rows_along(G), ind.train)
```

### With bigstatsr and bigsnpr

```{r}
# GWAS on training set
y01 <- celiac$fam$affection - 1
system.time(
  gwas.train <- big_univLogReg(G, y01.train = y01[ind.train],
                               ind.train = ind.train, 
                               covar.train = obj.svd$u[ind.train, ], 
                               ncores = NCORES)
)

# Clumping
system.time(
  ind.keep <- snp_clumping(G, infos.chr = celiac$map$chromosome,
                           ind.row = ind.train,
                           S = abs(gwas.train$score),
                           size = 250, # as PLINK default
                           is.size.in.bp = TRUE,
                           infos.pos = celiac$map$physical.pos,
                           ncores = NCORES)
)

# PRS
(thrs <- c(0, exp(seq(log(0.2), log(550), length.out = 50))))
lpS <- -predict(gwas.train)
system.time(
  prs <- snp_PRS(G, betas = gwas.train$estim,
                 ind.test = ind.test,
                 ind.keep = ind.keep,
                 lpS = lpS, 
                 thr.list = thrs)
)

# AUC
aucs <- apply(prs, 2, AUC, target = y01[ind.test])
```

### With PLINK (and R)

```{r}
# Get a file with only training set
write.table2(cbind(celiac$fam[ind.train, 1:2], 
                   obj.svd$u[ind.train, ]), "covar_train.txt")
# GWAS on training set
system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --keep covar_train.txt",
              " --logistic hide-covar beta",
              " --covar covar_train.txt",
              " --threads {NCORES}"))
)

gwas2 <- fread2("plink.assoc.logistic")
all.equal(gwas2$P, 10^predict(gwas.train))

# Clumping
system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --keep covar_train.txt",
              " --clump plink.assoc.logistic",
              " --clump-p1 1 --clump-p2 1 --clump-r2 0.2",
              " --threads {NCORES}"))
)

clump <- fread2("plink.clumped")
ind.keep2 <- match(clump$SNP, celiac$map$marker.ID)
mean(ind.keep %in% ind.keep2)

gwas2.keep <- gwas2[ind.keep, ]
write.table2(gwas2.keep, "assoc_clumped.txt")

# PRS
write.table2(celiac$fam[ind.test, 1:2], "ind_test.txt")
write.table2(cbind(paste0("S", seq_along(thrs)), 0, 10^(-thrs)),
             "range.txt")

system.time(
  system(glue("{plink} --bfile {prefix_bed}",
              " --keep ind_test.txt",
              " --extract assoc_clumped.txt",
              " --score assoc_clumped.txt 2 4 7 sum",
              " --q-score-range range.txt assoc_clumped.txt 2 9",
              " --threads {NCORES}"))
)

scores <- matrix(NA_real_, length(ind.test), length(thrs))
for (i in seq_along(thrs)) {
  scores[, i] <- fread2(glue("plink.S{i}.profile"))$SCORESUM
}

#AUC
aucs2 <- apply(scores, 2, AUC, target = y01[ind.test])

# Number of predictor for each threshold
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
bigstatsr:::MY_THEME(qplot(nb.pred, aucs, color = "A", log = "x", 
                           cex = I(3), pch = I(1),
                           xlab = "#predictors", ylab = "AUC"), coeff = 2) +
  geom_line(show.legend = FALSE) + 
  geom_point(aes(nb.pred, aucs2, color = "B"), cex = 3, pch = 3) + 
  scale_colour_manual(
    name = NULL, 
    values = c("black", "red"),
    labels = c("PRS with bigstastr\nand bigsnpr", "PRS with plink")) + 
  guides(color = guide_legend(override.aes = list(shape = c(1, 3)))) + 
  theme(legend.position = c(0.7, 0.2)) 
```

```{r}
# Cleaning
file.remove(list.files(pattern = "plink*|*.txt"))
```

### 2017-07-31 compare-GWAS-PRS-laptop.Rmd
laptop OK

--- 2017-07-30 compare-GWAS-PRS-desktop.Rmd
@@
 # Get population from external files
-pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/",
+pop.files <- list.files(path = "../Dubois2010_data/",
                         pattern = "cluster_*",
@@
 # Cleaning
-file.remove(list.files(pattern = "plink*|*.txt"))
+file.remove(c(list.files(pattern = "plink*|*.txt"), "keep.in"))
 ```

### 2017-09-06 GWAS-biobank-scale.Rmd
use FBMs

@@
 system.time(
-  rdsfile <- snp_readBed(bedfile, backingfile = "tmp_biobank")
+  rdsfile <- snp_readBed(bedfile, backingfile = "backingfiles/tmp_biobank")
 ) # 1.6h / 1h
@@
 ```{r}
-NCORES <- parallel::detectCores() / 2
+NCORES <- nb_cores()
 # Get SNPs in long-range LD regions

### 2017-09-06 autoSVD.Rmd
use FBMs

@@
 knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
-                      fig.width = 16, fig.asp = 0.7, error = TRUE)
+                      fig.width = 16, fig.asp = 0.7)
 options(width = 85)
@@
 POS <- celiac$map$physical.pos
-NCORES <- parallel::detectCores() / 2
+NCORES <- nb_cores()
 ```
@@
 (plot_scoresA <- plot(obj.svd, type = "scores", scores = 4:5, coeff = 2) +
-    aes(color = as.factor(attach.BM(G)[, ind.keep[oneSNP]])) + 
+    aes(color = as.factor(G[, ind.keep[oneSNP]])) + 
     labs(title = NULL, color = "Genotype") +

### 2017-09-06 benchmark-pca.Rmd
use FBMs

@@
 
-NCORES <- parallel::detectCores() / 2
+NCORES <- nb_cores()
 
@@
              ind.col = ind.keep)
-snp_readBed("backingfiles/celiacQC_keep1.bed", "celiacQC_keep1")
+snp_readBed("backingfiles/celiacQC_keep1.bed", "backingfiles/celiacQC_keep1")
 
@@
              ind.col = ind.keep)
-snp_readBed("backingfiles/celiacQC_keep5.bed", "celiacQC_keep5")
+snp_readBed("backingfiles/celiacQC_keep5.bed", "backingfiles/celiacQC_keep5")
 
@@
              ind.col = ind.keep)
-snp_readBed("backingfiles/celiacQC_keep10.bed", "celiacQC_keep10")
+snp_readBed("backingfiles/celiacQC_keep10.bed", "backingfiles/celiacQC_keep10")
 ```

### 2017-09-06 compare-GWAS-PRS-desktop.Rmd
compare desktop OK
use FBMs

@@
 
-Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file containing 15,250 individuals and 281,122 SNPs.
+Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file.
 
@@
 smartpca <- "../EIG-6.1.4/bin/smartpca"
+plink <- download_plink()
 
@@
 
-NCORES <- parallel::detectCores() / 2
-plink <- "../plink_linux_x86_64/plink"
+NCORES <- nb_cores()
 ```
@@
 gwas2 <- fread2("plink.assoc.logistic")
-all.equal(gwas2$P, 10^predict(gwas)) 
+all.equal(gwas2$P, predict(gwas, log10 = FALSE)) 
 ```
@@
 system.time(
-  prs <- snp_PRS(G, betas = gwas.train$estim,
+  prs <- snp_PRS(G, betas.keep = gwas.train$estim[ind.keep],
                  ind.test = ind.test,
                  ind.keep = ind.keep,
-                 lpS = lpS, 
+                 lpS.keep = lpS[ind.keep], 
                  thr.list = thrs)

### 2017-09-06 imputation.Rmd
use FBMs

@@
 
-NCORES <- parallel::detectCores() / 2
-PLINK <- "../plink_linux_x86_64/plink"
-BEAGLE <- "../../beagle.21Jan17.6cc.jar"
+NCORES <- nb_cores()
 ```
@@
 # Fill a copy of the matrix with NAs (coded as 03)
-X <- attach.BM(G)
-XNA <- as.BM.code(deepcopy(X), code = bigsnpr:::CODE_IMPUTE_PRED)
-XNA[indNA] <- as.raw(3)
-
-GNA <- describe(XNA)
+GNA <- big_copy(G)
+GNA[indNA] <- as.raw(3)
 ```
@@
 ```{r}
-# Cleaning from previous runs 
-file.remove(list.files("backingfiles", pattern = "tmp*", full.names = TRUE))
-```
-
-```{r}
 # Write a new bedfile
@@
 popresNA$genotypes <- GNA
-bedfile <- snp_writeBed(popresNA, bedfile = "backingfiles/tmp.bed")
+bedfile <- snp_writeBed(popresNA, bedfile = tempfile(fileext = ".bed"))
 ```
@@
 system.time(
-  bedfile2 <- snp_beagleImpute(beagle.path = BEAGLE,
-                               plink.path = PLINK,
-                               bedfile.in = bedfile,
+  bedfile2 <- snp_beagleImpute(bedfile.in = bedfile,
                                memory.max = 30,
@@
 # Read bedfile into R and assess the errors of imputation
-popres.beagle <- snp_readBed(bedfile2, backingfile = "tmp_impute") %>%
+popres.beagle <- snp_readBed(bedfile2, backingfile = tempfile()) %>%
   snp_attach()
-X_beagle <- attach.BM(popres.beagle$genotypes)
+G.beagle <- popres.beagle$genotypes
 
-mean(X_beagle[indNA] != X[indNA]) # 3.1 / 3.1 / 3.1 %
-table(X_beagle[indNA], X[indNA])
+mean(G.beagle[indNA] != G[indNA]) # 3.1 / 3.1 / 3.1 %
+table(G.beagle[indNA], G[indNA])
 ```
@@
 system.time(
-  infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, ncores = NCORES)
+  infos <- snp_fastImpute(GNA, infos.chr = popres$map$chromosome, 
+                          ncores = NCORES)
 ) # Previous runs: 42 / 47 / 47 minutes
@@
 # Errors
-mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 / 4.8 %
-table(XNA[indNA], X[indNA])
+GNA$code256 <- bigsnpr:::CODE_IMPUTE_PRED
+mean(GNA[indNA] != G[indNA]) # 4.7 / 4.8 / 4.8 %
+table(GNA[indNA], G[indNA])
 
 # Plot the estimation of the errors of imputation
-data_frame(col = indNA[, 2], error = XNA[indNA] != X[indNA]) %>%
+data_frame(col = indNA[, 2], error = GNA[indNA] != G[indNA]) %>%
   group_by(col) %>%

### 2017-09-06 preprocessing.Rmd
use FBMs

--- 2017-07-20 preprocessing.Rmd
@@
 
-NCORES <- parallel::detectCores() / 2
-plink <- "../plink_linux_x86_64/plink"
+NCORES <- nb_cores()
 
 popresQC.bed <- snp_plinkQC(
-  plink.path = plink,
   prefix.in = "../POPRES_data/POPRES_allchr",
@@
 
-popresQC2.bed <- snp_plinkIBDQC(plink, popresQC.bed, ncores = NCORES)
+popresQC2.bed <- snp_plinkIBDQC(popresQC.bed, ncores = NCORES)
 ```
@@
 celiacQC.bed <- snp_plinkQC(
-  plink.path = plink,
   prefix.in = "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300",
@@
 
-celiacQC2.bed <- snp_plinkIBDQC(plink, celiacQC.bed, ncores = NCORES)
+celiacQC2.bed <- snp_plinkIBDQC(celiacQC.bed, ncores = NCORES)
 ```
 
-## Transform PLINK files in big.matrix format
+## Transform PLINK files in bigSNP format
 
 ```{r}
-popresQC.rds <- snp_readBed(popresQC2.bed, "popresQC")
+popresQC.rds <- snp_readBed(popresQC2.bed, "backingfiles/popresQC")
 
 system.time(
-  celiacQC.rds <- snp_readBed(celiacQC2.bed, "celiacQC")
+  celiacQC.rds <- snp_readBed(celiacQC2.bed, "backingfiles/celiacQC")
 )
@@
 ```{r}
-{{G@code <- bigsnpr:::CODE_IMPUTE_PRED}}
+# You need to change the code of G
+G$code256 <- bigsnpr:::CODE_IMPUTE_PRED
 big_counts(G, ind.col = 1:10)
 ```
+
+```{r}
+# To make this permanent, you need to save (modify) the file on disk
+celiac$genotypes$code256 <- bigsnpr:::CODE_IMPUTE_PRED
+saveRDS(celiac, "backingfiles/celiacQC.rds")
+```
+
+```{r, include=FALSE}
+library(gmailr)
+send_message(
+  mime(To = "florian.prive.21@gmail.com",
+       From = "florian.prive.21@gmail.com",
+       Subject = sprintf("Knitting of %s has finished",
+                         knitr::current_input()))
+)
+```

### 2017-09-11 cabios-template/draft.tex
last changes

@@
 \\
-\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} and \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
-\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
+\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
+\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 This diverse and extremely rich software environment has been of tremendous benefit for the genetic community.
-However, it has two limitations: analysis pipelines become very complex and researchers have limited access to diverse analysis tools due to growing data sizes.
+However, it has two limitations: analysis pipelines are becoming very complex and researchers have limited access to diverse analysis tools due to growing data sizes.
 
@@
 Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
-Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}.
-Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}.
-Then, implementation of GWAS analyses also depends on the genotype format, e.g.
-ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple genetic variants \cite[]{Vilhjalmsson2015,Euesden2015}.
+Similarly, quality control software for genome-wide analysis have been developed such as PLINK and the Bioconductor package GWASTools \cite[]{Gogarten2012}.
+There are also several software for the computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
+Then, implementation of GWAS analyses also depends on the data format and model analyzed.
+For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1 is limited to best guess genotypes because of its input file format.
+Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants.
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
@@
 For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations.
-Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
+Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
+We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
 Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
@@
 We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping.
-Package bigstatsr implements many statistical tools for several types of ``big.matrix'' objects (raw, char, short, integer, float and double).
+Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
 This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries.
 The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices.
-Package bigsnpr depends on bigstatsr, using a special type of ``big.matrix'' object to store the genotypes.
-Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (pruning, testing, plotting).
-
+Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''.
+Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, plotting).
 We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores.
@@
 The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory.
-They use what we could call an hybrid solution: memory-mapping.
+They use an hybrid solution: memory-mapping.
 Memory-mapping is used to access data, possibly stored on disk, as if it were in memory.
-This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory \cite[]{Kane2013}.
+This solution is made available within R through the BH package, providing access to Boost C++ Header Files\footnote{http://www.boost.org/}.
 
-We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}.
+We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008} and of the R package BEDMatrix\footnote{https://github.com/QuantGen/BEDMatrix} which provides memory-mapping directly for binary PLINK files.
 With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
 The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
-For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.
+For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.
 
-Usually, memory-mapping provides seamless and faster access than standard read/write operations.
-When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory.
+Memory-mapping provides transparent and faster access than standard read/write operations. 
+When an element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. 
 When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
 All this is managed by the Operating System so that it is seamless and efficient. 
-It means that if you use the same chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on.
-Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.
+It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
+Of course, if the memory size of the computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.
 
@@
 
-Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this format.
-In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}).
-PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs.
-We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.
+Package bigsnpr currently takes as input a variety of formats (e.g.
+vcf, bed/bim/fam, ped/map).
+However, it uses PLINK for conversion to bed/bim/fam format and for Quality Control (QC) of the data, so that we first provide R functions that use system calls to PLINK for the conversion and QC steps (Figure~\ref{fig:qc}).
+Then, fast read/write operations from/to bed/bim/fam PLINK files are implemented.
+PLINK files are then read into a ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM), a data frame with information on samples and another data frame with information on SNPs.
+We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
 
-We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
+We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
 Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
@@
 
-Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
-The first function is a wrapper to PLINK and Beagle which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}.
+We also provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
+The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
-This algorithm doesn't use phasing and is very fast, as previous Machine Learning Approaches for genetic imputation \cite[]{Wang2012}.
+This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and doesn't use phasing, thus allowing for a dramatic decrease in computation time.
 It only relies on some local XGBoost models.
 XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
-XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices.
-For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets.
+XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices.
+Systematically, for each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets.
 The training set is used to build a model for predicting missing data.
-The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.
+The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
 
@@
 For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
+While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
+Package bigstatsr use the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
+The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.
 
-The function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
-FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
-The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the ``big.matrix'' format, which enables parallel computations.
-
-Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed.
-For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}.
-There are at least 3 different options to thin SNPs based on Linkage Disequilibrium.
-The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.
-
+SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}.
+There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium, two of them named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}.
+First, pruning, the most naive approach, is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
 A variant of pruning is clumping.
@@
 
-The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
+As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
 Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}.
-However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them.
+However, the pattern of LD might be population specific, so we developed an algorithm that automatically detects these regions and removes them.
 This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
@@
 Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
-Statistical tests based on linear and logistic regressions are also available.
-For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
-Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
+We also implement statistical tests based on linear and logistic regressions.
+For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $$\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender).
+Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $$\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
 
-The R packages also implement functions to compute Polygenic Risk Scores. 
-First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
+The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
+The first method is the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
 Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
@@
 Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
-Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Dudbridge2013,Chatterjee2013,Golan2014}.
-
-Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
+A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}.
+The second approach doesn't use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
 The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
-These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Tibshirani1996,Zou2005,Friedman2010}.
+These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}.
 Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}.
 The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}.
-These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).
+These modifications allow to include covariates in the models and to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr.
 
@@
 The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-
-For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset.
-To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.
+For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset.
+To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
 \label{sec:rep}
@@
 We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort.
-We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores.
+We compare performances on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores).
 For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
@@
 We performed an association study and computed a polygenic risk score for the Celiac cohort. 
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
+The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
 For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
-For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with the XGBoost method.
-If we used a standard R matrix to store the genotypes, this data would take 32GB of memory.
-On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB only.
+For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%.
+The 0.04\% missing genotype values were imputed with the XGBoost method.
+If we used a standard R matrix to store the genotypes, this data would require 32GB of memory.
+On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.
 
-We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}).
-We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
+We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}).
+We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
 As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.
@@
 Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs.
-Finally, on the whole dataset, we made a linear association test for each SNP, using the 10 first PCs as covariables.
+Finally, we performed a linear association test in the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables.
 
@@
 We used the datasets of growing size simulated from the Celiac dataset.
-Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure~\ref{fig:bench-pca}).
+Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
 We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA.
-FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
+FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
 
@@
 The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}).
-We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
+We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
 For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
 
-For the Celiac dataset, we compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$).
-When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
-When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure~\ref{fig:scores}).
-Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}).
+For the Celiac dataset, we further compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). 
+When not removing any long range LD region, we show that PC4 and PC5 don't capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
+When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~\ref{fig:scores}).
+Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~\ref{fig:loadings}).
 This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
@@
 
-For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset.
-The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}).
-We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values.
+For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset.
+The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value.
+We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}).
 Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017).
-Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
-For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
+Overall, snp\_fastImpute made 4.7\% of imputation errors whereas Beagle made only 3.1\% of errors but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
 We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
+For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 6 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
 
@@
 Technically, bigstatsr and bigsnpr could handle any size of datasets.
-However, if accesses demand the OS to often swap between the file and the memory, this would slow down analysis.
+However, if the OS has to often swap between the file and the memory for accessing the data, this would slow down data analysis.
 For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times.
@@
 
-The two R packages don't use some specific file format nor load the entire matrix in memory but rather use a special type of matrix.
-Using a matrix-like format makes it easy to develop new functions in order to experiment and develop new ideas.
-Integration in R makes it possible to take advantage of all what R has to offer, for example using the widely-used machine learning algorithm XGBoost to easily make a fast and accurate imputation algorithm for genotyped SNPs.
+The two R packages use a matrix-like format, which makes it easy to develop new functions in order to experiment and develop new ideas.
+Integration in R makes it possible to take advantage of the vast and diverse R libraries.
+For example, we developed a fast and accurate imputation algorithm for genotyped SNPs using the widely-used machine learning algorithm XGBoost available in the R package xgboost.
 Other functions, not presented here, are also available and all the functions available within the package bigstatsr are not specific to SNP arrays, so that they could be used for other omic data or in other fields of research.
@@
 We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing PRS.
-As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files \cite[]{Palmer2016}.
+For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing Polygenic Risk Scores.
+As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS \cite[]{Palmer2016}, and it could be implemented with the data format ``FBM.code256'' without having to write multiple files.
 
@@
 \caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr.
-Some SNPs in chromosome 6 have p-values smaller than the $10^{-10}$ threshold used for vizualisation purposes.}\label{fig:gwas}
+Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
 \end{figure}
@@
 \hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times} \\
+\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
  \cline{2-3}
@@
 \hline
-Reading PLINK files & n/a & 5 / 20 sec \\
-Pruning & 4 / 6 sec & 13 / 46 sec \\
-Computing 10 PCs & 6 / 7 min & 45 / 136 sec \\
-GWAS (binary phenotype) & 5 min & 5 / 14.5 min \\
+Reading PLINK files & n/a & 5 / 20 \\
+Pruning & 4 / 4 & 14 / 52 \\
+Computing 10 PCs & 306 / 315 & 58 / 180 \\
+GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
 \hline
-Total & 11 / 12 min & 6 / 18 min \\
+Total & 649 / 612 & 378 / 1113 \\
 \hline
@@
 \caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
-The first time is with a desktop computer and the second time is with a laptop computer.}
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
 \label{tab:bench-gwas}
@@
 \hline
-GWAS (binary phenotype) & 4 / min & 3 / min \\
-Clumping & 49 / sec & 9 / sec \\
-PRS & 9 / sec & 4 / sec \\
+GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
+Clumping & 49 / 58 & 10 / 35 \\
+PRS & 9 / 10 & 2 / 3 \\
 \hline
-Total & 5 / min & 3 / min \\
+Total & 290 / 307 & 190 / 688 \\
 \hline
@@
 \end{adjustbox}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for the Celiac dataset.
-The first time is with a desktop computer and the second time is with a laptop computer.}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
 \label{tab:bench-prs}
@@
 It shows the computation time in minutes as a function of the number of samples.
-The computation corresponds to 10 Principal Components and 93,083 SNPs which remain after thinning.}\label{fig:bench-pca}
+The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
 \end{figure}
@@
 \end{tabular}
-\caption{Regions found by snp\_autoSVD for the POPRES dataset. In parentheses are regions referenced in \cite[]{Price2008}.}
+\caption{Regions found by snp\_autoSVD for the POPRES dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
 \label{tab:lrldr-popres}
-\end{table}
-
-% latex table generated in R 3.3.3 by xtable 1.8-2 package
-% Tue Jul 18 15:59:51 2017
-\begin{table}[!h]
-\centering
-\begin{tabular}{lccc}
-  \hline
- & Chromosome & Start (Mb) & Stop (Mb) \\ 
-  \hline
-1 &  2 & 134.4 (134.5) & 138.1 (138) \\ 
-  2 &  6 & 23.8 (25.5) & 35.8 (33.5) \\ 
-  3 &  8 & 6.3 (8) & 13.5 (12) \\ 
-  4 &  3 & 163.1 (n/a) & 164.9 (n/a) \\ 
-  5 & 14 & 46.6 (n/a) & 47.5 (n/a) \\ 
-   \hline
-\end{tabular}
-\caption{Regions found by snp\_autoSVD for the celiac dataset. In parentheses are regions referenced in \cite[]{Price2008}.}
-\label{tab:lrldr-celiac}
 \end{table}
@@
 \label{tab:pc-popres}
+\end{table}
+
+% latex table generated in R 3.3.3 by xtable 1.8-2 package
+% Tue Jul 18 15:59:51 2017
+\begin{table}[!h]
+\centering
+\begin{tabular}{lccc}
+  \hline
+ & Chromosome & Start (Mb) & Stop (Mb) \\ 
+  \hline
+1 &  2 & 134.4 (134.5) & 138.1 (138) \\ 
+  2 &  6 & 23.8 (25.5) & 35.8 (33.5) \\ 
+  3 &  8 & 6.3 (8) & 13.5 (12) \\ 
+  4 &  3 & 163.1 (n/a) & 164.9 (n/a) \\ 
+  5 & 14 & 46.6 (n/a) & 47.5 (n/a) \\ 
+   \hline
+\end{tabular}
+\caption{Regions found by snp\_autoSVD for the celiac dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
+\label{tab:lrldr-celiac}
 \end{table}

### 2017-09-19 cabios-template/draft.tex
update last results

@@
 To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs.
-The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer.
-The GWAS analysis was composed of three main steps.
-First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs.
-Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs.
-Finally, we performed a linear association test in the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables.
+The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer.
+The GWAS analysis was composed of four main steps.
+First we read from PLINK files in our format "bigSNP" in 1 hour.
+Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h.
+Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h.
+Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.
 
@@
 We used the datasets of growing size simulated from the Celiac dataset.
-Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
+Overall, our function big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
 We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA.
@@
 We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
-For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 6 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
+For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
 

### 2017-10-03 cabios-template/draft.tex
try again

@@
 \title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+\author{Florian Priv?\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
+\noindent$^{\text{\sf 1}}$Universit? Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int?grative (C3BI), Institut Pasteur, Paris, France \\
 \noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

### 2017-10-03 cabios-template2/bigsnpr-final-supp.tex
try again

--- 2017-09-11 cabios-template/draft.tex
@@
-%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
-\usepackage{times}
-%\usepackage{algorithm2e}
-\usepackage{url}
-\usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{amsmath, amsthm, amssymb}
-\usepackage{setspace}
-\usepackage{lineno}
-\usepackage{hyperref}
-\usepackage{bbm}
-
-\linenumbers
-\doublespacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
-
-%Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
-
-\usepackage{adjustbox}
-\usepackage{multirow}
+\documentclass[a4paper, 11pt]{article}
 \usepackage{graphicx}
@@
 
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Bold symbol macro for standard LaTeX users
-%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
-
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-
-\usepackage{babel}
-\makeatother
-
 \begin{document}
 
-\title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
-
-\date{~ }
-\maketitle
-
-\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
-\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
-
-\noindent$^\ast$To whom correspondence should be addressed.
-
-\newpage
-\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-This increase in data size is a major challenge severely slowing down genomic analyses.
-Specialized software for every part of the analysis pipeline have been developed to handle large genomic data.
-However, combining all these software into a single data analysis pipeline might be technically difficult.
-\\
-\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework.
-To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM.
-To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods.
-In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs.
-We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
-Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
-\\
-\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
-\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
-online.}
-
-\newpage
-\section{Introduction}
-
-Genome-wide datasets produced for association studies have dramatically increased in size over the past years.
-A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation.
-This diverse and extremely rich software environment has been of tremendous benefit for the genetic community.
-However, it has two limitations: analysis pipelines are becoming very complex and researchers have limited access to diverse analysis tools due to growing data sizes.
-
-Consider first the basic tools necessary to perform a standard genome-wide analysis.
-Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
-Similarly, quality control software for genome-wide analysis have been developed such as PLINK and the Bioconductor package GWASTools \cite[]{Gogarten2012}.
-There are also several software for the computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
-Then, implementation of GWAS analyses also depends on the data format and model analyzed.
-For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1 is limited to best guess genotypes because of its input file format.
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants.
-As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
-Overall, this might be a brake on data exploration.
-%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.
-
-Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
-This is particularly a burden for commonly used analysis languages such as R, Python and Perl.
-Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented.
-Hopefully, strategies have been developed to avoid loading large datasets in RAM.
-For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations.
-Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
-We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
-Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
-
-\section{Approach}
-
-We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping.
-Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
-This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries.
-The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices.
-Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''.
-Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, plotting).
-We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores.
-We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.
-
-\section{Methods}
-
-\subsection{Memory-mapped files}
-
-The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory.
-They use an hybrid solution: memory-mapping.
-Memory-mapping is used to access data, possibly stored on disk, as if it were in memory.
-This solution is made available within R through the BH package, providing access to Boost C++ Header Files\footnote{http://www.boost.org/}.
-
-We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008} and of the R package BEDMatrix\footnote{https://github.com/QuantGen/BEDMatrix} which provides memory-mapping directly for binary PLINK files.
-With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
-The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
-For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.
-
-Memory-mapping provides transparent and faster access than standard read/write operations. 
-When an element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. 
-When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
-All this is managed by the Operating System so that it is seamless and efficient. 
-It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
-Of course, if the memory size of the computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.
-
-\subsection{Data management, preprocessing and imputation} \label{sec:preprocess}
-
-Package bigsnpr currently takes as input a variety of formats (e.g.
-vcf, bed/bim/fam, ped/map).
-However, it uses PLINK for conversion to bed/bim/fam format and for Quality Control (QC) of the data, so that we first provide R functions that use system calls to PLINK for the conversion and QC steps (Figure~\ref{fig:qc}).
-Then, fast read/write operations from/to bed/bim/fam PLINK files are implemented.
-PLINK files are then read into a ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM), a data frame with information on samples and another data frame with information on SNPs.
-We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
-
-We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
-Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
-With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
-
-We also provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
-The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R.
-The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
-This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and doesn't use phasing, thus allowing for a dramatic decrease in computation time.
-It only relies on some local XGBoost models.
-XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
-XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices.
-Systematically, for each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets.
-The training set is used to build a model for predicting missing data.
-The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
-
-\subsection{Population structure and SNP thinning based on Linkage Disequilibrium}
-
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
-While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
-Package bigstatsr use the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
-The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.
-
-SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}.
-There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium, two of them named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}.
-First, pruning, the most naive approach, is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
-A variant of pruning is clumping.
-Clumping is useful if a statistic is available to sort the SNPs by importance, e.g.
-association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest.
-Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.
-
-As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
-Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}.
-However, the pattern of LD might be population specific, so we developed an algorithm that automatically detects these regions and removes them.
-This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
-Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions.
-Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA.
-This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.
-
-\subsection{Association tests and Polygenic Risk Scores}
-
-Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
-Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
-We also implement statistical tests based on linear and logistic regressions.
-For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $$\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender).
-Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $$\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
-
-The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
-The first method is the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
-Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
-The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other.
-Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
-A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}.
-The second approach doesn't use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
-The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
-These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}.
-Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}.
-The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}.
-These modifications allow to include covariates in the models and to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr.
-
-\subsection{Data analyzed}
-
-In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
-The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
-The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset.
-To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
-\label{sec:rep}
-
-\subsection{Reproducibility}
-
-All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data.
-
-\section{Results}
-
-\subsection{Overview}\label{sec:overview}
-
-We present the results for three different analyses.
-First, we illustrate the application of R packages bigstatsr and bigsnpr.
-Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT).
-Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes.
-We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort.
-We compare performances on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores).
-For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
-
-\subsection{Application}
-
-We performed an association study and computed a polygenic risk score for the Celiac cohort. 
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
-For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
-For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%.
-The 0.04\% missing genotype values were imputed with the XGBoost method.
-If we used a standard R matrix to store the genotypes, this data would require 32GB of memory.
-On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.
-
-We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}).
-We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
-As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.
-
-%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
-To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs.
-The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer.
-The GWAS analysis was composed of three main steps.
-First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs.
-Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs.
-Finally, we performed a linear association test in the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables.
-
-\subsection{Method Comparison}
-
-We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
-For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.
-pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
-Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}).
-However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.
-
-On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr.
-For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs.
-We used the datasets of growing size simulated from the Celiac dataset.
-Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
-We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA.
-FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
-
-\subsection{Automatic detection of long-range LD regions}
-
-For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets.
-For the POPRES dataset, the algorithm converged in two iterations.
-The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}).
-We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
-For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
-
-For the Celiac dataset, we further compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). 
-When not removing any long range LD region, we show that PC4 and PC5 don't capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
-When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~\ref{fig:scores}).
-Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~\ref{fig:loadings}).
-This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
-
-\subsection{Imputation of missing values for genotyped SNPs}\label{sec:impute}
-
-For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset.
-The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value.
-We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}).
-Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017).
-Overall, snp\_fastImpute made 4.7\% of imputation errors whereas Beagle made only 3.1\% of errors but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
-We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
-For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 6 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
-
-\section{Discussion}
-
-We have developed two R packages, bigstatsr and bigsnpr, which enable multiple analyses of large-scale genotype datasets in a single comprehensive framework.
-Linkage Disequilibrium pruning, Principal Component Analysis, association tests and computations of polygenic risk scores are made available in this software.
-Implemented algorithms are both fast and memory-efficient, allowing the use of laptops or desktop computers to make genome-wide analyses.
-Technically, bigstatsr and bigsnpr could handle any size of datasets.
-However, if the OS has to often swap between the file and the memory for accessing the data, this would slow down data analysis.
-For example, the Principal Component Analysis (PCA) algorithm in bigstatsr is iterative so that the matrix has to be sequentially accessed over a hundred times.
-If the number of samples times the number of SNPs remaining after pruning is larger than the available memory, this slowdown would happen.
-For instance, a 32GB computer would be slow when computing PCs on more than 100K samples and 300K SNPs remaining after LD thinning.
-
-The two R packages use a matrix-like format, which makes it easy to develop new functions in order to experiment and develop new ideas.
-Integration in R makes it possible to take advantage of the vast and diverse R libraries.
-For example, we developed a fast and accurate imputation algorithm for genotyped SNPs using the widely-used machine learning algorithm XGBoost available in the R package xgboost.
-Other functions, not presented here, are also available and all the functions available within the package bigstatsr are not specific to SNP arrays, so that they could be used for other omic data or in other fields of research.
-
-We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing Polygenic Risk Scores.
-As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS \cite[]{Palmer2016}, and it could be implemented with the data format ``FBM.code256'' without having to write multiple files.
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{conversion+QC.pdf}}
-\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{imputation.pdf}}
-\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{svd.pdf}}
-\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{celiac-pca}}
-\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
-\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr.
-Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
-\end{figure}
-
-\newpage
-\begin{table}[!h]
-\centering
-\begin{adjustbox}{max width=\textwidth}
-\begin{tabular}{|c|c|c|}
-\hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
- \cline{2-3}
- & PLINK and FastPCA & bigstatsr and bigsnpr \\
-\hline
-Reading PLINK files & n/a & 5 / 20 \\
-Pruning & 4 / 4 & 14 / 52 \\
-Computing 10 PCs & 306 / 315 & 58 / 180 \\
-GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
-\hline
-Total & 649 / 612 & 378 / 1113 \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
-\label{tab:bench-gwas}
-\end{table}
-
-\newpage
-\begin{table}[!h]
-\centering
-\begin{adjustbox}{max width=\textwidth}
-\begin{tabular}{|c|c|c|}
-\hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
- \cline{2-3}
- & PLINK & bigstatsr and bigsnpr \\
-\hline
-GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
-Clumping & 49 / 58 & 10 / 35 \\
-PRS & 9 / 10 & 2 / 3 \\
-\hline
-Total & 290 / 307 & 190 / 688 \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
-\label{tab:bench-prs}
-\end{table}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{benchmark-pca}}
-\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the computation time in minutes as a function of the number of samples.
-The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{precision-pca}}
-\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
-\end{figure}
-
-\newpage
-\section*{Acknowledgements}
-
-Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
-
-\vspace*{-12pt}
-
-\bibliographystyle{natbib}
-\bibliography{document}
-
-\newpage
-\section*{Supplementary Data}
+\section{Supplementary Data}
 
@@
 
-\begin{table}[!h]
+\begin{table}[htpb]
 \centering
@@
 % Tue Jun 13 11:29:13 2017
-\begin{table}[!h]
+\begin{table}[htpb]
 \centering
@@
 % Tue Jul 18 15:59:51 2017
-\begin{table}[!h]
+\begin{table}[htpb]
 \centering
@@
 % Tue Jun 13 11:25:56 2017
-\begin{table}[!h]
+\begin{table}[htpb]
 \centering
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{scores}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{loadings}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{gini}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{hist-maf}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{hist-NA}}
@@
 
-\begin{figure}[!h]
+\begin{figure}[htpb]
 \centerline{\includegraphics[width=\textwidth]{error-impute}}
@@
 
+\vspace*{-12pt}
+
+\bibliographystyle{natbib}
+\bibliography{document}
+
 \end{document}

### 2017-10-03 cabios-template2/bigsnpr-final.tex
try again

--- 2017-10-03 cabios-template/draft.tex
@@
-%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
-\usepackage{times}
-%\usepackage{algorithm2e}
-\usepackage{url}
-\usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{amsmath, amsthm, amssymb}
-\usepackage{setspace}
-\usepackage{lineno}
-\usepackage{hyperref}
-\usepackage{bbm}
+\documentclass{bioinfo}
+\copyrightyear{2015} \pubyear{2015}
 
-\linenumbers
-\doublespacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
-
-%Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Manuscript Category}
 
@@
 \usepackage{multirow}
-\usepackage{graphicx}
-\graphicspath{{../figures/}}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Bold symbol macro for standard LaTeX users
-%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
-
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-
-\usepackage{babel}
-\makeatother
+%\textsc{\usepackage{graphicx}
+%\graphicspath{{../figures/}}}
+\usepackage{xr}
+\externaldocument{bigsnpr-final-supp}
 
 \begin{document}
+\firstpage{1}
 
-\title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author{Florian Priv?\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+\subtitle{Subject Section}
 
-\date{~ }
-\maketitle
+\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
+\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
+\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
+$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}
 
-\noindent$^{\text{\sf 1}}$Universit? Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int?grative (C3BI), Institut Pasteur, Paris, France \\
-\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
+\corresp{$^\ast$To whom correspondence should be addressed.}
 
-\noindent$^\ast$To whom correspondence should be addressed.
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-\newpage
+\editor{Associate Editor: XXXXXXX}
+
 \abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
@@
 
-\newpage
+\maketitle
+
 \section{Introduction}
@@
 
+%\enlargethispage{12pt}
+
 \section{Approach}
@@
 
+\begin{methods}
 \section{Methods}
@@
 
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=160pt]{conversion-QC.pdf}}
+\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
+\end{figure}
+
 We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
@@
 
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=235pt]{imputation.pdf}}
+\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
+\end{figure}
+
 \subsection{Population structure and SNP thinning based on Linkage Disequilibrium}
@@
 The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=235pt]{svd.pdf}}
+\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
+\end{figure}
 
@@
 Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
-We also implement statistical tests based on linear and logistic regressions.
-For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $$\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender).
-Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $$\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
+We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
+\begin{multline}
+  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
+\end{multline}
+and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender).
+Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
+\begin{multline}
+  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
+\end{multline}
+and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
 
@@
 
+\end{methods}
+
 \section{Results}
@@
 
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=200pt]{celiac-pca}}
+\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
+\end{figure}
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
+\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr.
+Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
+\end{figure}
+
 %The Celiac dataset is relatively small as compared to modern genetic cohorts. 
@@
 
+\begin{table}[!tpb]
+\centering
+\begin{adjustbox}{max width=235pt}
+\begin{tabular}{|c|c|c|}
+\hline
+\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
+ \cline{2-3}
+ & PLINK and FastPCA & bigstatsr and bigsnpr \\
+\hline
+Reading PLINK files & n/a & 5 / 20 \\
+Pruning & 4 / 4 & 14 / 52 \\
+Computing 10 PCs & 306 / 315 & 58 / 180 \\
+GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
+\hline
+Total & 649 / 612 & 378 / 1113 \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
+\label{tab:bench-gwas}
+\end{table}
+
+\begin{table}[!tpb]
+\centering
+\begin{adjustbox}{max width=235pt}
+\begin{tabular}{|c|c|c|}
+\hline
+\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
+ \cline{2-3}
+ & PLINK & bigstatsr and bigsnpr \\
+\hline
+GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
+Clumping & 49 / 58 & 10 / 35 \\
+PRS & 9 / 10 & 2 / 3 \\
+\hline
+Total & 290 / 307 & 190 / 688 \\
+\hline
+\end{tabular} 
+\end{adjustbox}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
+\label{tab:bench-prs}
+\end{table}
+
 On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr.
@@
 FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=200pt]{benchmark-pca.png}}
+\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
+It shows the computation time in minutes as a function of the number of samples.
+The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
+\end{figure}
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=200pt]{precision-pca.png}}
+\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
+It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
+\end{figure}
 
@@
 
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{conversion+QC.pdf}}
-\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{imputation.pdf}}
-\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{svd.pdf}}
-\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{celiac-pca}}
-\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
-\end{figure}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
-\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr.
-Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
-\end{figure}
-
-\newpage
-\begin{table}[!h]
-\centering
-\begin{adjustbox}{max width=\textwidth}
-\begin{tabular}{|c|c|c|}
-\hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
- \cline{2-3}
- & PLINK and FastPCA & bigstatsr and bigsnpr \\
-\hline
-Reading PLINK files & n/a & 5 / 20 \\
-Pruning & 4 / 4 & 14 / 52 \\
-Computing 10 PCs & 306 / 315 & 58 / 180 \\
-GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
-\hline
-Total & 649 / 612 & 378 / 1113 \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
-\label{tab:bench-gwas}
-\end{table}
-
-\newpage
-\begin{table}[!h]
-\centering
-\begin{adjustbox}{max width=\textwidth}
-\begin{tabular}{|c|c|c|}
-\hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
- \cline{2-3}
- & PLINK & bigstatsr and bigsnpr \\
-\hline
-GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
-Clumping & 49 / 58 & 10 / 35 \\
-PRS & 9 / 10 & 2 / 3 \\
-\hline
-Total & 290 / 307 & 190 / 688 \\
-\hline
-\end{tabular} 
-\end{adjustbox}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
-\label{tab:bench-prs}
-\end{table}
-
-\newpage
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{benchmark-pca}}
-\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the computation time in minutes as a function of the number of samples.
-The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{precision-pca}}
-\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
-\end{figure}
-
-\newpage
 \section*{Acknowledgements}
 
-Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
 
@@
 
-\newpage
-\section*{Supplementary Data}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-\subsection{Long-range LD regions}
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{lccc}
-  \hline
- & Chromosome & Start (Mb) & Stop (Mb) \\ 
-  \hline
-1 &  2 & 134.7 (134.5) & 137.3 (138) \\ 
-  2 &  6 & 27.5 (25.5) & 33.1 (33.5) \\ 
-  3 &  8 & 6.6 (8) & 13.2 (12) \\ 
-   \hline
-\end{tabular}
-\caption{Regions found by snp\_autoSVD for the POPRES dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
-\label{tab:lrldr-popres}
-\end{table}
-
-% latex table generated in R 3.3.3 by xtable 1.8-2 package
-% Tue Jun 13 11:29:13 2017
-\begin{table}[!h]
-\centering
-\begin{tabular}{l|cccccccccc}
-  \hline
- & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
-  \hline
-PC1 & 100.0 & -0.1 & -0.0 & 0.1 & -0.1 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\
-  PC2 & 0.1 & 100.0 & -0.0 & 0.1 & -0.0 & -0.0 & -0.0 & 0.2 & -0.1 & -0.0 \\
-  PC3 & 0.0 & -0.0 & 99.9 & 0.9 & 0.1 & -0.1 & -0.3 & 0.2 & 0.4 & 0.1 \\
-  PC4 & -0.1 & -0.1 & -0.9 & 99.7 & -1.0 & 0.7 & 0.6 & 0.2 & 0.3 & 0.9 \\
-  PC5 & 0.1 & 0.0 & -0.1 & 1.1 & 99.3 & 1.3 & -0.8 & 1.3 & -4.2 & -2.4 \\
-  PC6 & -0.0 & 0.0 & 0.1 & -0.7 & -1.0 & 97.7 & -3.5 & 6.1 & 7.9 & -6.2 \\
-  PC7 & -0.0 & -0.1 & 0.2 & -0.3 & -1.7 & 0.3 & 58.3 & 73.2 & -25.9 & 9.1 \\
-  PC8 & 0.1 & -0.1 & -0.3 & 0.4 & -0.5 & -5.3 & -73.5 & 59.5 & 15.8 & 13.2 \\
-  PC9 & 0.0 & 0.1 & -0.4 & -0.8 & 5.0 & -7.6 & 27.8 & 11.0 & 91.9 & 9.0 \\
-  PC10 & 0.1 & 0.0 & 0.0 & -0.9 & 1.6 & 10.2 & 3.9 & -19.6 & -6.3 & 89.2 \\
-   \hline
-\end{tabular}
-\caption{Correlation between scores of PCA for the POPRES dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
-\label{tab:pc-popres}
-\end{table}
-
-% latex table generated in R 3.3.3 by xtable 1.8-2 package
-% Tue Jul 18 15:59:51 2017
-\begin{table}[!h]
-\centering
-\begin{tabular}{lccc}
-  \hline
- & Chromosome & Start (Mb) & Stop (Mb) \\ 
-  \hline
-1 &  2 & 134.4 (134.5) & 138.1 (138) \\ 
-  2 &  6 & 23.8 (25.5) & 35.8 (33.5) \\ 
-  3 &  8 & 6.3 (8) & 13.5 (12) \\ 
-  4 &  3 & 163.1 (n/a) & 164.9 (n/a) \\ 
-  5 & 14 & 46.6 (n/a) & 47.5 (n/a) \\ 
-   \hline
-\end{tabular}
-\caption{Regions found by snp\_autoSVD for the celiac dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
-\label{tab:lrldr-celiac}
-\end{table}
-
-% latex table generated in R 3.3.3 by xtable 1.8-2 package
-% Tue Jun 13 11:25:56 2017
-\begin{table}[!h]
-\centering
-\begin{tabular}{l|cccccccccc}
-  \hline
- & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
-  \hline
-PC1 & 100.0 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
-  PC2 & 0.1 & 100.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 \\
-  PC3 & 0.1 & -0.0 & 99.9 & 0.2 & -0.0 & 0.1 & 0.1 & 0.1 & 0.0 & -0.1 \\
-  PC4 & -0.0 & -0.0 & -0.3 & 99.9 & -0.1 & 0.1 & -0.1 & 0.0 & 0.1 & 0.1 \\
-  PC5 & 0.0 & 0.0 & 0.0 & 0.1 & 99.7 & 0.9 & -0.3 & 0.1 & -0.8 & -0.6 \\
-  PC6 & -0.0 & 0.0 & -0.1 & -0.2 & -0.8 & 99.6 & 0.5 & -0.5 & -0.2 & -0.4 \\
-  PC7 & -0.0 & 0.0 & -0.1 & 0.0 & 0.5 & -0.4 & 98.9 & 3.1 & 0.7 & 1.6 \\
-  PC8 & 0.0 & 0.0 & -0.2 & -0.0 & -0.2 & 0.5 & -3.2 & 98.4 & -4.5 & -1.5 \\
-  PC9 & -0.0 & -0.0 & -0.0 & 0.0 & 0.6 & 0.1 & -0.7 & 4.6 & 96.9 & -10.7 \\
-  PC10 & -0.0 & -0.0 & 0.1 & -0.1 & 0.3 & 0.1 & -1.2 & 1.5 & 8.6 & 92.7 \\
-   \hline
-\end{tabular}
-\caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
-\label{tab:pc-celiac}
-\end{table}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{scores}}
-\caption{PC4 and PC5 of the celiac disease dataset.
-Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$).
-Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4.
-Right panel, PC scores obtained with the automatic detection and removal of long-range LD regions.
-Individuals are coloured according to their population of origin.}\label{fig:scores}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{loadings}}
-\caption{Loadings of first 6 PCs of the celiac disease dataset plotted as hexbins (2-D histogram with hexagonal cells).
-On the left, without removing any long range LD region (only clumping at $R^2 > 0.2$).
-On the right, with the automatic detection and removal of long-range LD regions.}\label{fig:loadings}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{gini}}
-\caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $R^2 > 0.2$) and with the automatic detection and removal of long-range LD regions.
-The dashed line corresponds to the theoretical value for gaussian loadings.}\label{fig:gini}
-\end{figure}
-
-\newpage
-\subsection{Imputation}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{hist-maf}}
-\caption{Histogram of the minor allele frequencies of the POPRES dataset used for comparing imputation methods.}\label{fig:maf}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{hist-NA}}
-\caption{Histogram of the number of missing values by SNP. These numbers were generated using a Beta-binomial distribution.}\label{fig:NA}
-\end{figure}
-
-\begin{figure}[!h]
-\centerline{\includegraphics[width=\textwidth]{error-impute}}
-\caption{Number of imputation errors vs the estimated number of imputation errors by SNP.
-For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
-The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
-\end{figure}
-
 \end{document}

### 2018-01-09 cabios-template2/bigsnpr-review.tex
copy text

--- 2017-10-03 cabios-template2/bigsnpr-final.tex
@@
 \usepackage{multirow}
-%\textsc{\usepackage{graphicx}
-%\graphicspath{{../figures/}}}
+\usepackage{graphicx}
+\graphicspath{{../figures/}}
 \usepackage{xr}
@@
 \title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
-\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
+\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
+\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France and \\
 $^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

### 2018-01-09 cabios-template2/bigsnpr-review.tex
intro

@@
 Then, implementation of GWAS analyses also depends on the data format and model analyzed.
-For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1 is limited to best guess genotypes because of its input file format.
+For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1.9 is limited to best guess genotypes because of its input file format.
+Note that PLINK 2.0 will completely handle this type of data, but is still in the alpha stage of development.
 Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants.
 As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
-Overall, this might be a brake on data exploration.
+Overall, this might be a barrier to data exploration.
 %Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.
@@
 Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
-We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
-Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
 
@@
 
-We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping.
+{\bf In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses.
+For instance, one might prefer using R because it makes easier to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
+We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
+Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
+This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.}
 Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
@@
 Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''.
-Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, plotting).
+Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting).
 We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores.

### 2018-01-09 cabios-template2/bigsnpr-review.tex
save

@@
 \usepackage{xr}
-\externaldocument{bigsnpr-final-supp}
+\externaldocument{bigsnpr-review-supp}
 
@@
 
-{\bf In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses.
+{\color{red}
+In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
 For instance, one might prefer using R because it makes easier to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
@@
 Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.
-This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.}
+This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
+}
 Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
@@
 
-The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory.
+The two R packages do not use standard read operations on a file nor load the genotype matrix entirely in memory.
 They use an hybrid solution: memory-mapping.
@@
 
-Package bigsnpr currently takes as input a variety of formats (e.g.
-vcf, bed/bim/fam, ped/map).
-However, it uses PLINK for conversion to bed/bim/fam format and for Quality Control (QC) of the data, so that we first provide R functions that use system calls to PLINK for the conversion and QC steps (Figure~\ref{fig:qc}).
-Then, fast read/write operations from/to bed/bim/fam PLINK files are implemented.
-PLINK files are then read into a ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM), a data frame with information on samples and another data frame with information on SNPs.
+{\color{red}
+We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
+Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls.
+With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
+So, we use a single data format that can store both genotype calls and dosages.
+
+For preprocessing steps, we think the user should use PLINK, which is an invaluable tool for this.
+For the sake of reproducibility, one could use PLINK directly from R via systems calls.
+For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.
+vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
+Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs.
 We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
+Finally, we provide two methods for reading-in dosages as proof of concept and will provide a more final solution when we get access to the UK Biobank dataset ([SUPPMAT]).
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=160pt]{conversion-QC.pdf}}
-\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
-\end{figure}
-
-We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
-Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''.
-With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
-
-We also provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}).
-The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R.
+Most modern SNP chips provide genotype data with large call-rates.
+For example, the Celiac data we use in this paper present only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control.
+Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
+So, we provide two functions for imputing missing values of genotyped SNPs.
+Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.
+imputation servers for human data \cite[]{mccarthy2016reference}.
+The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}).
+Beagle uses phasing before imputing, which could take several weeks even when there are only a tiny amount of missing values to be imputed.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
-This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and doesn't use phasing, thus allowing for a dramatic decrease in computation time.
+We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
+This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time.
 It only relies on some local XGBoost models.
@@
 XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
-Systematically, for each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets.
-The training set is used to build a model for predicting missing data.
-The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=235pt]{imputation.pdf}}
-\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
-\end{figure}
+Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP.
+Those latter individuals are further separated in a training set and a validation set (e.g.
+80\% training and 20\% validation).
+The training set is used to build the XGBoost model for predicting missing data.
+The prediction model is then evaluated on the validation set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
+The prediction model is also projected on the test set (missing values) in order to impute them.
+}
 
@@
 A variant of pruning is clumping. 
-Clumping is useful if a statistic is available to sort the SNPs by importance, e.g.
-association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest.
-Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.
+{\color{red}
+Clumping is useful if a statistic is available to sort the SNPs by importance.
+Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome.
+For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance.
+Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the higher MAF is kept.
+Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all so we advise to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping ([SUPMAT]).
+}
 
@@
 The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
-The first method is the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations.
-Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
-The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other.
+The first method is the widely-used Clumping + Thresholding (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.
+Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
+The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other.
 Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
 A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}.
-The second approach doesn't use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
+The second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
 The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
@@
 The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset.
+For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlation} as the original dataset.
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
@@
 Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT).
-Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes.
+Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes {\color{red} (for genotyped SNPs only)}.
 We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort.
-We compare performances on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores).
+We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores).
 For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
@@
 \begin{table}[!tpb]
-\centering
+\begin{center}
 \begin{adjustbox}{max width=235pt}
@@
 \hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
+\multirow{2}{*}{Operation} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
  \cline{2-3}
@@
 \end{adjustbox}
+\end{center}
 \caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
@@
 \end{adjustbox}
+\end{center}
 \caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
@@
 For the Celiac dataset, we further compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). 
-When not removing any long range LD region, we show that PC4 and PC5 don't capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
+When not removing any long range LD region, we show that PC4 and PC5 do not capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
 When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~\ref{fig:scores}).
@@
 We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
-For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle didn't finish imputing chromosome 1 in 48 hours.
+For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle did not finish imputing chromosome 1 in 48 hours.
 
@@
 We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing Polygenic Risk Scores.
+For example, we wish to use these packages to train much more accurate predictive models than the standard C+T model currently in use when computing Polygenic Risk Scores.
 As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS \cite[]{Palmer2016}, and it could be implemented with the data format ``FBM.code256'' without having to write multiple files.

### 2018-01-11 benchmark-pca.Rmd
save

@@
   library(bigsnpr)   
-  # devtools::install_github("gabraham/flashpca", subdir = "flashpcaR")
+  # devtools::install_github("gabraham/flashpca/flashpcaR")
   library(flashpcaR)
@@
 smartpca <- "../EIG-6.1.4/bin/smartpca"
+
+# PLINK 2.0: https://www.cog-genomics.org/plink/2.0/
+plink2 <- "./plink2"
 
@@
 
+plink2pca <- function(prefix, fast = TRUE, parallel = TRUE) {
+
+  time <- system.time(
+    system(glue("{plink2} --bfile {prefix} --out {prefix}",
+                " --pca {`if`(fast, 'approx 10', '10')}",
+                "{`if`(!parallel, ' --threads 1', '')}"))
+  )
+
+  evec <- data.table::fread(glue("{prefix}.eigenvec"), 
+                            select = paste0("PC", 1:10))
+
+  list(time[3], as.matrix(evec))
+}
+
 # bigstatsr::big_randomSVD
@@
   "bigstatsr - 1 core",   mypca,     list(),
-  "bigstatsr - 6 cores", mypca,    list(ncores = NCORES)
+  "bigstatsr - 6 cores",  mypca,     list(ncores = NCORES),
+  "PLINK 2.0 - 1 core",   plink2pca, list(parallel = FALSE),
+  "PLINK 2.0 - 11 cores", plink2pca, list()
 ) %>%
@@
 ```{r}
-# SmartPCA
-smartpca <- fastpca("backingfiles/celiacQC_keep1", fast = FALSE)
-true <- smartpca[[2]]
-cat(sprintf("Standard PCA takes forever: %.2gh.", smartpca[[1]] / 3600))
+# Exact PCA
+true <- plink2pca("backingfiles/celiacQC_keep1", fast = FALSE)[[2]]
 ```
@@
 (preci <- models %>%
+    mutate(method = forcats::fct_recode(
+      method, 
+      `bigstatsr` = "bigstatsr - 1 core",
+      `PLINK 2.0` = "PLINK 2.0 - 1 core", 
+      NULL = "bigstatsr - 6 cores", 
+      NULL = "PLINK 2.0 - 11 cores"
+    )) %>%
+    filter(!is.na(method)) %>%
     mutate(
@@
        color = "Number of individuals", shape = "Number of individuals", linetype = "Number of individuals") + 
-  theme(legend.position = c(0.8, 0.15),
+  theme(legend.position = c(0.25, 0.15),
         strip.text.x = element_text(size = rel(3)),
@@
 
-```{r, include=FALSE}
+```{r, include=FALSE, error=TRUE}
 library(gmailr)

### 2018-01-11 cabios-template2/bigsnpr-review.tex
save

@@
 In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
-For instance, one might prefer using R because it makes easier to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
+For instance, one might like using R because R makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
 We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.

### 2018-01-11 test-prsice.R

@@ -1,0 +1,87 @@
+
+# download.file("http://portals.broadinstitute.org/collaboration/giant/images/0/01/GIANT_HEIGHT_Wood_et_al_2014_publicrelease_HapMapCeuFreq.txt.gz",
+#               destfile = "sumstats-height.txt.gz")
+# R.utils::gunzip("sumstats-height.txt.gz")
+
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+
+sumstats <- data.table::fread("sumstats-height.txt", data.table = FALSE)
+
+indSNP <- match(celiac$map$marker.ID, sumstats$MarkerName)
+print(nrow(sumstats) - sum(!is.na(indSNP)))
+
+same_ref
+
+same <- same_ref(celiac$map$allele1, celiac$map$allele2, 
+                 sumstats$Allele1[indSNP], sumstats$Allele2[indSNP])
+sum(same, na.rm = TRUE)
+
+
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+NCORES <- nb_cores()
+betas <- sumstats$b[indSNP]
+pval <- sumstats$p[indSNP]
+lpS <- -log10(pval)
+
+
+path <- "PRSice_linux.bugfix.20171116/"
+prefix <- "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
+
+system.time(
+  system(glue::glue("Rscript {path}PRSice.R",
+                    " --prsice {path}PRSice_linux",
+                    " --base sumstats-height.txt",
+                    " --target {prefix}",
+                    " --thread 6",
+                    # " --full",
+                    # " --fastscore",
+                    " --pvalue p",
+                    " --stat b",
+                    " --beta T",
+                    " --snp MarkerName",
+                    " --A1 Allele1 --A2 Allele2",
+                    " --binary-target T"))
+)
+
+
+cor.pval0 <- data.table::fread("PRSice.prsice", select = "P")[["P"]] 
+
+library(dplyr)
+thrs <- -log10(data.table::fread("PRSice.prsice", select = "Threshold")[["Threshold"]])
+
+system.time(
+  ind.keep <- snp_clumping(G, infos.chr = CHR, S = lpS, thr.r2 = 0.1,
[…]

### 2018-01-14 public-data.Rmd
save

---
title: "Public data: domestic dogs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7, error = TRUE)
options(width = 85)
```

```{r, eval=FALSE}
bed <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104774/cornell_canine.bed"
bim <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104773/cornell_canine.bim"
fam <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104772/cornell_canine.fam"
lapply(c(bed, bim, fam), function(url) {
  download.file(url, destfile = paste0("backingfiles/", basename(url)))
})
```

```{r}
plink <- download_plink()
snp_plinkQC("backingfiles/cornell_canine", 
            plink.path = plink,
            maf = 0.05,
            geno = 0,
            hwe = 1e-10,
            autosome.only = TRUE,
            extra.options = "--dog")
```

```{r}
system.time(
  rds <- snp_readBed("backingfiles/cornell_canine_QC.bed", "backingfiles/dogs")
)
system.time(
  bigsnp <- snp_attach(rds)
)
```

```{r}
G <- bigsnp$genotypes
sum(big_counts(G)[4, ])
```

```{r}

```

### 2018-01-14 test-prsice.R

@@ -30,15 +30,19 @@ lpS <- -log10(pval)
 
 path <- "PRSice_linux.bugfix.20171116/"
 prefix <- "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
+tmp <- tempfile()
 
 system.time(
   system(glue::glue("Rscript {path}PRSice.R",
                     " --prsice {path}PRSice_linux",
                     " --base sumstats-height.txt",
                     " --target {prefix}",
+                    " --out {tmp}",
                     " --thread 6",
                     # " --full",
                     # " --fastscore",
+                    # " --cov-file covar.txt",
+                    # " --quantile 20",
                     " --pvalue p",
                     " --stat b",
                     " --beta T",
@@ -55,6 +59,7 @@ thrs <- -log10(data.table::fread("PRSice.prsice", select = "Threshold")[["Thresh
 
 system.time(
   ind.keep <- snp_clumping(G, infos.chr = CHR, S = lpS, thr.r2 = 0.1,
+                           ind.row = which(y01 == 0),
                            size = 250, is.size.in.bp = TRUE, infos.pos = POS,
                            exclude = which(is.na(same) | (lpS < min(thrs))), 
                            ncores = NCORES)
@@ -76,12 +81,25 @@ y01 <- celiac$fam$affection - 1
 # })
 cor.pval2 <- prs %>%
   big_copy() %>%
-  big_univLogReg(y01) %>%
+  big_univLogReg(y01, covar.train = obj.svd$u) %>%
   predict(log10 = FALSE)
 # plot(cor.pval, cor.pval2)
 # all.equal(cor.pval, cor.pval2)
 all.equal(cor.pval2, cor.pval0)
 plot(cor.pval2, cor.pval0, log = "xy"); abline(0, 1, col = "red")
-
+cor(log10(cor.pval2), log10(cor.pval0))
 
 CHR[lpS > 10]
+plot(10^(-thrs), -log10(cor.pval2))
+
+
+# obj.svd <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
+# write.table(cbind(celiac$fam[, 1:2], obj.svd$u), "covar.txt", 
+#             quote = FALSE, row.names = FALSE)
+# cor.pval2 <- prs %>%
+#   big_copy() %>%
+#   big_univLogReg(y01, covar.train = obj.svd$u) %>%
[…]

### 2018-01-14 from-hapgen2.R
save

@@ -1,0 +1,21 @@
+library(bigsnpr)
+
+rds <- snp_readBed("simus_hapgen2/simus_1_10.bed",
+                   tmp <- tempfile())
+
+bigsnp <- snp_attach(rds)
+G <- bigsnp$genotypes
+dim(G)
+CHR <- bigsnp$map$chromosome
+rle(CHR)
+POS <- bigsnp$map$physical.pos
+
+counts <- big_counts(G)
+counts[, 1:10]
+sum(counts[4, ])
+which(counts[4, ] != 0)
+indNA <- which(counts[4, ] != 0)
+
+NCORES <- nb_cores()
+ind.rm <- snp_indLRLDR(CHR, POS)
+ind.keep <- snp_clumping(G, CHR, exclude = union(indNA, ind.rm), ncores = NCORES)

### 2018-01-14 hapgen2.R

@@ -1,0 +1,98 @@
+library(glue)
+library(foreach)
+
+# http://mathgen.stats.ox.ac.uk/genetics_software/hapgen/hapgen2.html
+hapgen2 <- "hapgen2_x86_64/hapgen2"
+# https://mathgen.stats.ox.ac.uk/wtccc-software/HM3.tgz
+path <- file.path(dirname(hapgen2), "CEU.0908.impute.files")
+# http://www.well.ox.ac.uk/~cfreeman/software/gwas/gtool.html
+gtool <- "gtool_v0.7.5_x86_64/gtool"
+# https://www.cog-genomics.org/plink2
+plink <- bigsnpr::download_plink()
+simus.path <- "simus_hapgen2"
+dir.create(simus.path)
+
+get_simu <- function(i, n = 1000) {
+  system <- function(command) base::system(command, ignore.stdout = TRUE)
+  # prefix of created files
+  prefix <- glue("{simus.path}/simu{i}.out")
+  # simulate with hapgen2 in Oxford format
+  cmd.hapgen2 <- glue(
+    "{hapgen2}",
+    " -m {path}/genetic_map_chr2_combined_b36.txt",
+    " -l {path}/CEU.0908.chr2.legend",
+    " -h {path}/CEU.0908.chr2.hap",
+    " -o {prefix}.gz",
+    " -dl 205 1 1 1",
+    " -n {n} 1", 
+    " -no_haps_output"
+  )
+  system(cmd.hapgen2)
+  # convert to ped.gz/map.gz
+  cmd.gtool <- glue(
+    "{gtool} -G",
+    " --g {prefix}.controls.gen.gz", 
+    " --s {prefix}.controls.sample"
+  )
+  system(cmd.gtool)
+  # uncompress
+  system(glue("gzip -d {prefix}.controls.gen.map.gz"))
+  system(glue("gzip -d {prefix}.controls.gen.ped.gz"))
+  # convert to bed/bim/fam
+  system(glue("{plink} --file {prefix}.controls.gen --make-bed --out {prefix}"))
+  # cleaning all but bed/bim/fam
+  # https://stackoverflow.com/a/23179318/6103040
+  list <- grep(pattern = "^[^.]+$|\\.(?!(bed|bim|fam)$)([^.]+$)", 
+               x = list.files(simus.path, full.names = TRUE), 
+               perl = TRUE, value = TRUE)
[…]

### 2018-01-14 popres-pops.R

@@ -1,0 +1,20 @@
+list.pops <- list(
+  "SW Europe" = c("Portugal", "Spain"),
+  "Italy" = "Italy",
+  "SE Europe" = c("Greece", "Turkey", "Cyprus", "Macedonia"),
+  "Anglo-Irish Isles" = c("United", "Scotland", "Ireland"),
+  "FennoScandia" = c("?orway", "Finland", "Sweden", "Denmark"),
+  "Western Europe" = c("France", "Swiss-French", "Swiss-German",
+                       "Swiss-Italian", "Belgium"),
+  "Central Europe" = c("Germany", "Austria", "Poland", "?etherlands"),
+  "Eastern Europe" = c("Czech", "Romania", "Bulgaria", "Hungary",
+                       "Albania",  "Macedonia,", "Slovakia"),
+  "Former Yugoslavia" = c("Croatia", "Serbia", "Bosnia", "Slovenia", "Kosovo"),
+  "Former USSR" = c("Russian", "Ukraine", "Latvia")
+)
+mat.pops <- cbind(
+  unlist(list.pops, use.names = FALSE),
+  rep(names(list.pops), times = sapply(list.pops, length))
+)
+ind.match <- match(popres$fam$family.ID, mat.pops[, 1])
+Population <- mat.pops[ind.match, 2]

### 2018-01-14 cabios-template2/bigsnpr-review-supp.tex

--- 2017-09-11 cabios-template/draft.tex
@@
 
+\subsection{Pre-processing}
+
+\begin{figure}[htpb]
+\centerline{\includegraphics[width=160pt]{conversion-QC.pdf}}
+\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
+\end{figure}
+
+\begin{figure}[htpb]
+\centerline{\includegraphics[width=235pt]{imputation.pdf}}
+\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
+\end{figure}
+
 \subsection{Long-range LD regions}
 
-\begin{table}[!h]
+\begin{table}[htpb]
 \centering

### 2018-01-14 compare-GWAS-PRS-desktop.Rmd

@@
   # install.packages("ggplot2")
-  library(ggplot2)
+  library(tidyverse)
 })
@@
 ind.keep2 <- match(scan("plink.prune.in", what = ""), celiac$map$marker.ID)
-mean(ind.keep %in% ind.keep2)
+length(intersect(ind.keep, ind.keep2)) / length(union(ind.keep, ind.keep2))
 
@@
 # PRS
-(thrs <- c(0, exp(seq(log(0.2), log(550), length.out = 50))))
+(thrs <- c(0, exp(seq(log(0.1), log(100), length.out = 100))))
 lpS <- -predict(gwas.train)
@@
 aucs <- apply(prs, 2, AUC, target = y01[ind.test])
+# Number of predictor for each threshold
+nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
 ```
@@
 
-```{r}
+```{r, fig.width = 10}
 # Get a file with only training set
@@
 ind.keep2 <- match(clump$SNP, celiac$map$marker.ID)
-mean(ind.keep %in% ind.keep2)
+length(intersect(ind.keep, ind.keep2)) / length(union(ind.keep, ind.keep2))
 
@@
 
-scores <- matrix(NA_real_, length(ind.test), length(thrs))
+scores2 <- matrix(NA_real_, length(ind.test), length(thrs))
+nb.pred2 <- numeric(length(thrs))
 for (i in seq_along(thrs)) {
-  scores[, i] <- fread2(glue("plink.S{i}.profile"))$SCORESUM
+  tmp <- fread2(glue("plink.S{i}.profile"))
+  scores2[, i] <- tmp$SCORESUM
+  nb.pred2[i]  <- tmp$CNT[[1]] / 2
 }
@@
 #AUC
-aucs2 <- apply(scores, 2, AUC, target = y01[ind.test])
+aucs2 <- apply(scores2, 2, AUC, target = y01[ind.test])
 
-# Number of predictor for each threshold
-nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-bigstatsr:::MY_THEME(qplot(nb.pred, aucs, color = "A", log = "x", 
-                           cex = I(3), pch = I(1),
-                           xlab = "#predictors", ylab = "AUC"), coeff = 2) +
-  geom_line(show.legend = FALSE) + 
-  geom_point(aes(nb.pred, aucs2, color = "B"), cex = 3, pch = 3) + 
+bind_rows(
+  data_frame(nb = nb.pred,  auc = aucs,  Software = "bigsnpr"),
+  data_frame(nb = nb.pred2, auc = aucs2, Software = "PLINK")
+) %>%
+  ggplot() %>%
+  bigstatsr:::MY_THEME() +
+  geom_point(aes(nb, auc, color = Software, shape = Software), size = 5) +
+  geom_line(aes(nb, auc, color = Software, linetype = Software), show.legend = FALSE) +
+  scale_x_log10() + 
   scale_colour_manual(
-    name = NULL, 
     values = c("black", "red"),
     labels = c("PRS with bigstastr\nand bigsnpr", "PRS with plink")) + 
-  guides(color = guide_legend(override.aes = list(shape = c(1, 3)))) + 
-  theme(legend.position = c(0.7, 0.2)) 
+  scale_shape_manual(
+    values = c(1, 3),
+    labels = c("PRS with bigstastr\nand bigsnpr", "PRS with plink")) +
+  labs(x = "#predictors", y = "AUC") +
+  theme(legend.position = c(0.6, 0.2)) 
 ```

### 2018-01-18 compare-bioc.R
save

@@ -1,0 +1,141 @@
+library(bigsnpr)
+
+
+# Write temporary PLINK files
+tmp <- tempfile()
+bed <- snp_writeBed(snp_attach("backingfiles/celiacQC.rds"), paste0(tmp, ".bed"))
+
+library(SNPRelate)
+# Transform files in our format and attach them
+system.time(
+  gds <- snpgdsBED2GDS(bed.fn = bed, 
+                       fam.fn = sub("\\.bed$", ".fam", bed),
+                       bim.fn = sub("\\.bed$", ".bim", bed),
+                       out.gdsfn = paste0(tmp, ".gds"),
+                       cvt.snpid = "int")
+) # 33 sec
+
+genofile <- snpgdsOpen(gds)
+genofile
+snpgdsClose(genofile)
+
+
+# system.time(
+#   rds <- snp_readBed(bed, tmp)
+# ) # 43 sec
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+y01 <- celiac$fam$affection - 1
+NCORES <- nb_cores()
+
+# Pruning
+system.time(
+  snpset <- snpgdsLDpruning(genofile, method = "corr", ld.threshold = 0.2,
+                            num.thread = NCORES)
+) # 38 sec
+
+
+
+ind.excl <- union(which(CHR == 2), snp_indLRLDR(CHR, POS))
+
+system.time(
+  ind.keep <- snp_pruning(G, CHR, size = 500, is.size.in.bp = TRUE,
+                          exclude = snp_indLRLDR(CHR, POS),
+                          thr.r2 = 0.1,
+                          infos.pos = POS, ncores = NCORES)
+) # 73 sec
+length(ind.keep)
+
+system.time(
+  ind.keep2 <- snp_clumping(G, CHR, size = 500, is.size.in.bp = TRUE,
+                            exclude = snp_indLRLDR(CHR, POS),
+                            thr.r2 = 0.1,
+                            infos.pos = POS, ncores = NCORES)
+) # 66 sec
+length(ind.keep2)
+
+# system.time(
+#   pca <- snpgdsPCA(genofile, snp.id = celiac$map$marker.ID[ind.keep],
+#                    num.thread = NCORES, eigen.cnt = 10, algorithm = "randomized")
+# ) # 622
+
+
+system.time(
[…]

### 2018-01-18 optim-cmsa.R

@@ -1,0 +1,41 @@
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+y01 <- celiac$fam$affection - 1
+dim(G)
+
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+
+NCORES <- nb_cores()
+ind.excl <- snp_indLRLDR(CHR, POS)
+ind.keep <- snp_clumping(G, CHR, exclude = ind.excl, ncores = NCORES) 
+obj.svd <- snp_randomSVD(G, snp_scaleBinom(), ind.col = ind.keep, 
+                         ncores = NCORES, verbose = TRUE) 
+
+ind.train <- 1:12000
+system.time(
+  test <- big_spLogReg(G, y01[ind.train], ind.train = ind.train,
+                       alpha = 0.999, dfmax = 20e3)
+) # 490 sec
+
+ind.test <- setdiff(rows_along(G), ind.train)
+system.time(
+  preds <- predict(test, G, ind.test)
+) # 52 sec
+system.time(
+  aucs <- apply(preds, 2, function(pred) AUC(pred, y01[ind.test]))
+)
+summary(aucs[-1])
+#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+# 0.8156  0.8520  0.8596  0.8575  0.8725  0.8794 
+#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+# 0.8156  0.8520  0.8596  0.8575  0.8725  0.8796 
+
+plot(aucs[-1])
+
+# b <- test$beta
+# library(Matrix)
+# rowSums(b != 0)
+# colSums(b != 0)

### 2018-01-18 test-R-packages.R

@@ -1,0 +1,34 @@
+library(bigsnpr)
+
+# Write temporary PLINK files
+tmp <- tempfile()
+bed <- snp_writeBed(snp_attach("backingfiles/celiacQC.rds"), paste0(tmp, ".bed"))
+
+# gaston: Genetic Data Handling (QC, GRM, LD, PCA) & Linear Mixed Models
+library(gaston) 
+
+system.time(
+  x <- read.bed.matrix(bed)
+)
+
+x[1, 1]
+
+
+library(snpStats)
+snp <- read.plink(bed, 
+                  sub("\\.bed$", ".fam", bed),
+                  sub("\\.bed$", ".bim", bed)) ## All in memory
+
+
+library(GenABEL)  # pas updated depuis 2013
+
+?GenABEL::export.plink
+
+plink <- download_plink()
+system(glue::glue("{plink} --bfile {tmp} --recode --out {tmp}"))
+file.size(paste0(tmp, ".ped")) / 1024^3
+system.time(
+  test <- convert.snp.ped(pedfile = paste0(tmp, ".ped"),
+                          mapfile = paste0(tmp, ".map"),
+                          out = "test.dat"))
+)

### 2018-01-18 test-prsice.R

@@ -29,16 +29,17 @@ lpS <- -log10(pval)
 
 
 path <- "PRSice_linux.bugfix.20171116/"
-prefix <- "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
+# prefix <- "../Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
 tmp <- tempfile()
+snp_writeBed(celiac, paste0(tmp, ".bed"))
 
 system.time(
   system(glue::glue("Rscript {path}PRSice.R",
                     " --prsice {path}PRSice_linux",
                     " --base sumstats-height.txt",
-                    " --target {prefix}",
+                    " --target {tmp}",
                     " --out {tmp}",
-                    " --thread 6",
+                    " --thread {NCORES}",
                     # " --full",
                     # " --fastscore",
                     # " --cov-file covar.txt",
@@ -51,19 +52,18 @@ system.time(
                     " --binary-target T"))
 )
 
-
-cor.pval0 <- data.table::fread("PRSice.prsice", select = "P")[["P"]] 
+file.prsice <- paste0(tmp, ".prsice")
+cor.pval0 <- data.table::fread(file.prsice, select = "P")[["P"]]
 
 library(dplyr)
-thrs <- -log10(data.table::fread("PRSice.prsice", select = "Threshold")[["Threshold"]])
+thrs <- -log10(data.table::fread(file.prsice, select = "Threshold")[["Threshold"]])
 
 system.time(
   ind.keep <- snp_clumping(G, infos.chr = CHR, S = lpS, thr.r2 = 0.1,
-                           ind.row = which(y01 == 0),
                            size = 250, is.size.in.bp = TRUE, infos.pos = POS,
                            exclude = which(is.na(same) | (lpS < min(thrs))), 
                            ncores = NCORES)
-)
+) # 2 -> 19, 1 -> 29
 
 length(ind.keep)  # PRSice: 52946
 system.time(
@@ -79,9 +79,10 @@ y01 <- celiac$fam$affection - 1
 # cor.pval <- apply(prs, 2, function(scores) {
 #   summary(glm(y01 ~ scores, family = "binomial"))$coefficients[2, 4]
 # })
+library(magrittr)
 cor.pval2 <- prs %>%
   big_copy() %>%
[…]

### 2018-01-18 GWAS-comparison.Rmd
save

--- 2017-09-06 compare-GWAS-PRS-desktop.Rmd
@@
 ---
-title: "Comparing bigstatsr and bigsnpr with PLINK and EIGENSOFT"
+title: "Genome-Wide Association Study"
+subtitle: "Comparing bigstatsr and bigsnpr with PLINK and EIGENSOFT"
 output: html_document
@@
                       fig.width = 16, fig.asp = 0.7)
-options(width = 85)
+options(width = 90)
 ```
@@
 
-Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file.
+Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a bk/rds (format of packages bigstatsr and bigsnpr). See the notebook "preprocessing".
 
@@
   # install.packages("ggplot2")
-  library(ggplot2)
+  library(tidyverse)
 })
-# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
-smartpca <- "../EIG-6.1.4/bin/smartpca"
+# download.file("https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz",
+#               destfile = "EIG-6.1.4.tar.gz")
+# untar("EIG-6.1.4.tar.gz")
+smartpca <- "EIG-6.1.4/bin/smartpca"
 plink <- download_plink()
@@
 
-In this section, we will show how to perform a GWAS with or without our packages. For simplicity, everything will be done inside R.
+We will show how to perform a GWAS with or without our packages. For the sake of reproducibility and simplicity, everything will be done inside R (via possible system calls to command-line software).
 
@@
 ```{r}
+system.time(
   celiac <- snp_attach("backingfiles/celiacQC.rds")
+)
 G <- celiac$genotypes
@@
 
-ggsave("figures/celiac-pca.png", width = 1124, height = 945, scale = 1/75)
+ggsave("figures/celiac-pca.png", width = 1124, height = 945, scale = 1/90)
 ```
@@
 
-ggsave("figures/celiac-gwas-cut.png", width = 1475, height = 788, scale = 1/75)
+ggsave("figures/celiac-gwas-cut.png", width = 1475, height = 788, scale = 1/90)
 ```
@@
 # Pruning
-snp_writeBed(celiac, bedfile = "plink0.bed")
-prefix_bed <- "plink0"
+prefix_bed <- tempfile()
+snp_writeBed(celiac, bedfile = paste0(prefix_bed, ".bed"))
 write.table2(LD.wiki34, "LRLDR.txt")
 system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
+  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}",
               " --exclude LRLDR.txt --range",
@@
 
-ind.keep2 <- match(scan("plink.prune.in", what = ""), celiac$map$marker.ID)
-mean(ind.keep %in% ind.keep2)
+ind.keep2 <- match(scan(paste0(prefix_bed, ".prune.in"), what = ""), celiac$map$marker.ID)
+length(intersect(ind.keep, ind.keep2)) / length(union(ind.keep, ind.keep2))
 
@@
 system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
+  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}",
               " --extract keep.in --make-bed",
@@
 fastpca <- function(prefix, fast = TRUE) {
+
   parfile <- glue("{prefix}.pca.par")
@@
 system.time(
-  obj.svd2 <- fastpca("plink")
+  obj.svd2 <- fastpca(tmp)
 )
 
-# FastPCA is not accurate enough (see benchmark-pca.html)
+# FastPCA is not accurate enough (see Notebook "benchmark-pca")
 round(100 * cor(obj.svd2, obj.svd$u), 1)
@@
 system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
+  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}",
               " --logistic hide-covar",
@@
 
-gwas2 <- fread2("plink.assoc.logistic")
+gwas2 <- fread2(paste0(prefix_bed, ".assoc.logistic"))
 all.equal(gwas2$P, predict(gwas, log10 = FALSE)) 
 ```
-
-## Example of Polygenic Risk Scores (PRS)
-
-In this section, we will show how to create a PRS with or without our packages. For simplicity, everything will be done inside R.
-
-```{r}
-# Split in training/test set (80%/20%)
-ind.train <- sort(sample(nrow(G), 0.8 * nrow(G)))
-ind.test <- setdiff(rows_along(G), ind.train)
-```
-
-### With bigstatsr and bigsnpr
-
-```{r}
-# GWAS on training set
-y01 <- celiac$fam$affection - 1
-system.time(
-  gwas.train <- big_univLogReg(G, y01.train = y01[ind.train],
-                               ind.train = ind.train, 
-                               covar.train = obj.svd$u[ind.train, ], 
-                               ncores = NCORES)
-)
-
-# Clumping
-system.time(
-  ind.keep <- snp_clumping(G, infos.chr = celiac$map$chromosome,
-                           ind.row = ind.train,
-                           S = abs(gwas.train$score),
-                           size = 250, # as PLINK default
-                           is.size.in.bp = TRUE,
-                           infos.pos = celiac$map$physical.pos,
-                           ncores = NCORES)
-)
-
-# PRS
-(thrs <- c(0, exp(seq(log(0.2), log(550), length.out = 50))))
-lpS <- -predict(gwas.train)
-system.time(
-  prs <- snp_PRS(G, betas.keep = gwas.train$estim[ind.keep],
-                 ind.test = ind.test,
-                 ind.keep = ind.keep,
-                 lpS.keep = lpS[ind.keep], 
-                 thr.list = thrs)
-)
-
-# AUC
-aucs <- apply(prs, 2, AUC, target = y01[ind.test])
-```
-
-### With PLINK (and R)
-
-```{r}
-# Get a file with only training set
-write.table2(cbind(celiac$fam[ind.train, 1:2], 
-                   obj.svd$u[ind.train, ]), "covar_train.txt")
-# GWAS on training set
-system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
-              " --keep covar_train.txt",
-              " --logistic hide-covar beta",
-              " --covar covar_train.txt",
-              " --threads {NCORES}"))
-)
-
-gwas2 <- fread2("plink.assoc.logistic")
-all.equal(gwas2$P, 10^predict(gwas.train))
-
-# Clumping
-system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
-              " --keep covar_train.txt",
-              " --clump plink.assoc.logistic",
-              " --clump-p1 1 --clump-p2 1 --clump-r2 0.2",
-              " --threads {NCORES}"))
-)
-
-clump <- fread2("plink.clumped")
-ind.keep2 <- match(clump$SNP, celiac$map$marker.ID)
-mean(ind.keep %in% ind.keep2)
-
-gwas2.keep <- gwas2[ind.keep, ]
-write.table2(gwas2.keep, "assoc_clumped.txt")
-
-# PRS
-write.table2(celiac$fam[ind.test, 1:2], "ind_test.txt")
-write.table2(cbind(paste0("S", seq_along(thrs)), 0, 10^(-thrs)),
-             "range.txt")
-
-system.time(
-  system(glue("{plink} --bfile {prefix_bed}",
-              " --keep ind_test.txt",
-              " --extract assoc_clumped.txt",
-              " --score assoc_clumped.txt 2 4 7 sum",
-              " --q-score-range range.txt assoc_clumped.txt 2 9",
-              " --threads {NCORES}"))
-)
-
-scores <- matrix(NA_real_, length(ind.test), length(thrs))
-for (i in seq_along(thrs)) {
-  scores[, i] <- fread2(glue("plink.S{i}.profile"))$SCORESUM
-}
-
-#AUC
-aucs2 <- apply(scores, 2, AUC, target = y01[ind.test])
-
-# Number of predictor for each threshold
-nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-bigstatsr:::MY_THEME(qplot(nb.pred, aucs, color = "A", log = "x", 
-                           cex = I(3), pch = I(1),
-                           xlab = "#predictors", ylab = "AUC"), coeff = 2) +
-  geom_line(show.legend = FALSE) + 
-  geom_point(aes(nb.pred, aucs2, color = "B"), cex = 3, pch = 3) + 
-  scale_colour_manual(
-    name = NULL, 
-    values = c("black", "red"),
-    labels = c("PRS with bigstastr\nand bigsnpr", "PRS with plink")) + 
-  guides(color = guide_legend(override.aes = list(shape = c(1, 3)))) + 
-  theme(legend.position = c(0.7, 0.2)) 
-```
-
-```{r}
-# Cleaning
-file.remove(c(list.files(pattern = "plink*|*.txt"), "keep.in"))
-```

### 2018-01-18 cabios-template2/bigsnpr-review.tex
save

@@
 Then, implementation of GWAS analyses also depends on the data format and model analyzed.
-For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1.9 is limited to best guess genotypes because of its input file format.
+For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data,
+{\color{red}
+while PLINK version 1.9 is limited to best guess genotypes because of its input file format.
 Note that PLINK 2.0 will completely handle this type of data, but is still in the alpha stage of development.
+}
 Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants.
@@
 Most modern SNP chips provide genotype data with large call-rates.
-For example, the Celiac data we use in this paper present only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control.
+For example, the Celiac data we use in this paper presents only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control.
 Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
@@
 Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the higher MAF is kept.
-Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all so we advise to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping ([SUPMAT]).
+Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]).
+So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping .
+In practice, we tested the effect of using clumping instead of pruning for the thinning step before PCA and the results were very similar.
 }
@@
 The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlation} as the original dataset.
+For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset.
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
@@
 All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data. 
+{\color{red}
+We now provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs ([SUPMAT]) \cite[]{hayward2016complex}.
+}
 

### 2018-01-18 preprocessing.Rmd
save

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
+                      fig.width = 16, fig.asp = 0.7)
+options(width = 90)
 ```
@@
 
+plink <- download_plink()
 NCORES <- nb_cores()
@@
 popresQC.bed <- snp_plinkQC(
-  prefix.in = "../POPRES_data/POPRES_allchr",
+  prefix.in = "data/POPRES_allchr",
+  plink.path = plink,
   file.type = "--file", # ped/map
@@
 
-popresQC2.bed <- snp_plinkIBDQC(popresQC.bed, ncores = NCORES)
+popresQC2.bed <- snp_plinkIBDQC(
+  bedfile.in = popresQC.bed, 
+  plink.path = plink, 
+  ncores = NCORES
+)
 ```
@@
 celiacQC.bed <- snp_plinkQC(
-  prefix.in = "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300",
+  prefix.in = "data/FinnuncorrNLITUK1UK3hap300",
+  plink.path = plink,
   geno = 0.05,
@@
 
-celiacQC2.bed <- snp_plinkIBDQC(celiacQC.bed, ncores = NCORES)
+celiacQC2.bed <- snp_plinkIBDQC(
+  celiacQC.bed, 
+  plink.path = plink,
+  ncores = NCORES
+)
 ```
@@
 ```{r}
+system.time(
   popresQC.rds <- snp_readBed(popresQC2.bed, "backingfiles/popresQC")
+)
 
@@
 ```{r}
-system.time(
 celiac <- snp_attach(celiacQC.rds)
-)
 str(celiac, max.level = 2)
@@
 
+The above plot is useful for post-processing the imputation.
+No SNP (point) is above the red line. This means that no SNP has more than 1% of error.
+
 ```{r}

### 2018-01-18 proof-same-decomposition.Rmd
save

---
title: "Same eigen decomposition: proof"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proof for the same pearson correlation coefficient

Let us note $x$ and $y$ two SNPs and $\overline{x}$ the mean of $x$.

Then the pearson correlation coefficient between $x$ and $y$ is: 
$$r = \dfrac{\sum\limits_{i = 1}^n (x_i - \overline{x}) (y_i - \overline{y})}{\sqrt{\sum\limits_{i = 1}^n (x_i - \overline{x})^2} \sqrt{\sum\limits_{i = 1}^n (y_i - \overline{y})^2}}~~.$$

If we note $x_K$ and $y_K$ the SNPs $x$ and $y$ replicated $K$ times, then, $\overline{x_K} = \frac{K \sum\limits_{i = 1}^n x_i}{K~n} = \frac{1}{n} \sum\limits_{i = 1}^n x_i = \overline{x}$ and $r_K$ the pearson correlation coefficient between $x_K$ and $y_K$ is 
$$r_K = \dfrac{K \sum\limits_{i = 1}^n (x_i - \overline{x_K}) (y_i - \overline{y_K})}{\sqrt{K \sum\limits_{i = 1}^n (x_i - \overline{x_K})^2} \sqrt{K \sum\limits_{i = 1}^n (y_i - \overline{y_K})^2}} = \dfrac{\sum\limits_{i = 1}^n (x_i - \overline{x}) (y_i - \overline{y})}{\sqrt{\sum\limits_{i = 1}^n (x_i - \overline{x})^2} \sqrt{\sum\limits_{i = 1}^n (y_i - \overline{y})^2}} = r~.$$

## Proof for the same eigen analysis

First, as $\overline{x_K}$ = $\overline{x}$, then the allele frequency of $x_K$ is the same as for $x$ so that the scaling $\frac{x - 2 p}{\sqrt{2 p (1 - p)}}$ is also the same (because $p = \overline{x} / 2$).

For the exact singular value decomposition of $G = U \Delta V^T$ (where $G$ is the scaled genotype matrix), we can first compute $\Sigma = G^T G = V \Delta^2 V^T$, then remark that $\Sigma V = V \Delta^2$ so that $V$ is the matrix of the eigen vectors of $\Sigma$ and $\Delta^2$ is the matrix of the eigen values of $\Sigma$. Finally to get $U$, we can compute $G V \Delta^{-1} = U \Delta V^T V \Delta^{-1} = U \Delta \Delta^{-1} = U$.

For replicated individuals, we want the decomposition of $G_K = U_K \Delta_K V_K^T$. Then, $\Sigma_K = G_K^T G_K = KG^T G = K \Sigma$ so that $V_K = V$ (same PC loadings) and $\Delta_K^2 = K \Delta^2$ resulting in $\Delta_K = \sqrt{K} \Delta$ (same eigen values, up to a constant). Finally $U_K = G_K V_K \Delta_K^{-1} = G_K V \left(\sqrt{K} \Delta\right)^{-1} = \frac{1}{\sqrt{K}} G_K V \Delta^{-1}$ (PCs scores are the same, up to a constant).

### 2018-01-18 public-data.Rmd
save

@@
 system.time(
-  bigsnp <- snp_attach(rds)
+  bigsnp <- snp_attach("backingfiles/dogs.rds")
 )
@@
 ```
-
-```{r}
-
-```

### 2018-01-18 diff-PCs-clumping.R

@@ -1,0 +1,47 @@
+library(bigsnpr)
+bigsnp <- snp_attach("backingfiles/dogs.rds")
+G <- bigsnp$genotypes
+CHR <- bigsnp$map$chromosome
+POS <- bigsnp$map$physical.pos
+NCORES <- nb_cores()
+
+system.time(
+  ind.keep <- snp_pruning(G, CHR, size = 500, is.size.in.bp = TRUE,
+                          # thr.r2 = 0.1,
+                          infos.pos = POS, ncores = NCORES)
+) # 9 sec
+length(ind.keep)
+
+system.time(
+  ind.keep2 <- snp_clumping(G, CHR, size = 500, is.size.in.bp = TRUE,
+                            # thr.r2 = 0.1,
+                            infos.pos = POS, ncores = NCORES)
+) # 6 sec
+length(ind.keep2)
+
+
+system.time(
+  obj.svd <- big_randomSVD(G, snp_scaleBinom(), ind.col = ind.keep, k = 50, ncores = NCORES)
+) # 8 / 18
+
+system.time(
+  obj.svd2 <- big_randomSVD(G, snp_scaleBinom(), ind.col = ind.keep2, k = 50, ncores = NCORES)
+) # 7 / 18
+
+diag(round(100 * cor(obj.svd$u, obj.svd2$u), 1))
+
+cowplot::plot_grid(
+  plot(obj.svd), 
+  plot(obj.svd2), 
+  ncol = 1
+)
+
+
+library(ggplot2)
+PCs <- 26:27
+cowplot::plot_grid(
+  plot(obj.svd, type = "scores", scores = PCs),
+  plot(obj.svd2, type = "scores", scores = PCs), 
+  ncol = 1
+)
+

### 2018-01-23 GWAS-comparison.Rmd
GWAS + PRS

@@
                       fig.width = 16, fig.asp = 0.7)
-options(width = 90)
+options(width = 110)
 ```
@@
 ```{r}
-# GWAS
+# GWAS (logistic)
+y01 <- celiac$fam$affection - 1
 system.time(
-  gwas <- big_univLogReg(G, y01.train = celiac$fam$affection - 1,
-                         covar.train = obj.svd$u, ncores = NCORES)
+  gwas <- big_univLogReg(G, y01.train = y01, covar.train = obj.svd$u, 
+                         ncores = NCORES)
 )
+gwas.pval <- predict(gwas, log10 = FALSE)
 
 labels <- 1:22; labels[c(15, 17, 19, 21)] <- ""
-snp_manhattan(snp_gc(gwas), CHR, POS, 
-              coeff = 2, labels = labels) + 
+snp_manhattan(snp_gc(gwas), CHR, POS, coeff = 2, labels = labels) + 
   geom_hline(yintercept = -log10(5e-8), col = "red") +
@@
 
+```{r}
+# GWAS (linear)
+y <- rnorm(nrow(G))
+system.time(
+  gwas.lin <- big_univLinReg(G, y.train = y, covar.train = obj.svd$u, 
+                             ncores = NCORES)
+)
+gwas.lin.pval <- predict(gwas.lin, log10 = FALSE) 
+```
+
 ### With PLINK and EIGENSOFT (and R)
@@
 ```{r}
+# Write temporary PLINK files
+prefix_bed <- tempfile()
+bed <- snp_writeBed(snp_attach("backingfiles/celiacQC.rds"), 
+                    bedfile = paste0(prefix_bed, ".bed"))
+```
+
+```{r}
+# Long-range LD regions
+write.table2(LD.wiki34, "LRLDR.txt")
 # Pruning
-prefix_bed <- tempfile()
-snp_writeBed(celiac, bedfile = paste0(prefix_bed, ".bed"))
-write.table2(LD.wiki34, "LRLDR.txt")
 system.time(
@@
 ind.keep2 <- match(scan(paste0(prefix_bed, ".prune.in"), what = ""), celiac$map$marker.ID)
+# Same sets of SNPs
 length(intersect(ind.keep, ind.keep2)) / length(union(ind.keep, ind.keep2))
+```
 
+```{r}
 # Make a pruned dataset
@@
 system.time(
-  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}",
+  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}_sub",
               " --extract keep.in --make-bed",
@@
 )
+```
 
+```{r}
 # PCA with EIGENSOFT
@@
 system.time(
-  obj.svd2 <- fastpca(tmp)
+  obj.svd2 <- fastpca(paste0(prefix_bed, "_sub"))
 )
 
-# FastPCA is not accurate enough (see Notebook "benchmark-pca")
+# FastPCA is not accurate enough (see notebook "benchmark-pca")
 round(100 * cor(obj.svd2, obj.svd$u), 1)
@@
 gwas2 <- fread2(paste0(prefix_bed, ".assoc.logistic"))
-all.equal(gwas2$P, predict(gwas, log10 = FALSE)) 
+all.equal(gwas2$P, gwas.pval) 
 ```
+
+```{r}
+# GWAS (linear)
+write.table2(cbind(celiac$fam[, 1:2], y), "qt.phe")
+system.time(
+  system(glue("{plink} --bfile {prefix_bed} --out {prefix_bed}",
+              " --linear hide-covar --pheno qt.phe",
+              " --covar covar.txt",
+              " --threads {NCORES}"))
+)
+
+gwas2.lin <- fread2(paste0(prefix_bed, ".assoc.linear"))
+all.equal(gwas2.lin$P, gwas.lin.pval) 
+```
+
+### With R packages SNPRelate and GWASTools
+
+```{r}
+library(SNPRelate)
+
+# Transform files in our format and attach them
+system.time(
+  gds <- snpgdsBED2GDS(bed.fn = bed, 
+                       fam.fn = sub("\\.bed$", ".fam", bed),
+                       bim.fn = sub("\\.bed$", ".bim", bed),
+                       out.gdsfn = paste0(prefix_bed, ".gds"),
+                       cvt.snpid = "int")
+)
+```
+
+```{r}
+system.time(
+  genofile <- snpgdsOpen(gds)
+)
+```
+
+```{r}
+# Pruning
+system.time(
+  snpset <- snpgdsLDpruning(genofile, method = "corr", ld.threshold = 0.2,
+                            num.thread = NCORES)
+) # 38 sec
+ind.keep3 <- unlist(snpset, use.names = FALSE)
+# Not the same set of SNPs (because not the same algo)
+length(intersect(ind.keep3, ind.keep2)) / length(union(ind.keep3, ind.keep2))
+```
+
+```{r}
+snpgdsClose(genofile)
+# Need sudo yum install netcdf-devel.x86_64
+library(GWASTools)
+(genofile <- GdsGenotypeReader(gds, YchromCode=24L, XYchromCode=25L))
+scanID <- getScanID(genofile)
+# sex <- getVariable(genofile, "sample.annot/sex")
+phenotype <- getVariable(genofile, "sample.annot/phenotype")
+covar <- obj.svd$u
+colnames(covar) <- paste0("PC", 1:10)
+scanAnnot <- ScanAnnotationDataFrame(
+  cbind.data.frame(scanID, pheno = phenotype - 1, covar)
+)
+snpID <- getSnpID(genofile)
+chromosome <- getChromosome(genofile)
+position <- getPosition(genofile)
+alleleA <- getAlleleA(genofile)
+alleleB <- getAlleleB(genofile)
+snpAnnot <- SnpAnnotationDataFrame(data.frame(
+  snpID = seq_along(snpID), chromosome, position, rsID = snpID,
+  alleleA, alleleB, stringsAsFactors=FALSE),
+  YchromCode=24L, XYchromCode=25L)
+
+genoData <- GenotypeData(genofile, scanAnnot = scanAnnot, snpAnnot = snpAnnot)
+
+# Logistic regression
+system.time(
+  gwas3 <- assocRegression(genoData, "pheno", 
+                           model.type = "logistic",
+                           covar = colnames(covar),
+                           snpEnd = 10e3)
+) * ncol(G) / 10e3
+all.equal(gwas3$Wald.pval, gwas.pval[rows_along(gwas3)]) 
+# assocRegression uses glm() so that it is the most precise
+# We are very precise and PLINK is the least precise
+true.pval <- sapply(1:9, function(j) {
+  summary(glm(y01 ~ G[, j] + covar, family = "binomial"))$coefficients[2, 4]
+})
+rbind(`true (glm)` = true.pval, bigstatsr = gwas.pval[1:9], 
+      plink = gwas2$P[1:9], gwastools = gwas3$Wald.pval[1:9])
+```
+
+```{r}
+scanAnnot <- ScanAnnotationDataFrame(
+  cbind.data.frame(scanID, pheno = y, covar)
+)
+
+genoData <- GenotypeData(genofile, scanAnnot = scanAnnot, snpAnnot = snpAnnot)
+
+# Linear regression
+system.time(
+  gwas3.lin <- assocRegression(genoData, "pheno", 
+                               model.type = "linear",
+                               covar = colnames(covar),
+                               snpEnd = 10e3)
+) * ncol(G) / 10e3
+all.equal(gwas3.lin$Wald.pval, gwas.lin.pval[rows_along(gwas3)])
+
+# Again, we are very precise
+true.lin.pval <- sapply(1:9, function(j) {
+  summary(lm(y ~ G[, j] + covar))$coefficients[2, 4]
+})
+rbind(`true (lm)` = true.lin.pval, bigstatsr = gwas.lin.pval[1:9], 
+      plink = gwas2.lin$P[1:9], gwastools = gwas3.lin$Wald.pval[1:9])
+```
+
+```{r}
+# Cleaning
+file.remove(c("LRLDR.txt", "covar.txt", "keep.in", "qt.phe"))
+```

### 2018-01-23 PRS-comparison.Rmd
GWAS + PRS

---
title: "Polygenic Risk Scores (PRS)"
subtitle: "Comparing bigstatsr and bigsnpr with PRSice"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 8, fig.asp = 0.7)
options(width = 110)
```

## Prerequisites

Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a bk/rds (format of packages bigstatsr and bigsnpr). See the notebook "preprocessing".

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # install.packages("glue")
  library(glue)      
  # install.packages("ggplot2")
  library(tidyverse)
})
# install.packages("data.table")
fread2 <- function(...) data.table::fread(data.table = FALSE, ...)
NCORES <- nb_cores()
```

We also need some summary statistics. Let's use ones for the height.

```{r, eval=FALSE}
download.file("http://portals.broadinstitute.org/collaboration/giant/images/0/01/GIANT_HEIGHT_Wood_et_al_2014_publicrelease_HapMapCeuFreq.txt.gz",
              destfile = "sumstats-height.txt.gz")
R.utils::gunzip("sumstats-height.txt.gz")
```

## Example of Polygenic Risk Scores (PRS)

We will show how to perform a GWAS with or without our packages. For the sake of reproducibility and simplicity, everything will be done inside R (via possible system calls to command-line software).

### With PRSice v2.1.0.beta

```{r}
# Write temporary PLINK files
prefix_bed <- tempfile()
bed <- snp_writeBed(snp_attach("backingfiles/celiacQC.rds"), 
                    bedfile = paste0(prefix_bed, ".bed"))
```

```{r}
# download.file("https://github.com/choishingwan/PRSice/releases/download/2.1.0.beta/PRSice_linux.zip",
#               destfile = "PRSice_linux.zip")
# unzip("PRSice_linux.zip", files = c("PRSice.R", "PRSice_linux"))
system("chmod +x PRSice_linux")
# Use PRSice
system.time(
  system(glue("Rscript PRSice.R",
              " --prsice PRSice_linux",
              " --base sumstats-height.txt",
              " --target {prefix_bed}",
              " --out {prefix_bed}",
              " --thread {NCORES}",
              " --pvalue p",
              " --stat b",
              " --beta T",
              " --snp MarkerName",
              " --A1 Allele1 --A2 Allele2",
              " --binary-target T"))
)
# Read results
file.prsice <- paste0(prefix_bed, ".prsice")
res.prsice <- fread2(file.prsice)
```

### With bigstastr and bigsnpr

```{r}
system.time(
  celiac <- snp_attach("backingfiles/celiacQC.rds")
)
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
```

```{r}
# Read the summary stats
system.time(
  sumstats <- fread2("sumstats-height.txt")
)
```

```{r}
system.time({
  # Get the SNP indices that correspond in sumstats
  indSNP <- match(celiac$map$marker.ID, sumstats$MarkerName)
  betas <- sumstats$b[indSNP]
  pval <- sumstats$p[indSNP]
  lpS <- -log10(pval)
  # And check if the reference alleles are switched
  same <- same_ref(celiac$map$allele1, celiac$map$allele2, 
                   sumstats$Allele1[indSNP], sumstats$Allele2[indSNP])
})
# Non-matching SNPs
sum(is.na(indSNP))
```

```{r}
# Use the same p-value thresholds
thrs <- -log10(res.prsice$Threshold)
# Clumping
system.time(
  ind.keep <- snp_clumping(G, infos.chr = CHR, S = lpS, thr.r2 = 0.1,
                           size = 250, is.size.in.bp = TRUE, infos.pos = POS,
                           exclude = which(is.na(same) | (lpS < min(thrs))), 
                           ncores = NCORES)
)
```
```{r}
# PRS
system.time(
  prs <- snp_PRS(G, betas.keep = betas[ind.keep],
                 ind.test = rows_along(G),
                 same.keep = same[ind.keep], 
                 lpS.keep = lpS[ind.keep], 
                 ind.keep = ind.keep,
                 thr.list = thrs)
)
```

```{r}
# p-values
system.time(
  pval.bigsnpr <- prs %>%
    big_copy() %>%
    big_univLogReg(celiac$fam$affection - 1) %>%
    predict(log10 = FALSE)
)
```

### Verification

```{r}
# p-values
cor(log10(pval.bigsnpr), log10(res.prsice$P))
bigstatsr:::MY_THEME(qplot(pval.bigsnpr, res.prsice$P, log = "xy")) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(x = "P-values returned with bigsnpr (log-scale)",
       y = "P-values returned with PRSice (log-scale)")
```

```{r}
# Number of SNPs for each threshold
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr, na.rm = TRUE))
bigstatsr:::MY_THEME(qplot(nb.pred, res.prsice$Num_SNP)) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(x = "Number of SNPs with bigsnpr",
       y = "Number of SNPs with PRSice")
```

These small differences are due to the way LD is assessed in the clumping algorithm.

### 2018-01-23 cabios-template2/bigsnpr-review-supp.tex
GWAS + PRS

@@
 
+\begin{figure}[htpb]
+\centerline{\includegraphics[width=\textwidth]{post-imputation}}
+\caption{For each SNP (point), the estimated proportion of imputation errors ($\hat{p}(error~|~NA)$) vs the proportion of missing values ($p(NA) > 0.001$).
+These results come from the imputation of the Celiac dataset with function snp\_fastImpute (Supplementary notebook ``preprocessing''). 
+Colored curves are representing the estimated proportion of wrong genotypes per SNP ($\hat{p}(error~\&~NA) = \hat{p}(error~|~NA) \cdot p(NA)$).
+This particularly shows that no SNP has more than 1\% of wrong genotypes, allowing for no post-processing.}\label{fig:post-imputation}
+\end{figure}
+
 \vspace*{-12pt}

### 2018-01-23 cabios-template2/bigsnpr-review.tex
GWAS + PRS

@@
 For the sake of reproducibility, one could use PLINK directly from R via systems calls.
-For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.
-vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
+For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
 Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs.
@@
 So, we provide two functions for imputing missing values of genotyped SNPs.
-Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.
-imputation servers for human data \cite[]{mccarthy2016reference}.
+Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}.
 The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}).
@@
 Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP.
-Those latter individuals are further separated in a training set and a validation set (e.g.
-80\% training and 20\% validation).
+Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation).
 The training set is used to build the XGBoost model for predicting missing data.
@@
 The second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
-The currently available models are linear and logistic regressions and Support Vector Machine (SVM).
+The currently available models are very fast sparse linear and logistic regressions.
 These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}.
 Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}.
-The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}.
-These modifications allow to include covariates in the models and to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr.
+The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}.
+These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr, to remove the need of the choice of the regularization parameter and to fasten the overall algorithms.
 
@@
 {\color{red}
-We now provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs ([SUPMAT]) \cite[]{hayward2016complex}.
+In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
 }
@@
 We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
-For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.
-pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
+For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
 Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}).
@@
  \cline{2-3}
- & PLINK and FastPCA & bigstatsr and bigsnpr \\
+ & PLINK 1.9 and FastPCA & bigstatsr and bigsnpr \\
 \hline
@@
 \end{center}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) for making a GWAS for the Celiac dataset.
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
 \label{tab:bench-gwas}
@@
  \cline{2-3}
- & PLINK & bigstatsr and bigsnpr \\
+ & PRSice & bigstatsr and bigsnpr \\
 \hline
-GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
-Clumping & 49 / 58 & 10 / 35 \\
-PRS & 9 / 10 & 2 / 3 \\
+Reading PLINK files &  & 6 / 20 \\
+Reading summary stats &  & 4 /   \\
+Clumping &  & 9 /  \\
+PRS &  & 2 /  \\
+Compute pvalues &  & 1 /  \\
 \hline
-Total & 290 / 307 & 190 / 688 \\
+Total & 22 /  & 22 /  \\
 \hline
@@
 \end{center}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset.
-The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
+\caption{Execution times with bigstatsr and bigsnpr compared to PRSice for making a PRS on the Celiac dataset based on summary statistics for Height.
+The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
 \label{tab:bench-prs}
@@
 
-For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset.
+For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset (with no missing value). 
 The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value. 
 We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}).
-Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017).
-Overall, snp\_fastImpute made 4.7\% of imputation errors whereas Beagle made only 3.1\% of errors but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
-We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).
+Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017) 
+{\color{red}
+by counting the percentage of imputation errors (when the imputed genotype is different from the true genotype)
+}.
+Overall, in 3 runs, snp\_fastImpute made only 4.7\% of imputation errors and Beagle made only 3.1\% of errors.
+Yet, it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
+{\color{red}
+We also note that snp\_fastImpute makes less 0/2 switching errors, i.e.\ imputing with a 0 where it is in fact a 2 or the contrary (Supplementary ..notebook.. ``imputation'').
+We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~\ref{fig:error-impute}), which can be useful for post-processing of the imputation (Figure~\ref{fig:post-imputation}).
+}
 For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle did not finish imputing chromosome 1 in 48 hours.

### 2018-01-24 GWAS-comparison.Rmd
run on laptop

@@
 # Get population from external files
-pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/",
-                        pattern = "cluster_*",
-                        full.names = TRUE)
+pop.files <- list.files(path = "data", pattern = "cluster_*", full.names = TRUE)
 pop <- snp_getSampleInfos(celiac, pop.files)[[1]]

### 2018-01-24 GWAS-comparison.Rmd
save

@@
 ```{r}
+# PCA
+system.time(
+  pca <- snpgdsPCA(genofile, snp.id = ind.keep, eigen.cnt = 10,
+                   num.thread = NCORES, algorithm = "randomized")
+)
+```
+
+```{r}
+# Format objects
 snpgdsClose(genofile)
@@
 genoData <- GenotypeData(genofile, scanAnnot = scanAnnot, snpAnnot = snpAnnot)
+```
 
+```{r}
 # Logistic regression

### 2018-01-24 cabios-template2/bigsnpr-review.tex
save

@@
 \begin{adjustbox}{max width=235pt}
-\begin{tabular}{|c|c|c|}
+\begin{tabular}{|c|c|c|c|}
 \hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
- \cline{2-3}
- & PLINK 1.9 and FastPCA & bigstatsr and bigsnpr \\
+\multirow{3}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{3}{c|}{Execution times (in seconds)} \\
+ \cline{2-4}
+ & FastPCA & bigstatsr  & SNPRelate \\
+ & PLINK 1.9 & bigsnpr & GWASTools \\
 \hline
-Reading PLINK files & n/a & 5 / 20 \\
-Pruning & 4 / 4 & 14 / 52 \\
-Computing 10 PCs & 306 / 315 & 58 / 180 \\
-GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
+Reading PLINK files & n/a & 6 / 20 & 12 /  \\
+Pruning & 4 / 4 & 14 / 52 & 33 / \\  
+Computing 10 PCs & 304 / 315 & 61 / 180 &  /  \\
+GWAS (binary phenotype) & 336 / 293 & 301 / 861 & 15806 /  \\
+GWAS (continuous phenotype) & 1339 /  & 9 /  & 5933 /  \\
 \hline
-Total & 649 / 612 & 378 / 1113 \\
+Total & 649 / 612 & 378 / 1113 &  /  \\
 \hline
@@
 \end{center}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) for making a GWAS for the Celiac dataset.
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset.
 The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
@@
 \hline
-\multirow{2}{*}{Operation} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
+\multirow{2}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
  \cline{2-3}

### 2018-01-25 cabios-template2/bigsnpr-review.tex
new draft of review

@@
 \abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-This increase in data size is a major challenge severely slowing down genomic analyses.
-Specialized software for every part of the analysis pipeline have been developed to handle large genomic data.
-However, combining all these software into a single data analysis pipeline might be technically difficult.
-\\
-\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework.
+{\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
+\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R.
 To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM.
@@
 
-Genome-wide datasets produced for association studies have dramatically increased in size over the past years.
-A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation.
-This diverse and extremely rich software environment has been of tremendous benefit for the genetic community.
-However, it has two limitations: analysis pipelines are becoming very complex and researchers have limited access to diverse analysis tools due to growing data sizes.
+{\color{red}
+Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
+As a consequence, it has been required to continuously optimize algorithms and this has led to the reimplementation of some software that would have become obsolete otherwise.
+For computing Principal Component Analysis, commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
+PLINK 1.07, which is a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.
 
-Consider first the basic tools necessary to perform a standard genome-wide analysis.
-Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}.
-Similarly, quality control software for genome-wide analysis have been developed such as PLINK and the Bioconductor package GWASTools \cite[]{Gogarten2012}.
-There are also several software for the computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
-Then, implementation of GWAS analyses also depends on the data format and model analyzed.
-For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data,
-{\color{red}
-while PLINK version 1.9 is limited to best guess genotypes because of its input file format.
-Note that PLINK 2.0 will completely handle this type of data, but is still in the alpha stage of development.
-}
-Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants.
-As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions.
-Overall, this might be a barrier to data exploration.
-%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.
-
-Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
-This is particularly a burden for commonly used analysis languages such as R, Python and Perl.
-Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented.
+Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
+This is particularly a burden for commonly used analysis languages such as R.
+For analyzing genotype datasets in R, a range of software are available, including R packages snpStats, GenABEL, SNPRelate and GWASTools.
+Packages snpStats and GenABEL, which have been widely used and cited, are now outdated with current data sizes -- package snpStats still uses in-memory data and package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
+Solving the memory issues for languages such as R would give access to a broad range of tools for data analysis that have already been implemented.
 Hopefully, strategies have been developed to avoid loading large datasets in RAM.
@@
 Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
-
-%\enlargethispage{12pt}
+}
 
@@
 
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
+For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
 While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
-Package bigstatsr use the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
+Package bigstatsr uses the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra.
 The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.
 
-\begin{figure}[!tpb]
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=235pt]{svd.pdf}}
@@
 SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}.
-There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium, two of them named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}.
-First, pruning, the most naive approach, is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
-A variant of pruning is clumping. 
+There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium.
+Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}.
+First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
 {\color{red}
-Clumping is useful if a statistic is available to sort the SNPs by importance.
+Clumping is useful if a statistic is available for sorting the SNPs by importance.
 Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome.
-For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance.
-Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the higher MAF is kept.
-Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]).
-So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping .
-In practice, we tested the effect of using clumping instead of pruning for the thinning step before PCA and the results were very similar.
+For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the MAF as the statistic of importance. 
+Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest minor allele frequency (MAF) is kept.
+Yet, in some worst-case scenario, this pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
+So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
+In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
 }
@@
 As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
-Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}.
-However, the pattern of LD might be population specific, so we developed an algorithm that automatically detects these regions and removes them.
-This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}.
+Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in long-range LD regions begore computing PCs.
+However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them.
+This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}.
 Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions.
 Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA.
-This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.
+This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.
 
@@
 Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
-Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.
-We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
+Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}.
+Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
 \begin{multline}
@@
 and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
+{\color{red}
+These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
+}
+%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).
 
-The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
-The first method is the widely-used Clumping + Thresholding (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.
+The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
+{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
 Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
@@
 Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined).
-A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}.
-The second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
+A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}.
+On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}.
 The currently available models are very fast sparse linear and logistic regressions.
@@
 The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}.
-These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr, to remove the need of the choice of the regularization parameter and to fasten the overall algorithms.
+These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of the choice of the regularization parameter.
 
@@
 
-All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data. 
+All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Materials. 
 {\color{red}
@@
 
-We present the results for three different analyses.
+We present the results for four different analyses. 
 First, we illustrate the application of R packages bigstatsr and bigsnpr. 
-Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT).
-Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes {\color{red} (for genotyped SNPs only)}.
+Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}.
+For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
+Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
+Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (for genotyped SNPs only).
 We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
@@
 
-We performed an association study and computed a polygenic risk score for the Celiac cohort. 
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
+The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
 For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
@@
 We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
-As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.
+As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.
 
-\begin{figure}[!tpb]
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=200pt]{celiac-pca}}
@@
 
-\begin{figure}[!tpb]
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
@@
 The GWAS analysis was composed of four main steps.
-First we read from PLINK files in our format "bigSNP" in 1 hour.
+First we read from PLINK files in our format ``bigSNP'' in 1 hour.
 Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h.
@@
 
-\subsection{Method Comparison}
+\subsection{Performance and precision comparisons}
 
-We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
+First, we compared the GWAS computations obtained with bigstatsr and bigsnpr to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4, and also two R packages SNPRelate and GWASTools.
 For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
-Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}).
-However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.
+Overall, performing a GWAS on a binary outcome with our two R packages, bigstatsr and bigsnpr, is as fast as when using EIGENSOFT and PLINK, and much faster than when using R packages SNPRelate and GWASTools.
+For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset as the Celiac dataset.
+Note that the PC scores obtained are more precise (see the last paragraphe of this subsection), which is also the case for the p-values as compared to PLINK (see supplementary notebook ``GWAS-comparison'').
 
-\begin{table}[!tpb]
+\begin{table}[!htpb]
 \begin{center}
@@
 \hline
-Reading PLINK files & n/a & 6 / 20 & 12 /  \\
-Pruning & 4 / 4 & 14 / 52 & 33 / \\  
-Computing 10 PCs & 304 / 315 & 61 / 180 &  /  \\
-GWAS (binary phenotype) & 336 / 293 & 301 / 861 & 15806 /  \\
-GWAS (continuous phenotype) & 1339 /  & 9 /  & 5933 /  \\
+Reading PLINK files         &     n/a     &   6 /  20 &    13 /    33 \\
+Pruning                     &    4 /    4 &  14 /  52 &    33 /    32 \\  
+Computing 10 PCs            &  305 /  314 &  58 / 183 &   323 /   535 \\
+GWAS (binary phenotype)     &  337 /  284 & 291 / 682 & 16220 / 17425 \\
+GWAS (continuous phenotype) & 1348 / 1633 &  10 /  23 &  6115 /  7101 \\
 \hline
-Total & 649 / 612 & 378 / 1113 &  /  \\
+Total (binary)              &  646 /  602 & 369 / 937 & 16589 / 18025 \\
+Total (continuous)          & 1657 / 1951 &  88 / 278 &  6484 /  7701 \\
 \hline
@@
 \end{center}
-\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset.
+\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset (15,155 individuals and 281,122 SNPs).
 The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
@@
 
-\begin{table}[!tpb]
+Secondly, we compared the PRS analysis performed with the R packages to the one using PRSice-2.
+There are 5 main steps in such a analysis (see table~\ref{tab:bench-prs}), including 4 steps handled with functions of package bigstatsr and bigsnpr.
+The remaining step is the reading of summary statistics which is performed with function fread of R package data.table, a widely-used fast and convenient reading function.
+[FINIR TIMING]
+
+\begin{table}[!htpb]
 \begin{center}
@@
 Reading PLINK files   &  & 6 / 20 \\
-Reading summary stats &  & 4 /   \\
-Clumping &  & 9 /  \\
-PRS &  & 2 /  \\
-Compute pvalues &  & 1 /  \\
+Reading summary stats &  & 4 / 6  \\
+Clumping              &  & 9 / 31 \\
+PRS                   &  & 2 / 33? \\
+Compute pvalues       &  & 1 / 1  \\
 \hline
-Total & 22 /  & 22 /  \\
+Total                 & 22 / 29 & 22 /  \\
 \hline
@@
 
-On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr.
+Finally, on our desktop computer, we compared the computation times of FastPCA (fast mode of EIGENSOFT), FlashPCA2 and PLINK 2.0 (approx mode) to the similar function big\_randomSVD implemented in bigstatsr.
 For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs.
@@
 Overall, our function big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
-We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA.
-FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA (Fast mode of EIGENSOFT) whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).
+PLINK 2.0 proved to be faster than bigstatsr with a decrease in time of 20-40\%.
+We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with PLINK 2.0 (exact mode).
+Package bigstatsr and FlashPCA2 (that use the same algorithm) infer all PCs with a squared correlation of more than 0.999 between true PCs and approximated ones (Figure~\ref{fig:prec-pca}).
+Yet, FastPCA (fast mode of EIGENSOFT) and PLINK 2.0 (that use the same algorithm) infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for further PCs (Figure~\ref{fig:prec-pca}).
 
-\begin{figure}[!tpb]
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=200pt]{benchmark-pca.png}}
-\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
+\caption{Benchmark comparisons between randomized partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT), PLINK 2.0 (approx mode) and package bigstatsr.
 It shows the computation time in minutes as a function of the number of samples.
@@
 
-\begin{figure}[!tpb]
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=200pt]{precision-pca.png}}
-\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr.
-It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
+\caption{Precision comparisons between randomized partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT), PLINK 2.0 (approx mode) and package bigstatsr.
+It shows the squared correlation between approximated PCs and ``true'' PCs (produced by the exact mode of PLINK 2.0) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
 \end{figure}
@@
 
-For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets.
+For detecting long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets.
 For the POPRES dataset, the algorithm converged in two iterations.
@@
 We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
-For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
+For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
 
@@
 Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~\ref{fig:loadings}).
-This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
+This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing PCA with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
 
@@
 {\color{red}
-We also note that snp\_fastImpute makes less 0/2 switching errors, i.e.\ imputing with a 0 where it is in fact a 2 or the contrary (Supplementary ..notebook.. ``imputation'').
-We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~\ref{fig:error-impute}), which can be useful for post-processing of the imputation (Figure~\ref{fig:post-imputation}).
+We also note that snp\_fastImpute made less 0/2 switching errors, i.e.\ imputing with a 0 where it is in fact a 2 or the contrary (Supplementary notebook ``imputation'').
+We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~\ref{fig:error-impute}), which can be useful for post-processing the imputation by removing SNPs with too many errors (Figure~\ref{fig:post-imputation}).
 }
-For the Celiac dataset in which there was already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle did not finish imputing chromosome 1 in 48 hours.
+For the Celiac dataset in which there were already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle did not finish imputing chromosome 1 in 48 hours.
 
@@
 
-We have developed two R packages, bigstatsr and bigsnpr, which enable multiple analyses of large-scale genotype datasets in a single comprehensive framework.
-Linkage Disequilibrium pruning, Principal Component Analysis, association tests and computations of polygenic risk scores are made available in this software.
+We have developed two R packages, bigstatsr and bigsnpr, which enable multiple analyses of large-scale genotype datasets in R thanks to memory-mapping.
+Linkage Disequilibrium pruning, Principal Component Analysis, association tests and computation of polygenic risk scores are made available in this software.
 Implemented algorithms are both fast and memory-efficient, allowing the use of laptops or desktop computers to make genome-wide analyses.
@@
 Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+We would like to thank the reviewers of this paper because their comments and suggestions have led to a significant improvement of this paper.
 

### 2018-01-26 cabios-template2/bigsnpr-review.tex
before mike review

@@
 
-\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
+\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
 \author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
@@
 To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods.
-In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs.
-We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
+In particular, the packages implement fast and accurate computations of principal component analysis and association studies, functions to remove SNPs in linkage disequilibrium and algorithms to learn polygenic risk scores on millions of SNPs.
+We illustrate applications of the two R packages by analyzing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
 Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
@@
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
+\textbf{Supplementary information:} Supplementary materials are available at \textit{Bioinformatics}
 online.}
@@
 Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-As a consequence, it has been required to continuously optimize algorithms and this has led to the reimplementation of some software that would have become obsolete otherwise.
-For computing Principal Component Analysis, commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
-PLINK 1.07, which is a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.
+As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise.
+For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
+PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.
 
@@
 This is particularly a burden for commonly used analysis languages such as R.
-For analyzing genotype datasets in R, a range of software are available, including R packages snpStats, GenABEL, SNPRelate and GWASTools.
-Packages snpStats and GenABEL, which have been widely used and cited, are now outdated with current data sizes -- package snpStats still uses in-memory data and package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
-Solving the memory issues for languages such as R would give access to a broad range of tools for data analysis that have already been implemented.
-Hopefully, strategies have been developed to avoid loading large datasets in RAM.
+For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}.
+Package GenABEL, which has been widely used and cited, does not handle standard binary PLINK files and has not been updated since 2013.
+Solving the memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis.
+Fortunately, strategies have been developed to avoid loading large datasets in RAM.
 For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations.
@@
 In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
-For instance, one might like using R because R makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
+For instance, one might like using R because this programming language makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
 We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
@@
 Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
-This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries.
+This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries.
 The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices.
@@
 Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting).
-We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores.
-We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.
+We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS).
+We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice.
+We report execution times along with the code to perform major computational tasks.
 
@@
 With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}.
-The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
+The major advantage of manipulating genotype data within R, almost as if it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}.
 For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.
@@
 When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
-All this is managed by the Operating System so that it is seamless and efficient. 
+All this is managed by the operating system so that it is seamless and efficient. 
 It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
@@
 We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
-Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls.
+Indeed, each element is stored in one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls.
 With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).
@@
 
-For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).
+For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other one based on randomized projections, respectively named big\_SVD and big\_randomSVD (Figure~\ref{fig:svd}).
 While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}.
@@
 
-\begin{figure}[!htpb]
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=235pt]{svd.pdf}}
@@
 Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome.
-For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the MAF as the statistic of importance. 
-Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest minor allele frequency (MAF) is kept.
-Yet, in some worst-case scenario, this pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
+For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
+Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest MAF is kept.
+Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
 So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
@@
 
-As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
-Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in long-range LD regions begore computing PCs.
+As mentioned above, the third approach that is generally combined with pruning and consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
+Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in these regions before computing PCs.
 However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them.
@@
 \end{multline}
-and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender).
+and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender).
 Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
 \begin{multline}
-  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
+  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
 \end{multline}
@@
 {\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
-Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value.
+Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part).
 The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other.
@@
 The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}.
-These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of the choice of the regularization parameter.
+These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of choosing the regularization parameter.
 
@@
 The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset.
+For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset.
 To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
@@
 
-All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Materials. 
+All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the supplementary materials. 
 {\color{red}
@@
 Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
-Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (for genotyped SNPs only).
+Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only).
 We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
@@
 We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}).
-We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
+We then performed a Genome-Wide Association Study (GWAS) investigating how SNPs are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
 As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.
 
-\begin{figure}[!htpb]
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=200pt]{celiac-pca}}
@@
 
-\begin{figure}[!htpb]
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
@@
 The GWAS analysis was composed of four main steps.
-First we read from PLINK files in our format ``bigSNP'' in 1 hour.
+First we convert binary PLINK files in our format ``bigSNP'' in 1 hour.
 Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h.
@@
 Overall, performing a GWAS on a binary outcome with our two R packages, bigstatsr and bigsnpr, is as fast as when using EIGENSOFT and PLINK, and much faster than when using R packages SNPRelate and GWASTools.
-For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset as the Celiac dataset.
-Note that the PC scores obtained are more precise (see the last paragraphe of this subsection), which is also the case for the p-values as compared to PLINK (see supplementary notebook ``GWAS-comparison'').
+For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset such as the Celiac dataset (Table~\ref{tab:bench-gwas}).
+Note that the PC scores obtained are more precise as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').
 
-\begin{table}[!htpb]
+\begin{table}[!tpb]
 \begin{center}
@@
 \hline
-Reading PLINK files         &     n/a     &   6 /  20 &    13 /    33 \\
+Converting PLINK files         &     n/a     &   6 /  20 &    13 /    33 \\
 Pruning                     &    4 /    4 &  14 /  52 &    33 /    32 \\  
@@
 Secondly, we compared the PRS analysis performed with the R packages to the one using PRSice-2.
-There are 5 main steps in such a analysis (see table~\ref{tab:bench-prs}), including 4 steps handled with functions of package bigstatsr and bigsnpr.
-The remaining step is the reading of summary statistics which is performed with function fread of R package data.table, a widely-used fast and convenient reading function.
-[FINIR TIMING]
+There are 5 main steps in such an analysis (see table~\ref{tab:bench-prs}), including 4 steps handled with functions of packages bigstatsr and bigsnpr.
+The remaining step is the reading of summary statistics which can be performed with the widely-used function fread of R package data.table.
+Using bigstatsr and bigsnpr results in an analysis as fast as with PRSice-2 when using our desktop computer, and three times slower when using our laptop (Table~\ref{tab:bench-prs}).
 
-\begin{table}[!htpb]
+\begin{table}[!tpb]
 \begin{center}
@@
 \hline
-Reading PLINK files   &  & 6 / 20 \\
+Converting PLINK files   &  & 6 / 20 \\
 Reading summary stats &  & 4 / 6  \\
 Clumping              &  & 9 / 31 \\
-PRS                   &  & 2 / 33? \\
+PRS                   &  & 2 / 33 \\
 Compute pvalues       &  & 1 / 1  \\
 \hline
-Total                 & 22 / 29 & 22 /  \\
+Total                 & 22 / 29 & 22 / 91 \\
 \hline
@@
 For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs.
-We used the datasets of growing size simulated from the Celiac dataset.
-Overall, our function big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism (an option not currently possible with either FastPCA or FlashPCA2) with 6 cores (Figure~\ref{fig:bench-pca}).
+We used the datasets of growing size simulated from the Celiac dataset (from 15,155 to 151,550 individuals).
+Overall, our function  big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism with 6 cores, an option not currently possible with either FastPCA or FlashPCA2 (Figure~\ref{fig:bench-pca}).
 PLINK 2.0 proved to be faster than bigstatsr with a decrease in time of 20-40\%.
-We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with PLINK 2.0 (exact mode).
+We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact eigen decomposition obtained with PLINK 2.0 (exact mode).
 Package bigstatsr and FlashPCA2 (that use the same algorithm) infer all PCs with a squared correlation of more than 0.999 between true PCs and approximated ones (Figure~\ref{fig:prec-pca}).
@@
 
-\begin{figure}[!htpb]
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=200pt]{benchmark-pca.png}}
@@
 
-\begin{figure}[!htpb]
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=200pt]{precision-pca.png}}
@@
 For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset (with no missing value). 
-The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf} and there is no missing value. 
+The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf}. 
 We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}).
@@
 We have developed two R packages, bigstatsr and bigsnpr, which enable multiple analyses of large-scale genotype datasets in R thanks to memory-mapping.
-Linkage Disequilibrium pruning, Principal Component Analysis, association tests and computation of polygenic risk scores are made available in this software.
+Linkage disequilibrium pruning, principal component analysis, association tests and computation of polygenic risk scores are made available in these software.
 Implemented algorithms are both fast and memory-efficient, allowing the use of laptops or desktop computers to make genome-wide analyses.
@@
 We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data.
-For example, we wish to use these packages to train much more accurate predictive models than the standard C+T model currently in use when computing Polygenic Risk Scores.
+For example, we wish to use these packages to train much more accurate predictive models than the standard C+T model currently in use for computing polygenic risk scores.
 As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS \cite[]{Palmer2016}, and it could be implemented with the data format ``FBM.code256'' without having to write multiple files.

### 2018-01-29 cabios-template2/bigsnpr-review.tex
changes with Mike

@@
 As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise.
-For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
-PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.
+For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to the software EIGENSOFT, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
+PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9 to speed-up computations, and there is also an alpha version of PLINK 2.0 that will handle more data types \cite[]{chang2015second,Purcell2007}.
 
@@
 For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}.
-Package GenABEL, which has been widely used and cited, does not handle standard binary PLINK files and has not been updated since 2013.
-Solving the memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis.
+The popular package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
+Solving memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis.
 Fortunately, strategies have been developed to avoid loading large datasets in RAM.
@@
 In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
-For instance, one might like using R because this programming language makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
+R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
 We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
@@
 
-For preprocessing steps, we think the user should use PLINK, which is an invaluable tool for this.
+For preprocessing steps, PLINK is a widely-used software.
 For the sake of reproducibility, one could use PLINK directly from R via systems calls.
@@
 We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
-Finally, we provide two methods for reading-in dosages as proof of concept and will provide a more final solution when we get access to the UK Biobank dataset ([SUPPMAT]).
+Finally, we provide two methods for reading dosage data ([SUPPMAT]).
 
@@
 The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}).
-Beagle uses phasing before imputing, which could take several weeks even when there are only a tiny amount of missing values to be imputed.
+Beagle uses phasing before imputing, which could require several weeks of computation even when there are only a tiny amount of missing values to be imputed.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
@@
 For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
-Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest MAF is kept.
+This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
 Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
 So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
-In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
+In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (result not shown).
 }
 
-As mentioned above, the third approach that is generally combined with pruning and consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
+The third approach, which is generally combined with pruning, consists of removing SNPs in long-range LD regions \cite[]{Price2008}.
 Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in these regions before computing PCs.
@@
 Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries.
-Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}.
+Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes in bigsnpr \cite[]{Zheng2012}.
 Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
@@
 For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset.
-To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
+To assess scalibility of the packages for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
 \label{sec:rep}
@@
 
-We present the results for four different analyses. 
+We present the results of four different analyses. 
 First, we illustrate the application of R packages bigstatsr and bigsnpr. 
 Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}.
-For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
+PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
 Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
 Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only).
-We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
 We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores).
@@
 The 0.04\% missing genotype values were imputed with the XGBoost method.
-If we used a standard R matrix to store the genotypes, this data would require 32GB of memory.
+If we would have used a standard R matrix to store the genotypes, this data would have required 32GB of memory.
 On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.
@@
 The GWAS analysis was composed of four main steps.
-First we convert binary PLINK files in our format ``bigSNP'' in 1 hour.
+First we convert binary PLINK files in the format ``bigSNP'' in 1 hour.
 Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h.
@@
 For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}).
-Overall, performing a GWAS on a binary outcome with our two R packages, bigstatsr and bigsnpr, is as fast as when using EIGENSOFT and PLINK, and much faster than when using R packages SNPRelate and GWASTools.
-For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset such as the Celiac dataset (Table~\ref{tab:bench-gwas}).
+Overall, performing a GWAS on a binary outcome with bigstatsr and bigsnpr is as fast as when using EIGENSOFT and PLINK, and 19-45 times faster than when using R packages SNPRelate and GWASTools. 
+For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset such as the Celiac dataset.
+This analysis was 7-19 times faster as compared to PLINK 1.9 and 28-74 times faster as compared to SNPRelate and GWASTools (Table~\ref{tab:bench-gwas}).
 Note that the PC scores obtained are more precise as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').
@@
 We used the datasets of growing size simulated from the Celiac dataset (from 15,155 to 151,550 individuals).
-Overall, our function  big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism with 6 cores, an option not currently possible with either FastPCA or FlashPCA2 (Figure~\ref{fig:bench-pca}).
-PLINK 2.0 proved to be faster than bigstatsr with a decrease in time of 20-40\%.
+Overall, function  big\_randomSVD is almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast as when using parallelism with 6 cores, an option not currently available in either FastPCA or FlashPCA2 (Figure~\ref{fig:bench-pca}).
+PLINK 2.0 is faster than bigstatsr with a decrease in time of 20-40\%.
 We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact eigen decomposition obtained with PLINK 2.0 (exact mode).
@@
 We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
-For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
+For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
 
@@
 Overall, in 3 runs, snp\_fastImpute made only 4.7\% of imputation errors and Beagle made only 3.1\% of errors.
-Yet, it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less).
+Yet, it took Beagle 14.6 hours to complete while snp\_fastImpute only took 42 minutes (20 times less).
 {\color{red}

### 2018-01-29 cabios-template2/bigsnpr-review.tex
need changes from H

@@
 We report execution times along with the code to perform major computational tasks.
+{\color{red}
+For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".
+}
 
@@
 We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
-Finally, we provide two methods for reading dosage data ([SUPPMAT]).
+Finally, we provide two methods for reading dosage data (Supplementary notebook ``dosage'').
 
 Most modern SNP chips provide genotype data with large call-rates.
-For example, the Celiac data we use in this paper presents only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control.
+For example, the Celiac data we use in this paper presents only a proportion of 0.04\% of missing values for the genotyped SNPs after quality control.
 Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
@@
 The training set is used to build the XGBoost model for predicting missing data.
-The prediction model is then evaluated on the validation set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
+The prediction model is then evaluated on the validation set for which we know the true genotype values, providing an estimator of the number of genotypes that have been wrongly imputed for that particular SNP.
 The prediction model is also projected on the test set (missing values) in order to impute them.
@@
 This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
-Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
+Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all (Supplementary notebook ``pruning-vs-clumping''). 
 So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
-In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (result not shown).
+In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).
 }
@@
 
-Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
 We would like to thank the reviewers of this paper because their comments and suggestions have led to a significant improvement of this paper.

### 2018-01-29 public-data.Rmd
need changes from H

@@
 
-```{r, eval=FALSE}
+```{r}
 bed <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104774/cornell_canine.bed"
@@
 ```{r}
+library(bigsnpr)
 plink <- download_plink()

### 2018-01-29 cabios-template2/bigsnpr-review-supp.tex
make final changes

@@
 
-\section{Supplementary Data}
+\section{Supplementary Materials}
 
@@
 
-\begin{figure}[htpb]
-\centerline{\includegraphics[width=160pt]{conversion-QC.pdf}}
-\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\linewidth]{conversion-QC.pdf}}
+\caption{Conversion and quality control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
 \end{figure}
 
-\begin{figure}[htpb]
-\centerline{\includegraphics[width=235pt]{imputation.pdf}}
+\newpage
+
+\begin{figure}[!h]
+\centerline{\includegraphics[width=\linewidth]{imputation.pdf}}
 \caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
@@
 
+\newpage
+
 \subsection{Long-range LD regions}
 
-\begin{table}[htpb]
+\begin{table}[!h]
 \centering
@@
 \end{tabular}
-\caption{Regions found by snp\_autoSVD for the POPRES dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
-\label{tab:lrldr-popres}
+\caption{\label{tab:lrldr-popres}Regions found by snp\_autoSVD for the POPRES dataset. Numbers in parentheses correspond to regions referenced in Price et al., 2008.}
 \end{table}
+
+\vspace{5em}
 
@@
 % Tue Jun 13 11:29:13 2017
-\begin{table}[htpb]
+\begin{table}[!h]
 \centering
@@
 
+\newpage
+
 % latex table generated in R 3.3.3 by xtable 1.8-2 package
 % Tue Jul 18 15:59:51 2017
-\begin{table}[htpb]
+\begin{table}[!h]
 \centering
@@
 \end{tabular}
-\caption{Regions found by snp\_autoSVD for the celiac dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
+\caption{Regions found by snp\_autoSVD for the celiac dataset. Numbers in parentheses correspond to regions referenced in Price et al., 2008.}
 \label{tab:lrldr-celiac}
@@
 
+\vspace{5em}
+
 % latex table generated in R 3.3.3 by xtable 1.8-2 package
 % Tue Jun 13 11:25:56 2017
-\begin{table}[htpb]
+\begin{table}[!h]
 \centering
@@
 
-\begin{figure}[htpb]
+\newpage
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{scores}}
 \caption{PC4 and PC5 of the celiac disease dataset.
-Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$).
+Left panel, PC scores obtained without removing any long range LD region (only clumping at $r^2 > 0.2$).
 Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4.
@@
 
-\begin{figure}[htpb]
+\vspace{3em}
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{loadings}}
 \caption{Loadings of first 6 PCs of the celiac disease dataset plotted as hexbins (2-D histogram with hexagonal cells).
-On the left, without removing any long range LD region (only clumping at $R^2 > 0.2$).
+On the left, without removing any long range LD region (only clumping at $r^2 > 0.2$).
 On the right, with the automatic detection and removal of long-range LD regions.}\label{fig:loadings}
@@
 
-\begin{figure}[htpb]
+\vspace{3em}
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{gini}}
-\caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $R^2 > 0.2$) and with the automatic detection and removal of long-range LD regions.
+\caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $r^2 > 0.2$) and with the automatic detection and removal of long-range LD regions.
 The dashed line corresponds to the theoretical value for gaussian loadings.}\label{fig:gini}
@@
 \newpage
+
 \subsection{Imputation}
 
-\begin{figure}[htpb]
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{hist-maf}}
@@
 
-\begin{figure}[htpb]
+\vspace{3em}
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{hist-NA}}
@@
 
-\begin{figure}[htpb]
+\vspace{3em}
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{error-impute}}
@@
 For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect.
-The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
+The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute.}\label{fig:error-impute}
 \end{figure}
 
-\begin{figure}[htpb]
+\vspace{3em}
+
+\begin{figure}[!h]
 \centerline{\includegraphics[width=\textwidth]{post-imputation}}
@@
 
-\vspace*{-12pt}
-
-\bibliographystyle{natbib}
-\bibliography{document}
-
 \end{document}

### 2018-01-29 cabios-template2/bigsnpr-review.tex
make final changes

@@
 \address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France and \\
+$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
 $^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}
@@
 Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise.
+As a consequence, most existing software and algorithms have to be continuously optimized in order to avoid obsolescence.
 For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to the software EIGENSOFT, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
@@
 This is particularly a burden for commonly used analysis languages such as R.
-For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}.
-The popular package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
+For analyzing genotype datasets in R, a range of software are available, including for example the popular R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}.
 Solving memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis.
@@
 Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting).
+{\color{red}
 We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS).
-We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice.
+We compare results from bigstatsr and bigsnpr with those obtained by using command-line sofwtare PLINK, EIGENSOFT and PRSice, and R packages SNPRelate and GWASTools.
 We report execution times along with the code to perform major computational tasks.
-{\color{red}
 For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".
@@
 For the sake of reproducibility, one could use PLINK directly from R via systems calls.
-For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
+We therefore provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
 Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs.
 We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.
-Finally, we provide two methods for reading dosage data (Supplementary notebook ``dosage'').
+Finally, we provide two methods for converting dosage data to the format ``bigSNP'' (Supplementary notebook ``dosage'').
 
 Most modern SNP chips provide genotype data with large call-rates.
-For example, the Celiac data we use in this paper presents only a proportion of 0.04\% of missing values for the genotyped SNPs after quality control.
+For example, the Celiac data we use in this paper presents only 0.04\% of missing values after quality control.
 Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
@@
 The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}).
-Beagle uses phasing before imputing, which could require several weeks of computation even when there are only a tiny amount of missing values to be imputed.
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
-We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
+This function also provides an estimator of the imputation error rate by SNP for post quality control.
 This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time.
-It only relies on some local XGBoost models.
-XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
-XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
+It only relies on some local XGBoost models \cite[]{Chen2016}.
+XGBoost, which is available in R, builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
 Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP.
@@
 Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all (Supplementary notebook ``pruning-vs-clumping''). 
-So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
+So, we suggest to use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr.
 In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).
@@
 {\color{red}
-These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
+These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of optimized implementations, partly based on previous work by \cite{sikorska2013gwas}. 
 }
@@
 This analysis was 7-19 times faster as compared to PLINK 1.9 and 28-74 times faster as compared to SNPRelate and GWASTools (Table~\ref{tab:bench-gwas}).
-Note that the PC scores obtained are more precise as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').
+Note that the PC scores obtained are more accurate as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').
 

### 2018-01-30 cabios-template2/bigsnpr-review-supp.tex
add andrey

@@
 \graphicspath{{../figures/}}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{adjustbox}
+\usepackage{multirow}
 
@@
 
-\newpage
+\clearpage
 
@@
 
+\clearpage
+\subsection{Overview}
+
+\begin{table}[!h]
+\caption{Comparison of tools in low-level data management (first three rows) and high-level genetic analyses (the rest of rows).
+The bigstatsr and bigsnpr packages provide the most comprehensive and versatile infrastructure for data storage, processing and extensions; implement the main genetic analysis pipelines; and intrincically exploit memory-mapped files and parallel data processing.
+Notes:
+1: binary data format;
+2: multi-core and cluster parallel computation;
+3: support of imputed genotype data.
+When comparing tools in terms of parallel computation, we do not consider the case of jobs manually distributed, e.g.\ splitting genotypes into batches.
+We are aware of the bitwise parallelism in PLINK 1.9, but this feature is used internally for specific computations; instead, we follow the official documentaion (https://www.cog-genomics.org/PLINK/1.9/parallel).
+The official documentation for the GenABEL project is at http://www.genabel.org/packages/ParallABEL.
+GRM: Genetic Relationship Matrix; PCA: Principal Component Analysis; GWAS: Genome-Wide Association Study; PRS: Polygenic Risk Score.}
+\begin{center}
+\begin{adjustbox}{max width=\textwidth}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+\multirow{2}{*}{Toolkit/Feature} & bigstatsr~/ & \multirow{2}{*}{PLINK 1.9} & \multirow{2}{*}{GCTA} & SNPRelate~/ & \multirow{2}{*}{GenABEL} \\
+ & bigsnpr & & & GWASTools &  \\
+\hline
+\hline
+Proper genotype data format & FBM$^{2,3}$ & ped/bed$^2$ & - & GDS$^{2,3}$ & databel$^{2,3}$ \\ \hline
+Data format extension, e.g. -omics & $\checkmark$ & - & - & - & $\checkmark$ \\ \hline
+Vector/Matrix operations & $\checkmark^1$ & - & - & - & $\checkmark^1$ \\ \hline
+Quality Control & $\checkmark$ & $\checkmark$ & - & $\checkmark$ & $\checkmark$ \\ \hline
+GRM & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ \\ \hline
+PCA & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ & $\checkmark^1$ & - \\ \hline
+GWAS & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ \\ \hline
+PRS & $\checkmark$ & $\checkmark$ & - & - & - \\ \hline
+\end{tabular}
+\end{adjustbox}
+\end{center}
+\label{tab:andrey}
+\end{table}
+
 \end{document}

### 2018-01-30 cabios-template2/bigsnpr-review.tex
add andrey

@@
 \title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
+\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
 \address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
@@
 For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
+We present a table summarizing the features of different software in table \ref{tab:andrey}.
 

### 2018-01-31 cabios-template2/bigsnpr-review.tex
prepare resubmission

@@
 \author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
-\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\address{$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
 $^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
@@
 In particular, the packages implement fast and accurate computations of principal component analysis and association studies, functions to remove SNPs in linkage disequilibrium and algorithms to learn polygenic risk scores on millions of SNPs.
-We illustrate applications of the two R packages by analyzing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores.
+We illustrate applications of the two R packages by analyzing a case-control genomic dataset for celiac disease, performing an association study and computing Polygenic Risk Scores.
 Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer.
@@
 For the sake of reproducibility, one could use PLINK directly from R via systems calls.
-We therefore provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
+We therefore provide wrappers as R functions that use system calls to PLINK for conversion and quality control and a variety of formats can be used as input (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
 Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs.
@@
 Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome.
-For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
+For PCA, the thinning procedure should remain unsupervised (no phenotype must be used) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
 This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
@@
 Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes in bigsnpr \cite[]{Zheng2012}.
-Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
+Package bigstatsr implements statistical tests based on linear and logistic regressions. For linear regression, a t-test is performed for each SNP $j$ on $\beta^{(j)}$ where
 \begin{multline}
@@
 and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender).
-Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
+Similarly, for logistic regression, a Z-test is performed for each SNP $j$ on $\beta^{(j)}$ where
 \begin{multline}
@@
 
-In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}.
+In this paper, two datasets are analyzed: the celiac disease cohort and POPRES \cite[]{Dubois2010,Nelson2008}.
 The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs.
@@
 PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
-Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
+Thirdly, by performing a PRS analysis with summary statistics, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
 Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only).
@@
 We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}).
-We then performed a Genome-Wide Association Study (GWAS) investigating how SNPs are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
+We then performed a Genome-Wide Association Study (GWAS) investigating how SNPs are associated with celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}).
 As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.
@@
 {\color{red}
-We also note that snp\_fastImpute made less 0/2 switching errors, i.e.\ imputing with a 0 where it is in fact a 2 or the contrary (Supplementary notebook ``imputation'').
+We also note that snp\_fastImpute made less 0/2 switching errors, i.e.\ imputing with a homozygous referent where the true genotype is a homozygous variant, or the contrary (Supplementary notebook ``imputation'').
 We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~\ref{fig:error-impute}), which can be useful for post-processing the imputation by removing SNPs with too many errors (Figure~\ref{fig:post-imputation}).

### 2018-02-02 cabios-template2/bigsnpr-review.tex
replace suppmat refs

@@
 \usepackage{graphicx}
-\graphicspath{{../figures/}}
-\usepackage{xr}
-\externaldocument{bigsnpr-review-supp}
+%\graphicspath{{../figures/}}
+%\usepackage{xr}
+%\externaldocument{bigsnpr-review-supp}
 
@@
 \title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
-\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
+\author[Priv\'e \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
 \address{$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
@@
 \abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
-{\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
+This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.
+\\
 \textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R.
@@
 
-{\color{red}
 Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
@@
 Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
-}
 
@@
 
-{\color{red}
 In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
@@
 This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
-}
 Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double).
@@
 Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting).
-{\color{red}
 We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS).
@@
 For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".
-}
 
@@
 
-{\color{red}
 We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage.
@@
 For the sake of reproducibility, one could use PLINK directly from R via systems calls.
-We therefore provide wrappers as R functions that use system calls to PLINK for conversion and quality control and a variety of formats can be used as input (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
+We therefore provide wrappers as R functions that use system calls to PLINK for conversion and quality control and a variety of formats can be used as input (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~S1).
 Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs.
@@
 Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}.
-The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}).
+The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~S2).
 The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy.
@@
 The prediction model is also projected on the test set (missing values) in order to impute them.
-}
 
@@
 First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
-{\color{red}
 Clumping is useful if a statistic is available for sorting the SNPs by importance.
@@
 In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).
-}
 
@@
 and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
-{\color{red}
 These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of optimized implementations, partly based on previous work by \cite{sikorska2013gwas}.
-}
-%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).
 
 The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
-{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
+The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.
 Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part).
@@
 The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
-For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset.
+For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same eigen decomposition (up to a constant) and pairwise SNP correlations as the original dataset.
 To assess scalibility of the packages for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset.
@@
 All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the supplementary materials. 
-{\color{red}
 In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
-}
 
@@
 For the functions that enable parallelism, we use half of the cores available on the corresponding computer.
-We present a table summarizing the features of different software in table \ref{tab:andrey}.
+We present a table summarizing the features of different software in table S9.
 
@@
 
-The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
+The data was preprocessed following steps from figure~S1, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}.
 For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
@@
 For the POPRES dataset, the algorithm converged in two iterations.
-The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}).
-We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}).
-For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
+The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~S1).
+We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~S2).
+For the Celiac dataset, we found 5 long-range LD regions (Table~S3) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~S4).
 
 For the Celiac dataset, we further compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). 
-When not removing any long range LD region, we show that PC4 and PC5 do not capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}).
-When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~\ref{fig:scores}).
-Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~\ref{fig:loadings}).
-This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing PCA with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).
+When not removing any long range LD region, we show that PC4 and PC5 do not capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~S3 and~S4). 
+When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~S3).
+Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~S4).
+This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing PCA with snp\_autoSVD than when no long-range LD region is removed (Figure~S5).
 
@@
 For the imputation method based on XGBoost, we compared  the imputation accuracy and computation times with Beagle on the POPRES dataset (with no missing value). 
-The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~\ref{fig:maf}. 
-We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~\ref{fig:NA}).
-Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017) 
-{\color{red}
-by counting the percentage of imputation errors (when the imputed genotype is different from the true genotype)
-}.
+The histogram of the minor allele frequencies (MAFs) of this dataset is provided in figure~S6. 
+We used a Beta-binomial distribution to simulate the number of missing values by SNP and then randomly introduced missing values according to these numbers, resulting in approximately 3\% of missing values overall (Figure~S7).
+Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017) by counting the percentage of imputation errors (when the imputed genotype is different from the true genotype).
 Overall, in 3 runs, snp\_fastImpute made only 4.7\% of imputation errors and Beagle made only 3.1\% of errors.
 Yet, it took Beagle 14.6 hours to complete while snp\_fastImpute only took 42 minutes (20 times less).
-{\color{red}
 We also note that snp\_fastImpute made less 0/2 switching errors, i.e.\ imputing with a homozygous referent where the true genotype is a homozygous variant, or the contrary (Supplementary notebook ``imputation'').
-We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~\ref{fig:error-impute}), which can be useful for post-processing the imputation by removing SNPs with too many errors (Figure~\ref{fig:post-imputation}).
-}
+We also show that the estimation of the number of imputation errors provided by function snp\_fastImpute is accurate (Figure~S8), which can be useful for post-processing the imputation by removing SNPs with too many errors (Figure~S9).
 For the Celiac dataset in which there were already missing values, in order to further compare computation times, we report that snp\_fastImpute took less than 10 hours to complete for the whole genome whereas Beagle did not finish imputing chromosome 1 in 48 hours.
